---
title: "The complete workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The complete workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(DesiredGainR)
data(dgr_traits, package = "DesiredGainR")
data(dgr_G, package = "DesiredGainR")
data(dgr_P, package = "DesiredGainR")
data(dgr_candidates, package = "DesiredGainR")
data(dgr_gebv, package = "DesiredGainR")
data(dgr_history, package = "DesiredGainR")
traits <- dgr_traits$trait
lower_is_better <- dgr_traits$trait[dgr_traits$direction == "decrease"]
```

## 1. Purpose

Every other vignette is a deep dive into one layer. This one runs the whole
package, start to finish, on the shipped example programme, so that the
relationships between the stages are visible in one place rather than left to
be inferred across eight documents.

The order below is the order in which the decisions actually arise. It is not
the order of the function reference, and it deliberately places the objective
before the index, because an index built on an objective that was never tested
is a well-computed answer to an unexamined question.

Each stage gives the call and the decision it feeds. The vignette named at the
end of each stage carries the full argument reference and the statistical
detail.

| Phase | Stages | Question |
|---|---|---|
| Evidence | 1-2 | What do we have, and can it be trusted? |
| Objective | 3-5 | What should we select for, and is it possible? |
| Index | 6-8 | Which index, and what does it actually do? |
| Confidence | 9-10 | Would a different objective change the decision? |

---

## 2. Stage 1 — Inspect the covariance matrices

Index coefficients are obtained by inverting `P` and `G`. An ill-conditioned
matrix yields coefficients that are numerically meaningless while remaining
superficially plausible, so conditioning is checked before anything is built on
top of it.

```{r conditioning}
matrix_diagnostics(dgr_G, "G")$condition_number
matrix_diagnostics(dgr_P, "P")$condition_number
```

Those numbers look alarming until they are decomposed. Almost all of the
ill-conditioning comes from the trait scales rather than from the correlations:

```{r conditioning-decomposed}
matrix_diagnostics(stats::cov2cor(dgr_G))$condition_number
```

Standardising removes three orders of magnitude. This is the practical form of
the observation by Crosbie et al. (1980) that equal weights on unstandardised
traits concentrate selection on whichever trait carries the largest variance,
and it is why the package standardises by default.

See [Obtaining G and P](DesiredGainR-covariance.html).

---

## 3. Stage 2 — Confirm what the matrices represent

`G` should be a genetic covariance matrix from a fitted multi-trait model.
Where none exists, `estimate_genetic_covariance()` provides labelled
approximations, including the covariance of across-environment adjusted means
that the CGIAR guideline recommends for operational use. Each method records
its estimand rather than relabelling a covariance of predictions as a genetic
covariance.

```{r estimand, eval=FALSE}
G_working <- estimate_genetic_covariance(
  values = dgr_candidates, trait_cols = traits,
  method = "adjusted_means_surrogate"
)
G_working$estimand
#> [1] "working surrogate for genetic covariance. Covariance of
#>      across-environment adjusted means"
```

The example data ship a known `G`, so the rest of this vignette uses it
directly.

---

## 4. Stage 3 — State the objective

Desired gains are stated as improvements, and `lower_is_better` declares which
traits improve by falling. The covariance matrices are oriented internally, so
nothing needs signing by hand.

```{r objective}
desired_gains <- c(
  GY = 1.0, PHT = 0.4, AD = 0.6, ASI = 0.5,
  EPP = 0.4, GLS = 0.6
)

implied <- implied_economic_weights(
  desired_gains, dgr_G, dgr_P,
  lower_is_better = lower_is_better, gain_units = "genetic_sd"
)
round(implied, 2)
```

Two of those weights are negative for traits the objective asks to improve,
which is correct rather than an error. In addition, the target values are
already genetic-standard-deviation gains, whereas the returned weights are per
original trait unit because `dgr_G` and `dgr_P` are in original units. The
following multiplication expresses each weight per genetic standard deviation.
it does not standardise the target again:

```{r rescale}
genetic_sd <- stats::setNames(dgr_traits$genetic_sd, dgr_traits$trait)
round(sort(implied * genetic_sd[names(implied)], decreasing = TRUE), 2)
```

If the covariance matrices had already been transformed to genetic-standard-
deviation units, this multiplication would be incorrect because the weights
would already be on that scale.

See [Defining a breeding objective](DesiredGainR-objective.html).

---

## 5. Stage 4 — Test whether the objective is attainable

This stage exists because an optimisation can return a numerical answer even
when the stated one-cycle target is outside the achievable-response ellipsoid.
`gain_feasibility()` tests the exact desired-gain ray before any shortfall is
interpreted.

```{r feasibility}
feasibility <- gain_feasibility(
  desired_gains, dgr_G, dgr_P,
  n_candidates = nrow(dgr_candidates), n_select = 20L,
  lower_is_better = lower_is_better, gain_units = "genetic_sd"
)
feasibility
```

The required intensity is 3.2082, equivalent under normal truncation to the
best 0.1773%. Among 200 candidates, that proportion represents 0.355 candidate.
Rounding to one changes the selection proportion to 0.5% and lowers the
intensity. Therefore, the exact vector lies beyond this finite
candidate set.

The planned selection of 20 candidates has intensity 1.7550. Its ratio to the
required intensity is 0.547, so 54.7% of every target component is available
while the requested trait proportions are preserved.

```{r attainable}
round(feasibility$attainable_response_input_units, 3)
```

These values are in the same genetic-standard-deviation units as the input.
They describe the point on the requested ray, not independent upper limits for
the six traits. If the entries are minimum floors rather than a fixed ratio,
use population-driven or interval optimisation and inspect worst-trait and
joint attainment. The conclusion remains conditional on the linear model,
normal truncation selection and the estimated covariance matrices.

Carry this number forward to Stage 8, where the optimiser will be seen to
distribute the shortfall unevenly rather than uniformly.

---

## 6. Stage 5 — Recover what the programme already selects for

Where an index has never been used, the differentials a programme has achieved
already encode its objective.

```{r retrospective}
recovered <- retrospective_weights(
  selected_values = dgr_candidates[dgr_history$selected, traits],
  population_values = dgr_candidates[, traits],
  trait_cols = traits
)
round(recovered$selection_differential_sd, 2)
```

The recovered rule reproduces past decisions, including any bias in them, so it
is a starting point rather than a forward-looking objective. The objective
vignette verifies the recovery against the weights that generated the
historical decision.

---

## 7. Stage 6 — Build and compare index families

```{r indices}
economic_weights <- c(
  GY = 1.0, PHT = 0.2, AD = 0.5,
  ASI = 0.4, EPP = 0.3, GLS = 0.5
)

smith_hazel <- selection_index(
  dgr_candidates, traits,
  method = "smith_hazel",
  G = dgr_G, P = dgr_P, economic_weights = economic_weights,
  lower_is_better = lower_is_better, n_select = 20L, main_trait = "GY"
)
rank_sum <- selection_index(
  dgr_candidates, traits,
  method = "mulamba_mock",
  lower_is_better = lower_is_better, n_select = 20L
)
base_index <- selection_index(
  dgr_candidates, traits,
  method = "base",
  G = dgr_G, P = dgr_P, economic_weights = economic_weights,
  lower_is_better = lower_is_better, n_select = 20L, main_trait = "GY"
)
smith_hazel
```

Relative efficiency below one is the intended trade of response in the main
trait for response elsewhere in the objective, not a defect. Every value
reported by Rahimi and Debnath (2023) was below one.

The Mulamba-Mock rank-sum index needs neither weights nor covariance matrices.
This explains its value as a low-assumption comparator. The base index shows
how much the covariance adjustment contributes beyond direct weighting.

Use `comparison_objective()` to fix one target, utility and genetic covariance.
Then use `compare_selection_methods()` across every supported fitted family.
Read biological responses, response basis, target attainment, rank agreement,
selected-set agreement and common validation response. See [Multiple-trait
selection](DesiredGainR-index-families.html) for the complete workflow.

```{r compare}
comparison <- compare_selection_methods(
  list(
    Smith_Hazel = smith_hazel,
    Base = base_index,
    Rank_sum = rank_sum
  )
)
comparison$fairness
comparison$summary
round(comparison$rank_correlation, 2)
comparison$selected_overlap
```

Rank correlation uses all candidates. Selected-set overlap concerns the final
decision. Report both because close overall rankings can still diverge near the
selection boundary.

See [Multiple-trait selection](DesiredGainR-index-families.html).

---

## 8. Stage 7 — Check which traits the index acts on

```{r effective}
smith_hazel$effective_weights[, c("Trait", "Genetic_share")]
```

---

## 9. Stage 8 — Optimise the desired-gain vector

`run_dgsi()` applies the iterative optimisation procedure of Joukhadar et al.
(2024) to the established desired-gain index. It searches for an input
desired-gain vector whose realised response in the selected set approaches the
stated target.

```{r dgsi}
candidate_sd <- vapply(
  dgr_candidates[, traits], stats::sd, numeric(1L)
)
target_original_units <- desired_gains * genetic_sd
dg_target <- target_original_units / candidate_sd

dgsi <- run_dgsi(
  init_data = dgr_candidates[, c("GenoID", "Family")],
  cand_data = dgr_candidates,
  trait_cols = traits,
  dg = dg_target,
  G = dgr_G, P = dgr_P,
  lower_is_better = lower_is_better,
  scale_traits = TRUE,
  n_select = 20L, n_iter = 200L, n_rep = 5L, seed = 42L
)
round(dgsi$realised_response - dg_target, 3)
```

`realised_response` is a selected-candidate differential in candidate standard
deviations. Read the model-based transmitted response in the original
genetic-standard-deviation units before judging attainment.

```{r dgsi-transmitted}
transmitted_genetic_sd <-
  dgsi$theoretical_response$original_units / genetic_sd
round(transmitted_genetic_sd, 3)
round(transmitted_genetic_sd / desired_gains, 3)
```

Feasibility constrains the requested response ray. The iterative search can
leave that ray. Therefore, inspect every trait-specific attainment value.

The classical solution without iteration is returned as a comparator:

```{r non-iterated}
round(dgsi$non_iterated$realised_response, 3)
```

The anthesis-silking interval moves in the unfavourable direction in this
candidate set. Joukhadar et al. (2024) introduced their iterative search to
reduce this type of disagreement between a target and a selected-candidate
differential. This is a finite-candidate result. It does not contradict the
closed-form proportional-response identity under the covariance model.

See [Iterative optimisation of the desired-gain index](DesiredGainR-dgsi.html).

---

## 10. Stage 9 — Score genomic estimated breeding values

```{r qgsi}
quadratic_weights <- diag(c(
  GY = 0.05, PHT = -0.03, AD = -0.03, ASI = -0.04, EPP = 0.02, GLS = -0.03
))
dimnames(quadratic_weights) <- list(traits, traits)

qgsi <- run_qgsi(
  init_data = dgr_gebv["GenoID"], gebv_data = dgr_gebv,
  trait_cols = traits,
  linear_weights = economic_weights, W = quadratic_weights,
  lower_is_better = lower_is_better,
  scale_traits = TRUE, n_select = 20L
)
qgsi$expected_gain_per_trait
```

Use the same call with a zero `W` matrix as the linear genomic benchmark. Keep
all other inputs fixed. Compare per-trait response, rank correlation, selected
overlap, and validated nonlinear utility.

Standardisation matters more here than anywhere else in the package. On the
original scale, plant height carries half the linear index despite holding
nearly the smallest weight, and the expected gain in grain yield turns
*negative* while the index selects for more of it. `run_qgsi()` warns when the
trait scales make that likely.

See [Quadratic genomic selection](DesiredGainR-qgsi.html).

---

## 11. Stage 10 — Ask how much the objective matters

```{r sensitivity}
weight_sensitivity(
  economic_weights, dgr_candidates, dgr_G, dgr_P,
  n_select = 20L, trait_cols = traits, n_draws = 100L
)
```

Note that the weight the decision is most sensitive to is not the trait
carrying most of the index. Influence on the decision and contribution to the
index value are different quantities, and reading them together is the point.

---

## 12. What comes after

Two questions lie beyond a single cycle and beyond this package.

**Does this objective still look right in five cycles?** Truncation selection
erodes the genetic variance that response depends on, so a direction that
maximises first-cycle response can exhaust itself. The simulation layer
compares directions over successive cycles, with founders built from the
breeder's own phased marker data. See
[Comparing objectives over several cycles](DesiredGainR-simulation.html).

**Which candidates should actually be crossed?** Parent, cross and mating
allocation, coancestry control and operational feasibility are outside this
package's scope by design. Use
[HapBlockR](https://github.com/FAkohoue/HapBlockR) for parent selection, OCS,
optimum cross selection and mate allocation. HapBlockR's
`build_selection_index()` calls
DesiredGainR for DGSI or QGSI, so the index developed here passes directly into
that downstream decision. See
[Working with other breeding software](DesiredGainR-interoperation.html).

---

## 13. Where each stage is covered in depth

| Stage | Vignette |
|---|---|
| 1-2 | [Obtaining G and P](DesiredGainR-covariance.html) |
| 3-5 | [Defining a breeding objective](DesiredGainR-objective.html) |
| 6-7 | [Multiple-trait selection](DesiredGainR-index-families.html) |
| 8 | [Iterative optimisation of the desired-gain index](DesiredGainR-dgsi.html) |
| 9 | [Quadratic genomic selection](DesiredGainR-qgsi.html) |
| 10 | [Defining a breeding objective](DesiredGainR-objective.html) |
| Beyond | [Comparing objectives over several cycles](DesiredGainR-simulation.html) and [Working with other breeding software](DesiredGainR-interoperation.html) |

```{r session}
sessionInfo()
```
