---
title: "Using predictions, prediction errors, and restricted responses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using predictions, prediction errors, and restricted responses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(DesiredGainR)
data(dgr_traits, package = "DesiredGainR")
data(dgr_G, package = "DesiredGainR")
data(dgr_P, package = "DesiredGainR")
data(dgr_candidates, package = "DesiredGainR")
data(dgr_gebv, package = "DesiredGainR")
traits <- dgr_traits$trait
lower_is_better <- dgr_traits$trait[dgr_traits$direction == "decrease"]
```

## 1. Separate the evidence from the breeding objective

A selection index uses two vectors. The information vector contains the
measurements used to rank candidates. The objective vector contains the genetic
quantities that the breeder seeks to improve.

The two vectors often contain the same traits. In that case,
$\mathbf{C}=\mathbf{G}$. However, mixed-model and genomic analyses can provide
several sources of information for a smaller set of objective traits.

Define:

1. $\mathbf{P}=\operatorname{Var}(\mathbf{x})$, the covariance of the
   information vector.
2. $\mathbf{C}=\operatorname{Cov}(\mathbf{x},\mathbf{g})$, the covariance
   between information and objective traits.
3. $\mathbf{G}=\operatorname{Var}(\mathbf{g})$, the genetic covariance of the
   objective traits.

The three matrices must form a valid joint covariance:

$$
\begin{bmatrix}
\mathbf{P} & \mathbf{C}\\
\mathbf{C}^{\mathsf T} & \mathbf{G}
\end{bmatrix}.
$$

`selection_information()` checks this condition.

```{r information-model}
information <- selection_information(
  values = dgr_candidates,
  P = dgr_P,
  C = dgr_G,
  G = dgr_G
)
information
```

This example uses the same six traits in both vectors. Therefore it reproduces
the classical case.

## 2. Fit economic and desired-gain objectives

For economic weights $\mathbf{a}$, the general solution is

$$
\mathbf{b}=\mathbf{P}^{-1}\mathbf{C}\mathbf{a}.
$$

For desired gains $\mathbf{d}$, the solution is

$$
\mathbf{b}=
\mathbf{P}^{-1}\mathbf{C}
\left(
\mathbf{C}^{\mathsf T}\mathbf{P}^{-1}\mathbf{C}
\right)^{-1}\mathbf{d}.
$$

The expected response in the objective traits is

$$
\Delta\mathbf{g}
=
\frac{i\mathbf{C}^{\mathsf T}\mathbf{b}}
{\sqrt{\mathbf{b}^{\mathsf T}\mathbf{P}\mathbf{b}}},
$$

where $i$ is the standardised selection intensity.

```{r general-desired}
desired_gains_sd <- c(
  GY = 1.0, PHT = 0.4, AD = 0.6,
  ASI = 0.5, EPP = 0.4, GLS = 0.6
)
desired_gains <- desired_gains_sd * sqrt(diag(dgr_G))

general_dg <- generalized_index(
  information,
  objective = desired_gains,
  method = "desired_gain",
  n_select = 20L,
  lower_is_better = lower_is_better
)
general_dg
```

The coefficients refer to the information variables. The response refers to
the objective traits. These names can differ. The requested gains began in
genetic standard deviations. Multiplication by `sqrt(diag(dgr_G))` converted
them to the original trait units used by $\mathbf{P}$, $\mathbf{C}$, and
$\mathbf{G}$. The `lower_is_better` argument then oriented the reductions.

## 3. Use genomic predictions correctly

A genomic estimated breeding value (GEBV) is a prediction. Its covariance
describes the spread of predictions. The prediction error covariance (PEC)
describes the remaining uncertainty.

For calibrated best linear unbiased prediction (BLUP):

$$
\operatorname{Cov}(\widehat{\mathbf{g}},\mathbf{g})
=
\operatorname{Var}(\widehat{\mathbf{g}}).
$$

Hence a GEBV index can use:

1. $\mathbf{P}_{x}=\operatorname{Var}(\widehat{\mathbf{g}})$.
2. $\mathbf{C}=\operatorname{Var}(\widehat{\mathbf{g}})$.
3. $\mathbf{G}=\operatorname{Var}(\widehat{\mathbf{g}})
   +\overline{\mathbf{PEC}}$.

This identity requires calibrated predictions. Use cross-validation or a later
cohort to inspect calibration.

The full PEC remains valuable after fitting the index. Candidate-specific PEC
matrices provide score standard errors:

$$
\operatorname{SE}(I_i)
=
\sqrt{\mathbf{b}^{\mathsf T}\mathbf{PEC}_i\mathbf{b}}.
$$

```{r prediction-error}
example_pec <- diag(stats::setNames(rep(0.10, length(traits)), traits))

score_uncertainty <- candidate_score_uncertainty(
  general_dg,
  prediction_error_covariance = example_pec,
  n_draws = 1000L,
  seed = 91L
)
head(score_uncertainty$summary)
```

The selection probability shows how often a candidate enters the selected set
after prediction error is propagated. A candidate near the cut-off can carry a
wide score interval. That result helps the breeder decide where further
phenotyping has the greatest value.

## 4. Measure the value of each information source

Cunningham's deletion efficiency measures the contribution of each
information variable. It compares the full optimum index with the index formed
after deleting one variable.

```{r cunningham}
economic_weights <- c(
  GY = 1.0, PHT = 0.2, AD = 0.5,
  ASI = 0.4, EPP = 0.3, GLS = 0.5
)

smith_hazel <- selection_index(
  dgr_candidates,
  traits,
  method = "smith_hazel",
  G = dgr_G,
  P = dgr_P,
  economic_weights = economic_weights,
  scale_traits = TRUE,
  scale_by = "phenotypic",
  n_select = 20L
)

index_information_efficiency(smith_hazel)
```

An efficiency of 0.99 retains 99% of the full index standard deviation. Such a
variable may add little selection value. Its field cost can still justify
retention for another purpose. Therefore, combine this statistic with cost,
heritability, and programme needs.

## 5. Interpret restricted breeding values

Satoh (2024) showed that a restricted breeding value (RBV) is a projection of
an ordinary breeding value (BV). For a desired-gain direction $\mathbf{d}$:

$$
\mathbf{g}_{R}
=
\mathbf{d}
\left(
\mathbf{d}^{\mathsf T}\mathbf{G}^{-1}\mathbf{d}
\right)^{-1}
\mathbf{d}^{\mathsf T}\mathbf{G}^{-1}\mathbf{g}
=
\beta\mathbf{d}.
$$

The scalar $\beta$ has a direct interpretation. It measures progress along the
requested direction. Larger values indicate greater progress.

```{r satoh-rbv}
rbv <- restricted_breeding_values(
  dgr_gebv,
  G = dgr_G,
  direction = desired_gains,
  lower_is_better = lower_is_better
)
rbv
head(rbv$restricted)
head(rbv$beta)
```

Every projected vector follows the desired proportions. The candidate ranking
can therefore use $\beta$ when the objective is a complete proportional
direction.

An achieved population response can be evaluated in the same space.

```{r satoh-response}
response_check <- evaluate_restricted_response(
  response = general_dg$expected_response,
  direction = desired_gains,
  G = dgr_G,
  lower_is_better = lower_is_better
)
response_check
```

The output provides three complementary quantities:

1. `beta` measures progress along the desired-gain direction.
2. `mahalanobis_alignment` measures directional agreement.
3. `mahalanobis_residual` measures departure from the requested proportions.

## 6. Connect Yamada and restricted selection

Satoh (2024) proved that the Yamada desired-gain index is a special case of the
Kempthorne and Nordskog restricted index. A complete proportional direction
leaves a one-dimensional response space. Economic weights then have no effect
on the ranking.

```{r harville-satoh}
proportional <- restricted_index(
  dgr_candidates,
  traits,
  method = "harville",
  G = dgr_G,
  P = dgr_P,
  target_gains = desired_gains,
  lower_is_better = lower_is_better,
  scale_traits = FALSE,
  n_select = 20L
)

proportional$constraint$satoh_response
head(proportional$restricted_breeding_values$beta)
```

This result gives a direct bridge between desired-gain selection, restricted
selection, and simulation.

## 7. Use the Elston index as a careful comparator

The Elston multiplicative index ranks candidates by the product of their
margins above breeder-defined floors. DesiredGainR first applies every floor.
It then ranks eligible candidates. This ordering prevents a large value in two
traits from hiding a failed floor in another trait.

```{r elston}
elston <- selection_index(
  dgr_candidates,
  trait_cols = c("GY", "EPP"),
  method = "elston",
  culling_thresholds = c(GY = 0, EPP = 0),
  center_traits = TRUE,
  scale_traits = TRUE,
  n_select = 20L
)

head(elston$ranking)
```

The Elston index provides a useful weight-light comparator. Its response theory
is limited. Therefore, use the observed selected differential and simulation
when it changes the operational decision.

## 8. Practical sequence

The recommended sequence is:

1. Fit the multi-trait or multi-environment mixed model.
2. Export GEBVs, $\mathbf{G}$, and full PEC information.
3. Define $\mathbf{P}$, $\mathbf{C}$, and $\mathbf{G}$ explicitly.
4. Fit the economic or desired-gain index.
5. Propagate candidate-specific prediction error.
6. Inspect information deletion efficiency.
7. Evaluate restricted response with Satoh's criterion.
8. Validate the decision in a later cohort.

This sequence preserves the distinction between evidence, objective, expected
response, and uncertainty.

## References

- Beavis WD, Lamkey K, Mahama AA, Suza W (2023). Multiple Trait Selection.
  In *Quantitative Genetics for Plant Breeding*. Iowa State University Digital
  Press.
- Beavis WD, Mahama AA, Suza W (2023). Multi Environment Trials: Linear Mixed
  Models. In *Quantitative Genetics for Plant Breeding*. Iowa State University
  Digital Press.
- Cunningham EP (1969). The relative efficiencies of selection indexes.
  *Acta Agriculturae Scandinavica* **19**:45-48.
- Elston RC (1963). A weight-free index for ranking or selection with respect
  to several traits at a time. *Biometrics* **19**:85-97.
- Henderson CR, Quaas RL (1976). Multiple trait evaluation using relatives'
  records. *Journal of Animal Science* **43**:1188-1197.
- Satoh M (2024). Characteristics of restricted selection indices and
  geometrical interpretation of restricted breeding values. *Journal of
  Animal Breeding and Genetics* **141**:353-363.
  <https://doi.org/10.1111/jbg.12845>
