---
title: "Working with other breeding software"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with other breeding software}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
library(DesiredGainR)
data(dgr_traits, package = "DesiredGainR")
data(dgr_G, package = "DesiredGainR")
data(dgr_P, package = "DesiredGainR")
data(dgr_gebv, package = "DesiredGainR")
traits <- dgr_traits$trait
```

## 1. Where DesiredGainR enters the analysis

DesiredGainR begins after genetic evaluation. Mixed-model or genomic-prediction
software supplies candidate breeding values and, ideally, the fitted
multi-trait covariance matrices. DesiredGainR then defines and tests the
objective, builds selection indices, ranks candidates and reports uncertainty.

It does not analyse field designs, estimate marker effects, allocate crosses or
optimise matings. Keep those responsibilities in the specialised software that
produced or consumes the corresponding objects.

---

## 2. Minimum hand-off contract

Transfer the following objects explicitly:

1. one stable candidate identifier used by every table and matrix;
2. one GEBV or breeding-value column per trait;
3. trait names, units and favourable directions;
4. additive genetic covariance $\mathbf{G}$;
5. phenotypic or index-variable covariance $\mathbf{P}$;
6. prediction-error covariance or uncertainty draws when available; and
7. population, environment and model provenance.

Never match candidates or traits by position alone.

```{r identifiers}
stopifnot(!anyDuplicated(dgr_gebv$GenoID))
stopifnot(all(traits %in% names(dgr_gebv)))
stopifnot(identical(rownames(dgr_G), traits))
stopifnot(identical(colnames(dgr_G), traits))
```

---

## 3. Import covariance from upstream software

`import_covariance()` accepts matrices extracted from ASReml-R, `sommer`,
`breedR`, BGLR or a stage-wise workflow. The caller extracts the scientifically
correct model term; DesiredGainR validates its structure, orders the traits and
records its origin.

```{r import}
imported_G <- import_covariance(
  dgr_G[rev(traits), rev(traits)],
  trait_cols = traits,
  source = "matrix",
  estimand = "additive genetic covariance",
  P = dgr_P
)
imported_G
```

Because upstream interfaces change, passing the extracted covariance is more
stable than asking DesiredGainR to interpret an entire fitted model object.
The extraction script should remain with the analysis so the chosen random
term and covariance structure can be audited.

---

## 4. Pass phased genomes to the simulation layer

`haplotypes_from_inbred_dosage()` accepts inbred dosage data, while
`founder_haplotypes()` accepts phased homologues and a genetic map. Its input
shape is compatible with phased data returned by HapBlockR, but DesiredGainR
does not import HapBlockR and does not duplicate its haplotype analysis.

The resulting founder object can be passed to AlphaSimR through
`founder_population()`. Read
[Multi-cycle simulation](DesiredGainR-simulation.html) before interpreting a
simulated difference between objectives.

---

## 5. Hand selected candidates downstream

Export identifiers, ranks, scores, per-trait GEBVs and the exact objective used.
Do not send only a sorted list: mating-design software needs enough information
to apply diversity, relationship and crossing constraints that are outside the
selection index.

Also retain package versions, seeds, covariance provenance and the selected-set
stability analysis. A reproducible hand-off allows the next breeding stage to
distinguish a biological decision from a software default.

