---
title: "Empirical validation with real breeding-programme data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Empirical validation with real breeding-programme data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
validation_path <- function(name) {
  installed <- system.file("validation", name, package = "DesiredGainR")
  if (nzchar(installed)) {
    return(installed)
  }
  candidates <- c(
    file.path("..", "inst", "validation", name),
    file.path("inst", "validation", name)
  )
  found <- candidates[file.exists(candidates)]
  if (!length(found)) stop("Validation artifact not found: ", name)
  found[1]
}
results <- read.csv(
  validation_path("empirical-validation-summary.csv"),
  check.names = FALSE, stringsAsFactors = FALSE
)
```

## What the evidence establishes

DesiredGainR ships a reproducible real-data validation suite covering six
breeding-programme sources. The International Maize and Wheat Improvement
Center (CIMMYT) and Instituto Nacional de Investigación Agropecuaria (INIA)
are named in full at first use below. Together, these analyses demonstrate that the
package reproduces realised genetic-gain trajectories, recovers meaningful
genomic prediction, transports frozen multi-trait indices to later data, and
evaluates desired-gain directions in environments excluded from fitting. The
raw public data are too large for the package tarball, so the package ships the
compact results, evidence labels and interpretation guidance. The complete rerun is
`scripts/validate_empirical_programmes.R` in the source repository.

The suite keeps three questions separate:

1. Did observed population means change across real selection cycles?
2. Did an index frozen in an earlier period transport to later candidates,
   years or stages?
3. Did published predicted response agree in direction and scale with
   published realised response?

These complementary questions validate different layers of DesiredGainR. They
support its mathematical implementation, genomic-prediction foundation,
multi-environment transport and recommendation workflow. A future prospective
comparison assigning comparable populations to alternative vectors would add
direct estimates of incremental field response between strategies; it is a
next level of evidence, not a prerequisite for using the package's current
model-based decision support.

## Headline results

```{r headline, echo=FALSE}
headline_ids <- c(
  "cna6_yld_check_contrast",
  "hir_hif_connected_trend",
  "wheat_rcrgs_predicted_slope",
  "wheat_rcrgs_observed_slope",
  "inia_e1_advancement_auc",
  "inia_e1_auc_increment_over_yield",
  "inia_e1_advancement_enrichment",
  "g2f_index_transport_rho",
  "rcgs_gain_c0_c4_slope",
  "rcgs_raw_subsampled_gblup_cv_accuracy",
  "rcgs_nested_direction_worst_trait_increment"
)
headline <- results[
  match(headline_ids, results$validation_id),
  c(
    "dataset", "endpoint", "estimate", "lower", "upper",
    "unit", "status"
  )
]
headline$estimate <- round(headline$estimate, 3)
headline$lower <- round(headline$lower, 3)
headline$upper <- round(headline$upper, 3)
knitr::kable(headline, row.names = FALSE)
```

The evidence is informative because it distinguishes strong findings from
design-dependent ones. In CNA6 rice, the raw trial-mean yield trend was
positive, and the common-check sensitivity analysis provided the more precise
programme comparison. The
within-trial candidate-minus-check sensitivity analysis estimated about 235
kg/ha per cycle and excluded zero. Flowering and height trends remained
environment-dependent, as expected in a design where cycle and calendar year
coincide.

In the haploid-inducer maize population, the connected model used genotype
means and adjusted explicitly for calendar year. Real haploid induction
frequency (HIF) increased
increased by about 0.56 percentage points per cycle. The deposited data do not
identify the selected parents, so this validates the observed cycle trend, not
the historical index rule.

The CIMMYT wheat data are an external aggregate anchor. The predicted yield
trend from the published pedigree plus reproducing-kernel Hilbert-spaces with
kernel averaging model (P + RKHS-KA) was 0.158 t/ha per cycle;
the trend recomputed from the published two-year cycle means was 0.265 t/ha per
cycle. The raw yield plots and marker matrix are not in the local supplement,
so this is deliberately labelled a published aggregate rather than an
independent reanalysis.

## Out-of-time index validation

The INIA rice analysis used only records from the first evaluation stage (E1),
standardised within trial. A three-trait desired-gain index for grain yield, milling yield and head
rice recovery was fitted using cohorts available through 2008, frozen, and
then applied to 2009--2016 cohorts. Advancement within four years was known for
every E1 line.

The frozen multi-trait score had an area under the receiver operating
characteristic curve (AUC) of about 0.773 for advancement. Yield alone had an
AUC of about 0.756. A hierarchical bootstrap that resampled years and
candidates placed the AUC improvement near 0.017, with its 95% interval above
zero. The top 20% by the frozen index were about 1.77 times as likely to advance
as the full cohort. These are evidence of agreement with historical programme
decisions. They are not evidence that those decisions, or this particular
three-trait objective, were optimal.

Later-stage trait correlations are reported only among historically advanced
lines. Rejected lines have no later-stage phenotype, so presenting this as a
causal selected-versus-unselected comparison would be invalid.

The Genomes-to-Fields analysis froze an index on 2014--2017 environment-adjusted
means and evaluated the same genotypes in 2018--2021. Only 53 genotypes met the
predeclared requirement of at least two environments in both periods. The
early-to-late score correlation was about 0.56. Future yield response was
favourable, while anthesis and anthesis-silking interval responses showed the
expected environment sensitivity. This provides independent candidate-level
transport evidence that complements the recurrent-cycle datasets.

## CIMMYT rapid-cycle genomic selection

The rapid-cycle genomic selection (RCGS) deposit of Zhang, Pérez-Rodríguez,
Burgueño, Olsen, Jannink, Buckler, Atlin, Boddupalli, Vargas, San Vicente and
Crossa (2017) provides a
particularly useful combination: 1,000 base-cycle (C0) training families
evaluated in four environments, raw cycle-gain trials from C0 through the
fourth selection cycle (C4), 955,690 marker records, explicit
phenotype-to-sample keys,
and published genomic-prediction and diversity results.

The two-location raw-data model reproduced the published yield trajectory. It
estimated 0.175 t/ha per cycle from C0 to C4, compared with 0.158 in the paper,
and 0.253 t/ha per cycle from C1 to C4, compared with 0.225. C0--C3 adjusted
means were within 0.01 t/ha of the published table. The C4 deposit contains 44
candidate entries whereas the paper reports 43; retaining 43 rather than all
44 changed the adjusted mean by only about 0.001 t/ha, but the unidentified
entry discrepancy remains recorded.

The slower raw-genomic script streams the 1.9 gigabyte C0 haplotype-map
(HapMap) file, samples marker rows
deterministically across the genome, applies call-rate and minor-allele-frequency
filters, and constructs a VanRaden relationship matrix. Repeated five-fold
family prediction gave mean correlation 0.484 using 1,058 filtered sampled
markers. This is directionally and quantitatively concordant with the published
full-marker correlation of 0.55. The model was genomic best linear unbiased
prediction (GBLUP). A permuted-outcome negative control was 0.057.
This computationally bounded reproduction complements the paper's full
331,740-marker phenotype pipeline with an independently executable check.

For desired-gain direction validation, each outer fold withheld one complete
environment. Direction search occurred only within the other three environments,
using inner environment holdouts and a worst-trait criterion; the chosen vector
was then frozen. Across four outer folds, its average worst-trait improvement
over an equal-gain direction was 0.005 environmental standard deviations, with fold results from
-0.041 to 0.085. This is correctly labelled mixed. The joint high-gain event
occurred in one of four outer environments and its Brier score was 0.30. Four
outcomes make this a useful first external probability check rather than a
standalone calibration study. DesiredGainR therefore reports the probability
with its evidence level instead of overstating its precision.

The integrity audit found complete C0--C3 pedigree linkage and complete linkage
for all four training workbooks. Seventeen of 18 original-parent samples also
resolved through the supplied key. The one unresolved parent and the
byte-identical C3/C4 selected files are retained in the audit, consistent with
the paper's statement that C4 was not genotyped.

## Dataset attribution

The validation datasets are research outputs in their own right. Therefore,
the following citations name every author rather than replacing contributors
with *et al.*:

1. **CNA6 rice.** Jérôme Bartholomé, Adriano Castro, Flavio Breseghello,
   Isabela Furtini, Marley Utumi, José Almeida and Tuong-Vi Cao (2023).
   *Rice phenotypic data -- CNA6 population and advanced lines*. Dryad.
   <https://doi.org/10.5061/dryad.1g1jwsv28>
2. **Tropical maize haploid-inducer population.** Roberto Fritsche-Neto,
   Felipe Sabadin, Júlio César do Vale, Karina Borges, Pedro de Souza,
   José Crossa and Deoclécio Garbuglio (2023). *Realized genetic gains via
   recurrent selection in a tropical maize haploid inducer population and
   optimizing simultaneous selection for the next cycles*. *Crop Science*
   **63**:2865--2876. <https://doi.org/10.1002/csc2.21081>; data:
   <https://doi.org/10.5061/dryad.9p8cz8wns>
3. **CIMMYT spring wheat.** Susanne Dreisigacker, Paulino Pérez-Rodríguez,
   Leonardo Crespo-Herrera, Alison R. Bentley and José Crossa (2023).
   *Results from rapid-cycle recurrent genomic selection in spring bread
   wheat*. *G3: Genes|Genomes|Genetics* **13**:jkad025.
   <https://doi.org/10.1093/g3journal/jkad025>
4. **Instituto Nacional de Investigación Agropecuaria (INIA) rice.** Inés
   Rebollo, Sheila Scheffel, Pedro Blanco, Federico Molina, Sebastián Martínez,
   Gonzalo Carracelas, Ignacio Aguilar, Fernando Pérez de Vida and Juan E.
   Rosas (2023). *Consolidating 23 years of historical data from an irrigated
   subtropical rice breeding program in Uruguay*. *Crop Science*
   **63**:1300--1315. <https://doi.org/10.1002/csc2.20955>; data:
   <https://doi.org/10.5061/dryad.x69p8czn8>
5. **Genomes-to-Fields maize.** Marco Lopez-Cruz, Fernando M. Aguate,
   Jacob D. Washburn, Natalia de Leon, Shawn M. Kaeppler, Dayane Cristina Lima,
   Ruijuan Tan, Addie Thompson, Laurence Willard De La Bretonne and Gustavo de
   los Campos (2023). *Leveraging data from the Genomes-to-Fields Initiative
   to investigate genotype-by-environment interactions in maize in North
   America*. *Nature Communications* **14**:6904.
   <https://doi.org/10.1038/s41467-023-42687-4>; data:
   <https://doi.org/10.6084/m9.figshare.22776806.v1>
6. **CIMMYT tropical maize rapid-cycle genomic selection.** Xuecai Zhang,
   Paulino Pérez-Rodríguez, Juan Burgueño, Michael Olsen, Jean-Luc Jannink,
   Edward Buckler, Gary Atlin, M. Prasanna Boddupalli, Mateo Vargas,
   Félix San Vicente and José Crossa (2017). *Rapid cycling genomic selection
   in a multi-parental tropical maize population*. CIMMYT Research Data &
   Software Repository Network, version 2.
   <https://hdl.handle.net/11529/10927>. The associated article is by Xuecai
   Zhang, Paulino Pérez-Rodríguez, Juan Burgueño, Michael Olsen, Edward Buckler,
   Gary Atlin, Boddupalli M. Prasanna, Mateo Vargas, Félix San Vicente and José
   Crossa (2017). *G3: Genes|Genomes|Genetics* **7**:2315--2326.
   <https://doi.org/10.1534/g3.117.043141>

## Why the results are trustworthy

- Trials or genotypes are the independent units; plot rows are not treated as
  independent selection cycles.
- Every temporal index is frozen before the validation period.
- Adjusted-mean covariance is called a working surrogate, not additive genetic
  covariance.
- Common-check, yield-only and out-of-time comparisons were specified in the
  validation script rather than selected after seeing the result.
- Every result row carries an evidence tier, interpretation and appropriate
  scope of use.
- Haplotype-map and variant-call-format (VCF)-style headers are handled
  explicitly despite sharing the same
  `.hmp.txt` suffix in the RCGS deposit.
- Unphased bulk-family calls are not converted into fictitious phased inbred
  founders for AlphaSimR.
- The result table contains mixed and non-identifiable endpoints rather than
  filtering them out.

## Evidence interpretation and the next prospective study

```{r boundary, echo=FALSE}
extension <- results[
  results$validation_id == "prospective_vector_comparison_design",
  c("endpoint", "status", "interpretation", "limitation")
]
names(extension)[names(extension) == "limitation"] <- "prospective study design"
knitr::kable(extension, row.names = FALSE)
```

The current evidence supports DesiredGainR as rigorous, auditable decision
support: its mathematics, optimisation, genomic signal, cycle-gain
reproduction and out-of-sample recommendation behaviour have all been tested.
The probabilities are interpreted correctly as support under the fitted
genetic model and declared breeding scenario.

A prospective study can extend this evidence by assigning comparable
populations before selection to competing desired-gain vectors, holding
selection intensity, mating design and target environments constant, and
evaluating progeny in common trials. Such a study would estimate the
incremental field response among strategies and provide an additional layer of
programme-specific calibration. It does not negate the mathematical,
simulation and real-programme evidence already established here.
