---
title: "Full pipeline: from trait values to a defended selection decision"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Full pipeline: from trait values to a defended selection decision}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(DesiredGainR)
data(dgr_traits, package = "DesiredGainR")
data(dgr_G, package = "DesiredGainR")
data(dgr_P, package = "DesiredGainR")
data(dgr_candidates, package = "DesiredGainR")
data(dgr_gebv, package = "DesiredGainR")
data(dgr_history, package = "DesiredGainR")
data(dgr_hap1, package = "DesiredGainR")
data(dgr_hap2, package = "DesiredGainR")
data(dgr_map, package = "DesiredGainR")
```

## Purpose

This vignette assumes no prior knowledge of DesiredGainR. It runs the entire
package on the shipped example programme, one stage at a time, showing every
call, every output, and the decision each one feeds. Nothing is skipped and
nothing is left to be inferred from another document.

Sixteen stages in four phases:

| Phase | Stages | Question answered |
|---|---|---|
| **Evidence** | 1-4 | What do we have, and can it be trusted? |
| **Objective** | 5-8 | What should we select for, and is it possible? |
| **Index** | 9-12 | Which index, and what does it actually do? |
| **Looking ahead** | 13-16 | Does this objective still hold over several cycles? |

Every stage names the vignette that carries its full statistical detail. This
document is the connective walkthrough. Those are the references.

The walkthrough stops before the operational crossing decision. Throughout
this vignette, `n_select` is an analytical truncation count used to calculate
selection intensity and response, while `n_parents` and `n_crosses` in the
simulation are scenario settings. None of them is an optimised parent or mating
plan. Use [HapBlockR](https://github.com/FAkohoue/HapBlockR) after the index has
been built to choose parents, perform OCS, rank crosses and allocate matings.

---

# Phase 1: Evidence

## Stage 1 — What you need before you start

DesiredGainR begins *after* the genetic evaluation. It does not analyse field
trials, does not fit a mixed model, and does not estimate breeding values. You
must arrive with four things.

| Input | What it is | Where it comes from |
|---|---|---|
| Trait values | One adjusted mean, best linear unbiased prediction (BLUP) or genomic estimated breeding value (GEBV) per candidate per trait | Your multi-trait genetic evaluation |
| $\mathbf{G}$ | Genetic variance-covariance matrix | The same fitted model |
| $\mathbf{P}$ | Phenotypic variance-covariance matrix | The same fitted model |
| Trait directions | Which traits improve by rising, which by falling | The product profile |

If you have no fitted $\mathbf{G}$, Stage 3 shows the supported
approximations. If you have no trait directions written down, stop and write
them down: every later stage depends on them.

## Stage 2 — Load and inspect the data

```{r stage2}
dgr_traits
```

Read this table before anything else. It carries the units, the direction of
improvement, the heritability and the genetic standard deviation for each
trait. Two columns decide most of what follows.

```{r stage2-directions}
traits <- dgr_traits$trait
lower_is_better <- dgr_traits$trait[dgr_traits$direction == "decrease"]
lower_is_better
```

`lower_is_better` is passed to nearly every function in the package. Supplying
it means you state gains as *improvements* everywhere, and the package handles
the sign conventions internally. Signing values by hand is the commonest source
of a confidently wrong answer.

```{r stage2-candidates}
str(dgr_candidates, max.level = 1)
head(dgr_candidates[, c("GenoID", "GY", "PHT", "GLS")], 3)
```

## Stage 3 — Check the covariance matrices

Every index in this package is computed by inverting $\mathbf{P}$,
$\mathbf{G}$, or both. An ill-conditioned matrix produces coefficients that are
numerically meaningless while looking entirely plausible, so this check comes
before anything is built on top of them.

```{r stage3}
diagnostics_G <- matrix_diagnostics(dgr_G, "G")
c(
  condition_number = diagnostics_G$condition_number,
  positive_definite = diagnostics_G$positive_definite,
  numerical_rank = diagnostics_G$numerical_rank
)
```

A condition number near $1.9 \times 10^{4}$ looks alarming until it is
decomposed. Almost all of it comes from the trait scales rather than from the
correlations:

```{r stage3-decompose}
matrix_diagnostics(stats::cov2cor(dgr_G))$condition_number
```

Standardising the traits removes three orders of magnitude of ill-conditioning.
This is the practical form of an observation by Crosbie et al. (1980): equal
weights applied to unstandardised traits concentrate selection on whichever
trait happens to carry the largest variance. It is why DesiredGainR
standardises by default.

**If you have no fitted $\mathbf{G}$.** Use
`estimate_genetic_covariance()`, which returns labelled approximations rather
than relabelling a covariance of predictions as a genetic covariance:

```{r stage3-estimate}
G_working <- estimate_genetic_covariance(
  values = dgr_candidates, trait_cols = traits,
  method = "adjusted_means_surrogate"
)
G_working$estimand
```

Four methods are available: `prediction_covariance`, `pev_corrected`,
`se_diagonal_corrected`, `relationship_adjusted`, and
`adjusted_means_surrogate`. The last implements what the CGIAR implementation
guideline recommends for operational use when no fitted matrix exists.

## Stage 4 — Confirm units and heritabilities

The declared heritabilities must be recoverable from the two matrices. If they
are not, the metadata and the matrices are describing different things.

```{r stage4}
round(diag(dgr_G) / diag(dgr_P), 2)
dgr_traits$heritability
```

Check the spread of trait scales, because it determines whether standardisation
is optional or essential:

```{r stage4-scales}
genetic_sd <- stats::setNames(dgr_traits$genetic_sd, dgr_traits$trait)
round(sort(genetic_sd, decreasing = TRUE), 2)
round(max(genetic_sd) / min(genetic_sd))
```

A spread of 120-fold means standardisation is essential here. The package warns
automatically above fivefold.

---

# Phase 2: Objective

## Stage 5 — State a first objective

There are two ways to state a breeding objective, and they are equivalent.

**Economic weights** $\mathbf{w}$ say what a unit of each trait is worth.
**Desired gains** $\mathbf{d}$ say how much of each trait you want. The
Smith-Hazel index uses the first, $\mathbf{b} = \mathbf{P}^{-1}\mathbf{G}\mathbf{w}$.
the Pesek-Baker desired-gain index uses the second,
$\mathbf{b} = \mathbf{P}^{-1}\mathbf{G}(\mathbf{G}\mathbf{P}^{-1}\mathbf{G})^{-1}\mathbf{d}$.

Most breeders find desired gains easier, so start there. State them as
improvements, in genetic standard deviations, and let `lower_is_better` handle
direction.

```{r stage5}
desired_gains <- c(
  GY = 1.0, PHT = 0.4, AD = 0.6, ASI = 0.5,
  EPP = 0.4, GLS = 0.6
)
```

This says: one standard deviation of extra yield, four tenths of a standard
deviation shorter plants, six tenths earlier flowering, and so on.

## Stage 6 — Translate to economic weights

Setting the two index formulae equal gives an exact translation in both
directions:

$$\mathbf{w} = \mathbf{G}^{-1}\mathbf{P}\mathbf{G}^{-1}\mathbf{d},
\qquad
\mathbf{d} = \mathbf{G}\mathbf{P}^{-1}\mathbf{G}\mathbf{w}.$$

```{r stage6}
implied <- implied_economic_weights(
  desired_gains, dgr_G, dgr_P,
  lower_is_better = lower_is_better, gain_units = "genetic_sd"
)
round(implied, 2)
```

**Two of these are negative, for traits you asked to improve.** That is
correct. Once oriented, grain yield correlates $+0.55$ with the
anthesis-silking interval and $+0.45$ with ears per plant, so selecting hard
for yield would carry both *past* the smaller gains you requested. To deliver
the ratio you asked for, the index must hold them back. A negative weight never
means the trait should get worse.

**The desired gains are already standardised. The returned weights use a
different scale.**
Here `desired_gains` is expressed in genetic standard deviations, but `dgr_G`
and `dgr_P` are supplied in the original trait units. Therefore,
`implied_economic_weights()` first converts the target to those original units
and returns a weight per original unit of each trait. Compare coefficients only
after conversion to a common trait scale.

The multiplication below does not standardise `desired_gains` a second time.
It converts each returned raw-unit weight into the contribution associated with
one genetic standard deviation of its trait. These products are on a common
genetic-standard-deviation scale and may be compared as a diagnostic:

```{r stage6-rescale}
round(sort(implied * genetic_sd[names(implied)], decreasing = TRUE), 2)
```

Ears per plant had the largest raw coefficient but one of the smaller
per-genetic-standard-deviation effects. If `G` and `P` had already been
standardised to genetic-standard-deviation units, the returned weights would
already be per genetic standard deviation and this multiplication would be
wrong. Standardise either the covariance space or the returned weights, not
both.

Confirm the translation inverts, which costs one line and catches unit errors:

```{r stage6-invert}
round(
  implied_desired_gains(
    implied, dgr_G, dgr_P,
    lower_is_better = lower_is_better, gain_units = "genetic_sd"
  ),
  3
)
```

## Stage 7 — Test whether the objective is attainable

**This is the stage most often skipped, and the one that most often changes the
answer.**

Let $\mathbf{b}$ denote the linear-index coefficients, $\mathbf{G}$ the
additive genetic covariance matrix, $\mathbf{P}$ the phenotypic covariance
matrix, and $i$ the standardised selection intensity. The expected one-cycle
response is

$$\mathbf{R} =
i\,\frac{\mathbf{G}\mathbf{b}}
{\sqrt{\mathbf{b}^\mathsf{T}\mathbf{P}\mathbf{b}}}.$$

Consequently, every response available at intensity $i$ lies on the
achievable-response ellipsoid

$$\mathbf{R}^\mathsf{T}\mathbf{G}^{-1}\mathbf{P}
\mathbf{G}^{-1}\mathbf{R}=i^2.$$

A specified vector $\mathbf{d}$ lies on an ellipsoid whose intensity is

$$i_{\text{required}} = \sqrt{\mathbf{d}^\mathsf{T}\mathbf{G}^{-1}\mathbf{P}\mathbf{G}^{-1}\mathbf{d}}.$$

```{r stage7}
feasibility <- gain_feasibility(
  desired_gains, dgr_G, dgr_P,
  n_candidates = nrow(dgr_candidates), n_select = 20L,
  lower_is_better = lower_is_better, gain_units = "genetic_sd"
)
feasibility
```

Read the output line by line. The required intensity of 3.2082 corresponds,
under normal truncation selection, to retaining the best 0.1773% of an
effectively continuous population. In a population of 200 candidates, this is
0.355 candidate. A breeder cannot select a fraction of a candidate. Even
selecting the single best candidate retains 0.5%, which gives a lower intensity
than 3.2082. Therefore, the exact vector is unattainable in this candidate set.

The planned decision retains 20 of 200 candidates, or 10%. Its intensity is
1.7550. The ratio $1.7550/3.2082=0.547$ means that 54.7% of the requested vector
can be delivered while preserving its exact trait proportions:

```{r stage7-attainable}
round(feasibility$attainable_response_input_units, 3)
```

The returned values are in the same genetic-standard-deviation units as the
input. Thus, the planned exact-ray response is 0.547 standard deviations for
grain yield, 0.219 for plant height, 0.328 for anthesis date, 0.274 for the
anthesis-silking interval, 0.219 for ears per plant, and 0.328 for grey leaf
spot resistance. They are not observed gains and they are not six independent
upper limits. They are the model-based response at the planned intensity when
the requested **ratio among traits is held exactly**.

This interpretation has four consequences.

1. The classical desired-gain index honours the direction, or ratio, of
   $\mathbf{d}$. Multiplying all elements by the same positive constant leaves
   the ranking unchanged. It changes only the intensity required to attain the
   stated magnitude.
2. `feasible_at_planned_intensity = no` still allows the index to rank
   candidates. The complete stated vector lies beyond one-cycle response at
   the planned 10% selection proportion under the
   supplied $\mathbf{G}$ and $\mathbf{P}$.
3. An exact desired-gain vector is not the same objective as a set of minimum
   gains. For example, another attainable response may exceed the grain-yield
   target and still meet every minimum without preserving the ratio
   $1.0:0.4:0.6:0.5:0.4:0.6$. `gain_feasibility()` tests the exact ray.
   Population-driven and interval searches assess trait-specific floors and
   must be interpreted through their worst-trait and joint-attainment results.
4. The result is conditional on a linear index, one-cycle truncation selection,
   approximate normality, and the supplied covariance matrices. Therefore,
   covariance uncertainty and selected-set stability must be checked before a
   recommendation is treated as robust.

If the ratio itself is biologically required, scale the whole vector by 0.547
or relax the selection proportion. If each entry is a minimum rather than a
fixed ratio, use `suggest_desired_gains()` or interval optimisation and require
the lower confidence bound for joint attainment to support the recommendation.

## Stage 8 — Recover what the programme already selects for, then adjust

If your programme has been selecting without a formal index, the differentials
it achieved already encode its objective:

$$\mathbf{b} = \mathbf{P}^{-1}\mathbf{s}$$

where $\mathbf{s}$ holds the differences between the selected group and the
whole population.

```{r stage8}
recovered <- retrospective_weights(
  selected_values = dgr_candidates[dgr_history$selected, traits],
  population_values = dgr_candidates[, traits],
  trait_cols = traits
)
recovered
```

```{r stage8-differentials}
round(recovered$selection_differential_sd, 2)
```

This is how an index was introduced at both the International Maize and Wheat
Improvement Center (CIMMYT) and the International Rice Research Institute
(IRRI). Treat it as a starting point: it reproduces past decisions including any bias in them, and
must be adjusted deliberately against the product profile before it becomes a
forward-looking objective.

---

# Phase 3: Index

## Stage 9 — Build candidate indices

`selection_index()` provides every classical family through one interface.

```{r stage9}
economic_weights <- c(
  GY = 1.0, PHT = 0.2, AD = 0.5,
  ASI = 0.4, EPP = 0.3, GLS = 0.5
)

smith_hazel <- selection_index(
  dgr_candidates, traits,
  method = "smith_hazel",
  G = dgr_G, P = dgr_P, economic_weights = economic_weights,
  lower_is_better = lower_is_better, n_select = 20L, main_trait = "GY"
)
rank_sum <- selection_index(
  dgr_candidates, traits,
  method = "mulamba_mock",
  lower_is_better = lower_is_better, n_select = 20L
)
base_index <- selection_index(
  dgr_candidates, traits,
  method = "base",
  G = dgr_G, P = dgr_P, economic_weights = economic_weights,
  lower_is_better = lower_is_better, n_select = 20L, main_trait = "GY"
)
smith_hazel
```

| Method | Needs weights | Needs $\mathbf{G}$, $\mathbf{P}$ | Use when |
|---|---|---|---|
| `smith_hazel` | yes | yes | Weights are defensible |
| `base` | yes | no | A quick comparator. It is often nearly as good. |
| `pesek_baker` | desired gains | $\mathbf{G}$ only | Gains easier than weights |
| `yamada` | desired gains | yes | Same index, better conditioned route |
| `mulamba_mock` | no | no | Weights unavailable or untrusted |
| `independent_culling` | thresholds | no | Non-compensatory requirements |
| `elston` | thresholds | no | Firm limits followed by balanced margins |
| `tandem` | trait order | no | Comparator for sequential selection |
| `restricted_index()` | weights and restrictions | yes | One or more responses must be controlled |
| `run_dgsi()` | desired gains | yes | Iterative selected-differential search is required |
| `run_qgsi()` | linear and quadratic values | genomic covariance | Merit is curved or interactive |

## Stage 10 — Evaluate and compare

`evaluate_index()` reports the classical criteria used by Rahimi and Debnath
(2023), together with the response of every trait. Each quantity answers a
different question.

| Criterion | Full definition | Valid interpretation |
|---|---|---|
| $R_{HI}$ | Correlation between the selection index $I$ and aggregate genetic merit $H$ | It ranges from -1 to 1. Negative values select against the stated merit, zero means no linear association, and values nearer 1 indicate closer agreement. It exists only when the same aggregate-weight vector defines $H$ for every index compared. |
| $\Delta H$ | Expected change in aggregate genetic merit | This is $iR_{HI}\sigma_H$, where $\sigma_H$ is the genetic standard deviation of aggregate merit. Positive values improve the stated merit. Compare magnitudes only when all indices use the same merit definition, population and selection intensity. |
| $\Delta_j$ | Expected genetic response of trait $j$ | This is the primary biological result. Check its sign, magnitude and units for every trait. An aggregate statistic can conceal an unfavourable response. |
| RE | Relative efficiency for the declared main trait | RE = 1 matches direct response, 0.80 retains 80%, a value above 1 exceeds direct response through correlated information, and a negative value moves the main trait unfavourably. It says nothing about the other traits. |
| $h_I^2$ | Heritability of the index score treated as a composite trait | It ranges from 0 to 1 for compatible covariance matrices. Its square root is the accuracy with which the observed index predicts its own additive genetic component. It does not measure agreement with the breeding objective. |
| $CV_I$ | Coefficient of variation of index scores | This is $100s_I/|\bar I|$. It depends on the arbitrary zero of the score and is undefined for a centred index. Use it only for raw, uncentred scores. |

The software output name `delta_H` and the abbreviated print label `dH` both
mean $\Delta H$, the expected change in aggregate genetic merit. They do not
mean the response of an individual trait.

No classical scalar is sufficiently robust on its own. Start with $\Delta_j$
for every trait. Then test exact-ray feasibility. For minimum floors, inspect
the worst-trait and joint-attainment probabilities. Measure response alignment
with `index_uncertainty()`. Inspect rank correlation and selected-set overlap
under perturbation. For repeated cycles, finish with diversity or coancestry
diagnostics. This panel covers biological attainment, statistical stability,
and the retention of usable variation.

**Relative efficiency below 1 is not a failure.** It is the intended trade of
response in the main trait for response elsewhere. Every value reported by
Rahimi and Debnath was below 1. Conversely, a high relative efficiency is not
proof that the desired multi-trait objective was attained.

Use `compare_selection_methods()` to verify the common comparison conditions.
Then compare responses, summary criteria, rankings, and selected sets.

```{r stage10}
comparison <- compare_selection_methods(
  list(
    Smith_Hazel = smith_hazel,
    Base = base_index,
    Rank_sum = rank_sum
  )
)
comparison$fairness
comparison$summary
round(comparison$rank_correlation, 2)
comparison$selected_overlap
```

Rank correlation uses every candidate. Selected-set overlap concerns the final
decision. Report both because close overall rankings can diverge near the
selection boundary.

Check which traits the index is really acting on:

```{r stage10-effective}
smith_hazel$effective_weights[, c("Trait", "Genetic_share")]
```

## Stage 11 — Optimise the desired-gain vector

`run_dgsi()` implements the iterative method of Joukhadar et al. (2024). It
samples desired-gain vectors, builds the index from each, and keeps the one
whose *realised* response in the selected set comes closest to your target.

```{r stage11}
dgsi <- run_dgsi(
  init_data = dgr_candidates[, c("GenoID", "Family")],
  cand_data = dgr_candidates,
  trait_cols = traits,
  dg = desired_gains,
  G = dgr_G, P = dgr_P,
  lower_is_better = lower_is_better,
  scale_traits = TRUE,
  n_select = 20L, n_iter = 200L, n_rep = 5L, seed = 42L
)
round(dgsi$realised_response, 3)
```

Read the deviation from target against Stage 7:

```{r stage11-deviation}
round(dgsi$realised_response - desired_gains, 3)
```

Feasibility constrained the requested *proportion*. The optimiser is not bound
to that proportion, so it distributes the shortfall unevenly, overshooting
grain yield while undershooting anthesis date. Whether that trade is acceptable
is a breeding decision, not a numerical one.

Always inspect the replicate diagnostics. The search is stochastic, so
agreement across replicates is the evidence that the answer is stable:

```{r stage11-replicates}
dgsi$replicate_diagnostics
```

And compare against the classical solution without iteration:

```{r stage11-comparator}
round(dgsi$non_iterated$realised_response, 3)
```

The anthesis-silking interval moves the *wrong way* there. Joukhadar et al.
reported precisely this failure of the unoptimised method.

## Stage 12 — Score genomic estimated breeding values

`run_qgsi()` implements the quadratic genomic selection index of Cerón-Rojas et
al. (2026):

$$I_{qg,i} = \mathbf{w}^{\mathsf T}\widehat{\mathbf{g}}_i +
\widehat{\mathbf{g}}_i^{\mathsf T}\mathbf{W}\widehat{\mathbf{g}}_i$$

where $\widehat{\mathbf{g}}_i$ is the candidate's vector of genomic
estimated breeding values, $\mathbf{w}$ the linear economic weights, and
$\mathbf{W}$ a symmetric matrix of squared and cross-product weights. Negative
diagonal curvature in $\mathbf{W}$ favours intermediate values. Positive
curvature favours extremes.

```{r stage12}
quadratic_weights <- diag(c(
  GY = 0.05, PHT = -0.03, AD = -0.03, ASI = -0.04, EPP = 0.02, GLS = -0.03
))
dimnames(quadratic_weights) <- list(traits, traits)

qgsi <- run_qgsi(
  init_data = dgr_gebv["GenoID"], gebv_data = dgr_gebv,
  trait_cols = traits,
  linear_weights = economic_weights, W = quadratic_weights,
  lower_is_better = lower_is_better,
  scale_traits = TRUE, n_select = 20L
)
qgsi$expected_gain_per_trait
```

**Standardisation matters more here than anywhere else in the package**,
because economic weights multiply the breeding values directly. Left
unstandardised on these data, plant height carries half the linear index
despite holding nearly the smallest weight, and the expected gain in grain
yield comes out *negative* while the index selects for more of it.
`run_qgsi()` warns when the trait scales make this likely.

---

# Phase 4: Looking ahead

A single-cycle calculation cannot distinguish desired-gain directions beyond
what the feasibility ellipsoid already tells you. Over several cycles the
ranking can change, because truncation selection erodes the genetic variance
that the response depends on. Phase 4 represents that.

The simulation is calibrated to your own germplasm: founders are built from
real phased marker data, never from a simulated genome.

## Stage 13 — Build founders from phased marker data

```{r stage13}
founders <- founder_haplotypes(dgr_hap1, dgr_hap2, dgr_map)
founders
```

`hap1` and `hap2` are marker-by-individual matrices coded 0 or 1, giving the
allele on each chromosome copy. This is what a phase-preserving variant call
format reader returns.

**Dosage coded 0, 1, 2 is not accepted**, because it records how many
alternative alleles an individual carries but not which copy carries them, and
that co-occurrence is the linkage disequilibrium the simulation exists to
represent.

For highly inbred diploid material, dosage 0 and 2 are unambiguous, so phase
can be derived without external phasing. Measure the heterozygosity first:

```{r stage13-dosage}
dosage <- dgr_hap1 + dgr_hap2
diagnostics <- dosage_diagnostics(dosage)
c(
  heterozygosity = diagnostics$overall_heterozygosity,
  missing = diagnostics$overall_missing
)
```

```{r stage13-convert, eval=FALSE}
converted <- haplotypes_from_inbred_dosage(
  dosage,
  heterozygous_policy = "drop_variant"
)
```

No threshold is imposed on residual heterozygosity, because no universally
appropriate level exists. It is measured and reported instead. Heterozygous
calls are never resolved silently.

## Stage 14 — Calibrate the simulation

Genome structure comes from your markers. Trait architecture is calibrated to
the covariance matrix you already estimated, so the simulation reproduces both.
Quantitative trait loci (QTL) are sampled on that structure for the simulation
model.

```{r stage14, eval=FALSE}
setup <- founder_population(
  founders,
  G = dgr_G,
  h2 = stats::setNames(dgr_traits$heritability, traits),
  n_qtl_per_chromosome = 100L,
  seed = 42L
)
setup
```

```
#> <desiredgainr_sim_setup>
#>   Founders: 200 individuals, 3 chromosomes, 100 QTL per chromosome
#>   Traits: GY, PHT, AD, ASI, EPP, GLS
#>   Dominance simulated: no
```

For a **clonal** programme add `dominance_degree`, and supply the *genotypic*
rather than the additive covariance as `G`, because the clone inherits
dominance intact.

## Stage 15 — Simulate one objective across cycles

```{r stage15, eval=FALSE}
simulation <- simulate_selection_cycles(
  setup,
  desired_gains = desired_gains,
  n_cycles = 5L,
  mating_system = "outcross",
  n_parents = 20L, n_crosses = 50L, n_progeny_per_cross = 10L,
  lower_is_better = lower_is_better,
  seed = 11L
)
simulation
```

```
#> <desiredgainr_simulation>
#>   outcross system, 5 cycles, 20 parents recycled
#>   Index re-estimated each cycle: yes
#>   Cumulative genetic gain:
#>       GY     PHT      AD     ASI     EPP     GLS
#>    1.842  -8.031  -1.226  -0.244  -0.019  -0.487
#>   Final mean relationship among parents: 0.0913
```

Output above is illustrative of the object's shape. Your numbers depend on your
founders and parameters.

Three mating systems are supported, and they differ in more than naming:

| System | Crops | What differs |
|---|---|---|
| `"self"` | Wheat, rice, bean | Advanced by selfing or doubled haploidy. Recombination releases variance slowly. |
| `"outcross"` | Maize, sorghum, millet | Random mating occurs each cycle. Half the selection-induced disequilibrium decays per generation. |
| `"clonal"` | Cassava, sweetpotato, banana | Selection on *total* genetic value, because dominance is inherited intact |

The per-cycle table records the genetic mean and variance for each trait, the
mean relationship among selected parents, and the implied effective population
size, so that gain and the loss of diversity can be read together.

This is a scenario comparison, not OCS or optimum cross selection. The
simulation asks whether one desired-gain objective remains attractive under a
declared reproduction rule. It does not claim that its recycled parents or
crosses are the operational optimum. Pass the index merit score to HapBlockR
when the real parent, contribution and mating decisions are made.

`reestimate_index = TRUE`, the default, rebuilds the index from each cycle's own
data, which is what a programme actually does and which propagates estimation
error across cycles. Setting it `FALSE` isolates the effect of the
desired-gain direction alone. A large divergence between the two means the
recommendation is sensitive to estimation error.

## Stage 16 — Search for the best objective over cycles

Because only the *direction* of $\mathbf{d}$ matters, the search space is the
unit sphere with $p-1$ free parameters for $p$ traits — small enough to cover
densely.

```{r stage16, eval=FALSE}
optimisation <- optimize_desired_gains(
  setup,
  n_cycles = 5L,
  mode = "pareto",
  budget = 60L,
  n_replicates = 3L,
  mating_system = "outcross",
  n_parents = 20L, n_crosses = 50L, n_progeny_per_cross = 10L,
  lower_is_better = lower_is_better,
  checkpoint = "search.rds",
  seed = 42L
)
optimisation
```

```
#> <desiredgainr_optimisation>
#>   Mode: pareto   Cycles: 5   Evaluations: 60 (3 replicates each)
#>   Objectives: GY, PHT, AD, ASI, EPP, GLS, diversity
#>   Non-dominated directions: 9 of 60
#>   No single direction is recommended. Choose a point on the frontier.
```

**The simulation is the objective function and is never replaced by a cheaper
approximation.** A Gaussian process decides only where to spend the next
simulation. It never filters or substitutes for one. Because the
acquisition function keeps exploring, no region can be permanently excluded.

Four ranking modes are available:

| Mode | Objective | Needs |
|---|---|---|
| `"pareto"` | Non-dominated outcomes | nothing beyond the setup |
| `"economic"` | Maximise $\mathbf{w}^\mathsf{T}\mathbf{R}$ | `economic_weights` |
| `"target"` | Minimise distance to absolute targets | `target_gains` |
| `"constrained"` | Maximise one trait subject to floors | `focal_trait`, `gain_floors` |

`"pareto"` is the default because it is the only mode that does not require the
weights breeders find hardest to state. Choosing a point on a frontier is an
easier judgement, and it is weight elicitation by revealed preference.

A realistic budget takes hours, so supply `checkpoint` to make the search
resumable.

---

## Sign-off checklist

Complete this before promoting any output to a breeding recommendation.

| Check | Where it came from |
|---|---|
| Covariance matrices positive definite, conditioning inspected | Stage 3 |
| Heritabilities consistent with $\mathbf{G}$ and $\mathbf{P}$ | Stage 4 |
| Trait directions declared, not signed by hand | Stage 2 |
| Objective feasible at the planned intensity, or knowingly not | Stage 7 |
| Index family chosen and justified against a comparator | Stages 9-10 |
| Effective weights inspected. No unintended trait dominance | Stage 10 |
| Replicate diagnostics agree | Stage 11 |
| Traits standardised, or the warning read and accepted | Stages 3, 12 |
| Decision robust to plausible weight perturbation | *Objective vignette* |
| Final parents, contributions and crosses decided in HapBlockR, not inferred from `n_select` | [Working with other breeding software](DesiredGainR-interoperation.html) |
| Random seed, package version and session recorded | below |

```{r session}
packageVersion("DesiredGainR")
sessionInfo()
```

---

## Where each stage is covered in depth

| Stage | Vignette |
|---|---|
| 1-4 | [Obtaining G and P](DesiredGainR-covariance.html) |
| 5-8 | [Defining a breeding objective](DesiredGainR-objective.html) |
| 9-10 | [Multiple-trait selection](DesiredGainR-index-families.html) |
| 11 | [Optimising desired gains](DesiredGainR-dgsi.html) |
| 12 | [Quadratic genomic selection](DesiredGainR-qgsi.html) |
| 13-16 | [Comparing objectives over several cycles](DesiredGainR-simulation.html) |
| Downstream decision | [Working with other breeding software](DesiredGainR-interoperation.html) |

---

## References

- Cerón-Rojas JJ, Montesinos-López OA, Montesinos-López A, et al. (2026).
  Nonlinear genomic selection index accelerates multi-trait crop improvement.
  *Nature Communications* **17**:1991.
- Covarrubias-Pazaran G (2021). *Practical implementation of selection
  indices.* CGIAR Excellence in Breeding.
- Guimarães PHR et al. (2021). Index selection can improve the selection
  efficiency in a rice recurrent selection population. *Euphytica* **217**:95.
- Joukhadar R et al. (2024). Optimising desired gain indices to maximise
  selection response. *Frontiers in Plant Science* **15**:1337388.
- Rahimi M, Debnath S (2023). Estimating optimum and base selection indices.
  *Scientific Reports* **13**:18977.
