---
title: "Quadratic genomic selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quadratic genomic selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(DesiredGainR)
data(dgr_traits, package = "DesiredGainR")
data(dgr_G, package = "DesiredGainR")
data(dgr_gebv, package = "DesiredGainR")
traits <- dgr_traits$trait
lower_is_better <- dgr_traits$trait[dgr_traits$direction == "decrease"]
```

## 1. The decision that QGSI represents

The quadratic genomic selection index (QGSI) is useful when a straight line
gives an inadequate representation of a trait combination's value. Three
examples are an intermediate optimum, diminishing benefit beyond a useful
level, and complementarity between two traits. If every additional unit has
the same value, a linear index is clearer and should be preferred.

For candidate $i$, DesiredGainR calculates

$$
I_{qg,i}=\mathbf{w}^{\mathsf T}\widehat{\mathbf{g}}_i+
\widehat{\mathbf{g}}_i^{\mathsf T}\mathbf{W}
\widehat{\mathbf{g}}_i,
$$

where $\widehat{\mathbf{g}}_i$ contains genomic estimated breeding values
(GEBVs), $\mathbf{w}$ contains linear economic weights, and $\mathbf{W}$ is a
symmetric matrix of squared and cross-product economic weights. The package
ranks larger QGSI scores first.

**Desired gains are not QGSI weights.** A desired gain states the response a
breeder wants. A QGSI weight states the value assigned to a candidate's linear,
squared or cross-product genetic merit. DesiredGainR never substitutes one for
the other.

---

## 2. Translate the biology into weights

For one standardised, favourable-direction trait $x_j$, its contribution is

$$u_j(x_j)=w_jx_j+W_{jj}x_j^2.$$

The signs have precise meanings.

| Specification | Interpretation |
|---|---|
| $W_{jj}=0$ | The trait has a linear value. |
| $W_{jj}<0$ | The value is concave: improvement has diminishing benefit and eventually reaches an optimum. |
| $W_{jj}>0$ | Extreme values in either direction receive increasing value. Use this only when both extremes are genuinely valuable. |
| $W_{jk}>0$ | Same-sign deviations are rewarded and opposite-sign deviations are penalised. This includes both the jointly favourable and jointly unfavourable quadrants. |
| $W_{jk}<0$ | Opposite-sign deviations are rewarded and same-sign deviations are penalised. |

When $W_{jj}<0$ and $w_j>0$, the single-trait optimum is
$-w_j/(2W_{jj})$ standard deviations above the reference mean. This identity
provides a defensible way to set curvature. For example, $w_j=0.40$ and
$W_{jj}=-0.10$ place the optimum at two standard deviations.

For an off-diagonal entry, the coefficient of $x_jx_k$ in the score is
$2W_{jk}$ because a symmetric quadratic form contains both $W_{jk}$ and
$W_{kj}$. Enter half of the intended cross-product coefficient in each of the
two symmetric cells.

---

## 3. Prepare the GEBVs

The example contains 200 candidates and six traits. Grain yield and ears per
plant improve by increasing. Plant height, flowering time, anthesis-silking
interval and grey leaf spot severity improve by decreasing.

```{r inspect-data}
dgr_traits[, c("trait", "description", "direction", "unit")]
head(dgr_gebv[, c("GenoID", traits)])
```

`lower_is_better` declares the unfavourable measurement directions. The
function then changes their signs internally, so all weights below are stated
in a favourable-direction space. A positive weight for grey leaf spot
therefore rewards greater resistance, not greater disease severity.

We use `scale_traits = TRUE`. Consequently, both $\mathbf{w}$ and
$\mathbf{W}$ refer to GEBVs measured in standard-deviation units. Without
scaling, their units would depend on the original measurement scale: $w_j$ has
inverse trait units, $W_{jj}$ has inverse squared units, and $W_{jk}$ has
inverse cross-trait units.

---

## 4. Define one explicit quadratic objective

The following objective rewards improvement in all six traits, introduces
diminishing returns, and adds complementarity between grain yield and disease
resistance. These numbers demonstrate the mechanics. They are not universal
recommendations.

```{r define-objective}
linear_weights <- c(
  GY = 1.00, PHT = 0.15, AD = 0.10,
  ASI = 0.20, EPP = 0.40, GLS = 0.40
)

W <- diag(c(
  GY = -0.08, PHT = -0.04, AD = -0.04,
  ASI = -0.06, EPP = -0.08, GLS = -0.08
))
dimnames(W) <- list(traits, traits)

# The complete cross-product coefficient is 2 * 0.03 = 0.06.
W["GY", "GLS"] <- W["GLS", "GY"] <- 0.03
W
```

The diagonal entries make the objective concave along each trait axis. The
positive `GY` by `GLS` term gives additional value to candidates that are above
the reference mean for both yield and resistance. It also raises the score of
candidates below the mean for both traits, because the product of two negative
deviations is positive. The linear terms and negative diagonal curvature are
expected to oppose that second effect here, but this must be verified from the
candidate-specific contributions. Before using a cross-product term, the
breeding programme should state why the interaction has value beyond the sum
of the separate effects and inspect all four sign combinations.

---

## 5. Fit the index and select candidates

```{r fit-qgsi}
qgsi <- run_qgsi(
  init_data = dgr_gebv["GenoID"],
  gebv_data = dgr_gebv,
  trait_cols = traits,
  linear_weights = linear_weights,
  W = W,
  Gamma = dgr_G,
  lower_is_better = lower_is_better,
  center_traits = TRUE,
  scale_traits = TRUE,
  n_select = 20L
)
qgsi
```

`Gamma = dgr_G` supplies the genomic covariance used by the theoretical
calculations. When `Gamma` is absent, `run_qgsi()` estimates the covariance of
the supplied GEBVs and records that estimand. It does not relabel a covariance
of predictions as the total additive genetic covariance.

Inspect the ranking together with its linear and quadratic parts.

```{r ranking}
head(qgsi$ranked_geno[, c(
  "GenoID", "LinearPart", "QuadraticPart", "QGSI", "Rank", "Selected"
)])
```

The `LinearPart` and `QuadraticPart` are components of the same score. Their
signs do not say that a candidate is biologically good or bad by themselves.
the selection decision follows their sum, `QGSI`.

---

## 6. Interpret the predicted response

```{r expected-gain}
qgsi$expected_gain_per_trait
qgsi$observed_selection_differential
```

These two tables answer different questions.

- `expected_gain_per_trait` is the model-based expected genetic response. It
  uses the genomic covariance and divides by the standard deviation of the
  complete index, including its quadratic variance.
- `observed_selection_differential` is the difference between the selected
  candidates' mean GEBV and the full candidate mean. It describes the selected
  group. It is not automatically the response transmitted to progeny.

The expected-gain calculation is a linear-regression approximation. A
quadratic score is not normally distributed, so the normal-theory selection
differential becomes less accurate as curvature becomes stronger. Report the
assumptions stored with the result rather than presenting the approximation as
an observed outcome.

Under the centred multivariate-normal model,
$\operatorname{Cov}(\widehat{\mathbf{g}},
\widehat{\mathbf{g}}^{\mathsf T}\mathbf{W}
\widehat{\mathbf{g}})=0$ because the relevant third central moments are
zero. Therefore, the numerator of the linear-regression gain is
$\mathbf{\Gamma}\mathbf{w}$. $\mathbf{W}$ enters through the total index
variance. This is a property of that model, not proof that curvature has no
biological effect. When curvature materially changes the selected set, inspect
the observed GEBV differential and use simulation or later-cohort validation
to assess the non-linear decision.

```{r theory}
qgsi$theoretical_parameters[c(
  "linear_index_variance",
  "quadratic_index_variance",
  "total_index_variance",
  "index_standard_deviation",
  "selection_intensity"
)]
```

Accuracy and mean squared prediction error require a true genetic covariance,
which is normally available only in simulation. They are deliberately left
unavailable in empirical applications rather than being estimated by treating
the GEBV covariance as truth.

---

## 7. Identify what changed each candidate's score

Candidate-specific contributions make the non-linearity auditable.

```{r contributions}
head(qgsi$linear_contributions)
head(qgsi$quadratic_contributions[, seq_len(
  min(8L, ncol(qgsi$quadratic_contributions))
)])
```

Use these tables to answer concrete questions: was a candidate selected because
of yield, because it approached an intermediate plant-height optimum, or
because it combined yield and resistance? A large contribution should be
checked against the biological justification for the corresponding weight.

---

## 8. Determine whether the quadratic term matters

Fit the linear special case by setting $\mathbf{W}=0$, then compare rankings
and selected sets. If the decision barely changes, the quadratic complexity is
not buying a materially different decision.

```{r linear-comparator}
linear_only <- run_qgsi(
  init_data = dgr_gebv["GenoID"],
  gebv_data = dgr_gebv,
  trait_cols = traits,
  linear_weights = linear_weights,
  W = matrix(0, length(traits), length(traits),
    dimnames = list(traits, traits)
  ),
  Gamma = dgr_G,
  lower_is_better = lower_is_better,
  center_traits = TRUE,
  scale_traits = TRUE,
  n_select = 20L
)

rank_comparison <- merge(
  as.data.frame(qgsi$ranked_geno[, c("GenoID", "Rank", "Selected")]),
  as.data.frame(linear_only$ranked_geno[, c("GenoID", "Rank", "Selected")]),
  by = "GenoID", suffixes = c("_quadratic", "_linear")
)

c(
  spearman_rank_correlation = stats::cor(
    rank_comparison$Rank_quadratic,
    rank_comparison$Rank_linear,
    method = "spearman"
  ),
  selected_in_both = sum(
    rank_comparison$Selected_quadratic & rank_comparison$Selected_linear
  )
)
```

This comparison is more informative than asking whether the quadratic term is
statistically non-zero. The breeder's question is whether it changes which
candidates are retained, whether that change is stable, and whether the
non-linear value judgement is defensible.

`compare_selection_methods()` accepts both QGSI objects. Use
`comparison_objective()` to fix the utility and covariance matrix. Supply
common validation values when the objective contains curvature. The function
then evaluates the selected sets under the same linear and quadratic utility.
Use the same candidates, transformations, weights, covariance matrix and
selected count.

Report at least four results:

1. expected response for every trait;
2. Spearman rank correlation;
3. selected-set overlap and Jaccard similarity;
4. the linear and quadratic score contributions for candidates whose decision
   changed.

The Jaccard similarity is the number selected by both fits divided by the
number selected by either fit. A value of one means that both fits select the
same candidates. A value of zero means that the selected sets do not overlap.

```{r linear-quadratic-decision-summary}
selected_quadratic <- rank_comparison$Selected_quadratic
selected_linear <- rank_comparison$Selected_linear
selected_overlap <- sum(selected_quadratic & selected_linear)
selected_union <- sum(selected_quadratic | selected_linear)

c(
  selected_overlap = selected_overlap,
  selected_jaccard = selected_overlap / selected_union
)
```

Evaluate both selected sets under one quadratic utility. The fitted weights
refer to standardised GEBVs. Convert them back to weights per original GEBV
unit before defining the common objective.

```{r unified-linear-quadratic-comparison}
qg_scale <- qgsi$transformation$scale[traits]
common_linear_raw <- linear_weights / qg_scale
common_W_raw <- sweep(sweep(W, 1L, qg_scale, "/"), 2L, qg_scale, "/")

qg_objective <- comparison_objective(
  aggregate_weights = common_linear_raw,
  W = common_W_raw,
  G = dgr_G[traits, traits],
  gain_units = "trait"
)

qg_comparison <- compare_selection_methods(
  list(Linear = linear_only, Quadratic = qgsi),
  objective = qg_objective,
  validation_data = dgr_gebv[, c("GenoID", traits)]
)
qg_comparison$summary[, c(
  "Method", "Expected_response_basis",
  "Validation_utility_response",
  "Validation_utility_rank_correlation"
)]
qg_comparison$selected_jaccard
```

The validation utility is exact for the supplied GEBVs. Its evidential strength
depends on their provenance. Independent predictions or simulated true genetic
values provide stronger evidence than the values used to construct the scores.

The quadratic fit earns its extra complexity only when it improves a declared
non-linear utility in independent data or simulation. A change in rank alone
does not establish improvement.

---

## 9. Minimum evidence before operational use

Do not adopt a QGSI because it is more flexible than a linear index. Adopt it
only when the programme can defend the curvature.

1. State the biological or economic reason for every non-zero element of
   $\mathbf{W}$.
2. Record whether GEBVs were centred and scaled, and define the reference
   population used for both operations.
3. Confirm that $\mathbf{W}$ is symmetric and that trait names and directions
   match the GEBV columns.
4. Compare the quadratic ranking and selected set with the linear special case.
5. Inspect candidate-specific contributions for implausible dominance by one
   trait or interaction.
6. Examine the response for every trait, not only the total QGSI score.
7. Test rank and selected-set stability under plausible weight and covariance
   perturbations.
8. Validate the decision in later cohorts before assigning realised breeding
   value to the selected set.

The practical output is not “QGSI is better”. It is a documented statement
that an explicit non-linear objective changed the decision in a useful, stable
and biologically defensible way.

---

## 10. Related documentation

- [Multiple-trait selection](DesiredGainR-index-families.html) places QGSI
  beside the linear and desired-gain families.
- [Iterative optimisation of the desired-gain
  index](DesiredGainR-dgsi.html) explains the iterative search applied to the
  desired-gain selection index. It represents a different breeding objective
  from a quadratic economic index.
- [`run_qgsi()`](../reference/run_qgsi.html) gives the complete argument and
  return-value contract.
- [`compare_dg_and_qgsi()`](../reference/compare_dg_and_qgsi.html) compares
  scores, rankings and selected sets descriptively. Its `decision_summary`
  reports overlap and Jaccard similarity. These quantities do not imply that
  DGSI and QGSI optimise the same objective.

## References

Ceron-Rojas JJ, Montesinos-Lopez OA, Montesinos-Lopez A, et al. (2026).
Nonlinear genomic selection index accelerates multi-trait crop improvement.
*Nature Communications*, 17, 1991.
https://doi.org/10.1038/s41467-026-69890-3
