---
title: "Comparing objectives over several cycles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing objectives over several cycles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(DesiredGainR)
data(dgr_traits, package = "DesiredGainR")
data(dgr_G, package = "DesiredGainR")
data(dgr_hap1, package = "DesiredGainR")
data(dgr_hap2, package = "DesiredGainR")
data(dgr_map, package = "DesiredGainR")
traits <- dgr_traits$trait
lower_is_better <- dgr_traits$trait[dgr_traits$direction == "decrease"]
```

## 1. What this layer is for, and what it is not for

The purpose is narrow and deliberately kept so: **to compare desired-gain
directions over several breeding cycles.** It is not a crossing-plan tool, and
parent, cross and mating allocation are outside the package by design.

The heavy chunks below are marked `eval = FALSE` because a realistic run takes
minutes to hours. Output blocks shown as fixed text illustrate the shape of
each object; your numbers will depend on your founders and parameters.

`AlphaSimR` is required and lives in `Suggests`. Every function fails with an
explicit installation message when it is absent.

---

## 2. Why simulate at all

For any linear index the response is
$\mathbf{R} = i\,\mathbf{G}\mathbf{b}/\sqrt{\mathbf{b}^\mathsf{T}\mathbf{P}\mathbf{b}}$,
so attainable responses lie on the ellipsoid

$$\mathbf{R}^\mathsf{T}\mathbf{G}^{-1}\mathbf{P}\mathbf{G}^{-1}\mathbf{R} = i^{2}.$$

`gain_feasibility()` solves that exactly, without simulation. **A single-cycle
calculation therefore cannot distinguish desired-gain directions beyond what
the ellipsoid already states**, and any simulator that merely re-applies the
one-cycle formula $T$ times returns $T$ times the one-cycle answer.

Multi-cycle simulation earns its cost only by representing what makes cycle
five differ from cycle one:

1. **Bulmer effect.** Truncation selection removes variance anisotropically,
   by $k = i(i-x)$ along the index direction. A direction that hammers one axis
   exhausts the variance that axis depends on.
2. **Finite $N_e$.** Drift, and the fact that the most aggressive index usually
   erodes effective population size fastest.
3. **Antagonistic correlations.** A secondary trait can cross an unacceptable
   threshold in cycle three, which no single-cycle response reveals.
4. **Compounding estimation error**, when the index is rebuilt each cycle from
   that cycle's own data.

If none of those matter for your question, use `gain_feasibility()` and stop.

---

## 3. Founders come from your germplasm, not from a simulated genome

This is the design decision that determines whether the results mean anything.

A simulator that generates founder genomes from a coalescent model produces
linkage disequilibrium, allele frequencies and population structure belonging
to an assumed demography. Optimising a desired-gain direction against that is
optimising for a fictional population. DesiredGainR therefore requires the
breeder's own phased marker data.

### 3.1 The input contract

```{r founders}
founders <- founder_haplotypes(dgr_hap1, dgr_hap2, dgr_map)
founders
```

`hap1` and `hap2` are marker-by-individual matrices coded 0, 1 or `NA`, giving
the allele carried on the first and second homologue. Row names are variant
identifiers, column names individual identifiers, and both matrices must carry
identical dimnames in identical order.

```{r founders-shape}
dim(dgr_hap1)
dgr_hap1[1:3, 1:4]
head(dgr_map, 3)
```

### 3.2 Why dosage is refused

A haplotype is one chromosome copy, so at a biallelic locus it carries a single
allele rather than a count. Genotype dosage coded 0, 1, 2 is the *sum* of the
homologues:

$$\text{dosage} = \text{hap}_1 + \text{hap}_2.$$

It records how many alternative alleles an individual carries but not which
copy carries them, so it cannot say which alleles lie in *cis* across loci.
That co-occurrence is the linkage disequilibrium the simulation exists to
represent, and assigning heterozygous calls to homologues at random would
destroy it.

```{r dosage-refused, error=TRUE}
dosage <- dgr_hap1 + dgr_hap2
founder_haplotypes(dosage, dgr_hap2, dgr_map)
```

### 3.3 The exception: inbred material

In a diploid inbred line a dosage of 0 or 2 is unambiguous, so phase can be
derived without external phasing. Only heterozygous and missing calls need a
decision.

Measure the rates before converting, because a low overall level can conceal a
few badly affected individuals or markers:

```{r diagnostics}
diagnostics <- dosage_diagnostics(dosage)
diagnostics
```

```{r convert, eval=FALSE}
converted <- haplotypes_from_inbred_dosage(
  dosage,
  heterozygous_policy = "drop_variant",
  missing_policy = "drop_variant"
)
founders <- founder_haplotypes(converted$hap1, converted$hap2, dgr_map)
```

**No threshold is imposed on residual heterozygosity**, because no universally
appropriate level exists: it depends on the generation, mating history, crop,
population type, genotyping error rate and quality-control procedure. The rates
are measured and reported instead, and heterozygous calls are never resolved
silently. The four policies are `"error"`, `"drop_variant"`,
`"drop_individual"` and `"mask"`. None assigns a heterozygote to a homologue.

Use this route only for highly inbred diploid material. For outcrossing or
clonal germplasm, phase externally.

### 3.4 Diploid only

The contract is diploid and biallelic, and a polyploid dataset raises an
explanatory error rather than proceeding. Adding homologue matrices would not
make the simulation polyploid: autopolyploid and allopolyploid meiosis require
an explicit crop-specific pairing and recombination model, including
multivalent formation and double reduction. Applying a diploid meiotic model to
polyploid homologues would produce confidently wrong results.

---

## 4. Calibrating the trait architecture

Genome structure comes from your markers; trait variances and correlations are
calibrated to the covariance matrix you already estimated.

```{r setup-sim, eval=FALSE}
setup <- founder_population(
  founders,
  G = dgr_G,
  h2 = stats::setNames(dgr_traits$heritability, traits),
  n_qtl_per_chromosome = 100L,
  seed = 42L
)
setup
```

```
#> <desiredgainr_sim_setup>
#>   Founders: 200 individuals, 3 chromosomes, 100 QTL per chromosome
#>   Traits: GY, PHT, AD, ASI, EPP, GLS
#>   Dominance simulated: no
```

Trait variances are taken from $\operatorname{diag}(\mathbf{G})$ and
correlations from the corresponding correlation matrix, so the simulation
reproduces the observed germplasm structure *and* the observed trait
covariances.

**For a clonal programme**, supply `dominance_degree`, and `G` should then be
the **genotypic** rather than the additive covariance, because the unit of
selection is the clone and dominance is inherited intact.

```{r setup-clonal, eval=FALSE}
setup_clonal <- founder_population(
  founders,
  G = dgr_G, # genotypic covariance for a clonal crop
  h2 = stats::setNames(dgr_traits$heritability, traits),
  dominance_degree = stats::setNames(rep(0.3, length(traits)), traits),
  seed = 42L
)
```

---

## 5. Simulating one objective

```{r simulate, eval=FALSE}
simulation <- simulate_selection_cycles(
  setup,
  desired_gains = c(
    GY = 1.0, PHT = 0.4, AD = 0.6,
    ASI = 0.5, EPP = 0.4, GLS = 0.6
  ),
  n_cycles = 5L,
  mating_system = "outcross",
  n_parents = 20L,
  n_crosses = 50L,
  n_progeny_per_cross = 10L,
  lower_is_better = lower_is_better,
  reestimate_index = TRUE,
  seed = 11L
)
simulation
```

```
#> <desiredgainr_simulation>
#>   outcross system, 5 cycles, 20 parents recycled
#>   Index re-estimated each cycle: yes
#>   Cumulative genetic gain:
#>       GY     PHT      AD     ASI     EPP     GLS
#>    1.842  -8.031  -1.226  -0.244  -0.019  -0.487
#>   Final mean relationship among parents: 0.0913
```

### 5.1 The three mating systems

| System | Crops | Cycle structure |
|---|---|---|
| `"self"` | Wheat, rice, common bean, cowpea | Selected parents intercrossed, families advanced by selfing or doubled haploidy before evaluation. Recombination releases variance slowly. |
| `"outcross"` | Maize, sorghum, pearl millet | Random mating each cycle, so half the selection-induced disequilibrium decays per generation. |
| `"clonal"` | Cassava, sweetpotato, banana, potato | Recombination happens once per cycle, when selected parents are crossed to raise a seedling generation; every later stage is a copy, so dominance is transmitted intact and selection acts on **total genotypic value**. Requires a setup built with `dominance_degree`, and `G` is then the genotypic covariance. |

### 5.2 What a cycle means

Check this definition before reading any number below.

**Cycle 0** is the founder population, before any selection. It exists so that
the cycle 1 row has something to be a response *to*.

**Cycle \(t\)** records the population produced by selecting parents from the
cycle \(t-1\) candidates and crossing them. Every row from cycle 1 onward is a
transmitted selection response, so `n_cycles = 1` gives exactly one.

Versions before 0.5.0 measured the *candidates* of each cycle rather than their
selected descendants. Because random mating does not shift a population mean,
the cycle 1 gain was zero in expectation for every desired-gain direction, and
a one-cycle run could not distinguish directions at all. Re-run any stored
result.

```{r cycle-table, eval=FALSE}
simulation$cycle_table
```

```
#>    cycle  trait genetic_mean cumulative_gain genetic_variance
#> 1:     0     GY        0.000           0.000            0.612
#> 2:     0    PHT        0.000           0.000          148.203
#> 3:     1     GY        0.402           0.402            0.548
#> 4:     1    PHT       -1.907          -1.907          139.412
#> ...
#>    mean_relationship mean_inbreeding parent_inbreeding qtl_segregating
#>                0.021           0.014             0.038           0.981
#>    genic_variance_ratio effective_size selection_intensity n_evaluated
#>                   0.994           23.9               1.755         500
```

Gain and the loss of diversity are recorded together deliberately. A direction
that maximises cumulative gain while collapsing effective population size is
not obviously preferable, and the trade-off should be visible rather than
inferred.

`mean_relationship` and `mean_inbreeding` are measured on a **genome-wide
marker panel**, not on the quantitative trait loci, and against allele
frequencies fixed at the founders. Both choices matter. Measuring on QTL would
make the metric rise whenever selection succeeded — changing QTL frequencies is
what genetic gain is — so a direction would be penalised for working.
Recomputing the frequencies each cycle would pin the mean off-diagonal at
\(-\overline{G}_{ii}/(n-1)\), a function of population size alone. QTL
frequency change is reported separately as `qtl_segregating` and
`genic_variance_ratio`.

Supply `n_markers_per_chromosome` to `founder_population()` to create the
panel; without it, all segregating sites are used instead.

### 5.3 The re-estimation switch

`reestimate_index = TRUE`, the default, rebuilds the covariance of the observed
selection criterion from each cycle. It does **not** estimate genetic
covariance from hidden simulated breeding values. Phenotypic selection retains
the breeder-supplied `G_target`; genomic selection uses the calibrated-GEBV
identity described below. Setting it `FALSE` reuses the cycle-one coefficients
and isolates the effect of the desired-gain direction alone.

**Run both.** A large divergence means the recommendation is sensitive to
estimation error rather than to the objective, and the breeder should be told
so.

### 5.4 Phenotypic or genomic selection

The default is explicit phenotypic selection. To represent an additive genomic
selection programme, create a marker panel and request cross-fitted RR-BLUP:

```{r rrblup, eval=FALSE}
setup <- founder_population(
  founders,
  G = dgr_G, h2 = 0.4,
  n_qtl_per_chromosome = 20,
  n_markers_per_chromosome = 100,
  seed = 42
)

genomic <- simulate_selection_cycles(
  setup,
  desired_gains = c(
    GY = 1, PHT = 0.4, AD = 0.6,
    ASI = 0.5, EPP = 0.4, GLS = 0.6
  ),
  mating_system = "outcross",
  prediction = list(
    method = "rrblup",
    folds = 5,
    update_training = TRUE,
    max_training = 2000
  )
)
```

Every candidate is predicted by a model that excluded its fold, preventing
training-set leakage. Earlier cycles may be retained as a rolling training
population. The cycle table reports the correlation between held-out GEBV and
true simulated breeding value as `prediction_accuracy`; truth is used only for
this diagnostic, never for selection. `prediction_calibration_slope` additionally
reports `Cov(A, Ahat) / Var(Ahat)`; values far from one warn that the calibrated-
GEBV assumption used by the desired-gain index is poor. RR-BLUP currently
supports additive, non-clonal programmes. A clonal dominance programme must use
the phenotypic path until a validated total-genetic-value predictor is
implemented.

### 5.5 The selection rule is deliberately simple

Truncation on the index, and nothing more. Optimal contribution selection,
family quotas and coancestry constraints are the province of a mating-design
tool. Duplicating them here would mean maintaining two simulators with
different genetic engines.

---

## 6. Searching for the best objective

### 6.1 The search space

Only the *direction* of $\mathbf{d}$ affects the index, because the attainable
magnitude is fixed by selection intensity. The domain is therefore the unit
sphere,

$$\|\mathbf{d}\|_2 = 1,$$

with $p-1$ free parameters for $p$ traits. For the three to eight traits
breeders use, that is small enough to cover densely. By default the search is
restricted to the non-negative orthant, meaning improvement is sought in every
trait; set `non_negative = FALSE` to admit directions that deliberately concede
ground somewhere.

### 6.2 The simulation is the objective function

```{r optimise, eval=FALSE}
optimisation <- optimize_desired_gains(
  setup,
  n_cycles = 5L,
  mode = "pareto",
  budget = 60L,
  n_initial = 30L,
  n_replicates = 3L,
  include_diversity = TRUE,
  mating_system = "outcross",
  n_parents = 20L, n_crosses = 50L, n_progeny_per_cross = 10L,
  lower_is_better = lower_is_better,
  checkpoint = "search_state.rds",
  seed = 42L
)
optimisation
```

```
#> <desiredgainr_optimisation>
#>   Mode: pareto   Cycles: 5   Evaluations: 60 (3 replicates each)
#>   Objectives: GY, PHT, AD, ASI, EPP, GLS, diversity
#>   Non-dominated directions: 9 of 60
#>   No single direction is recommended; choose a point on the frontier.
```

A Gaussian process is fitted to the accumulated results, but **it decides only
where the next simulation should be spent.** It never filters, screens, or
substitutes for an evaluation, and because the acquisition function retains an
exploration term across the whole domain, no region can be permanently excluded
by the surrogate.

An earlier design used a deterministic infinitesimal-model recursion as the
surrogate's prior mean. That was rejected: it would embed a second genetic
model whose assumptions could fail in precisely the situations where simulation
is most needed.

### 6.3 The five ranking modes

| Mode | Objective | Required arguments |
|---|---|---|
| `"pareto"` | Non-dominated set of multi-cycle outcomes | none |
| `"interval"` | Maximise conservative joint probability of reaching all interval minima | `desired_gain_intervals` |
| `"economic"` | Maximise $\mathbf{w}^\mathsf{T}\mathbf{R}_{\text{cum}}$ | `economic_weights` |
| `"target"` | Minimise distance to stated absolute targets | `target_gains` |
| `"constrained"` | Maximise a focal trait subject to floors | `focal_trait`, `gain_floors` |

`"pareto"` is the default because it makes no scalar preference assumption.
`"interval"` is the breeder-facing recommendation mode when the team can state
minimum and aspirational gains but cannot defend economic weights. It reports
the joint probability of reaching every minimum and recommends conservatively
by the lower credible bound. See the dedicated desired-gain interval vignette.

The Pareto search uses randomised augmented-Chebyshev scalarisation, which
converges on the frontier while requiring only single-objective improvement. A
fresh weight vector is drawn each iteration so the search sweeps the frontier
rather than one corner of it.

### 6.4 Noise, and why the frontier is smoothed

Simulation output is stochastic. Two properties matter for interpretation.

**Common random numbers.** Every direction is evaluated with the same sequence
of seeds, so comparisons between directions share their stochasticity. This
removes far more comparison variance than increasing the replicate count.

**The frontier comes from posterior means, not raw draws.** A frontier built
from raw replicate averages is populated by fortunate runs. The reported
non-dominated set is computed from Gaussian-process posterior means.

```{r frontier, eval=FALSE}
optimisation$pareto_set[, c("d_GY", "d_ASI", "posterior_GY", "posterior_ASI")]
```

### 6.5 Recommendations retain their uncertainty

For the scalar modes the result reports a **stability region**: every direction
whose posterior outcome lies within `stability_tolerance` of the best.

```{r stability, eval=FALSE}
optimisation$stability_proportion
optimisation$recommended_direction
```

The optimum is conditional on the supplied genetic covariance, the founder
germplasm and the programme parameters. Reporting one direction to three
decimal places would imply a precision the analysis does not have.

### 6.6 Budget and resumption

A realistic budget takes hours. Supplying `checkpoint` writes the accumulated
evaluations after every simulation and reloads them automatically on a later
call, so an interruption does not discard the work.

```{r resume, eval=FALSE}
# Re-running the same call continues from where it stopped.
optimisation <- optimize_desired_gains(
  setup,
  n_cycles = 5L, mode = "pareto", budget = 120L,
  checkpoint = "search_state.rds", seed = 42L, ...
)
```

---

## 7. What the results are conditional on

State these alongside any recommendation.

| Assumption | Consequence if wrong |
|---|---|
| Founder panel represents the breeding population | The optimised direction suits a different population |
| $\mathbf{G}$ is the covariance of the target population | Trait architecture is miscalibrated throughout |
| Quantitative trait loci are many and of small effect | The infinitesimal approximation degrades |
| The programme parameters match practice | Intensity and drift are misrepresented |
| Truncation is the operative selection rule | Coancestry-controlled programmes will differ |

The simulation is a decision-support tool, not a prediction. Its value lies in
ranking directions under stated assumptions, and it should be reported with
those assumptions attached.

---

## 8. Function reference for this layer

| Function | Role |
|---|---|
| `dosage_diagnostics()` | Measure heterozygosity and missingness before converting |
| `haplotypes_from_inbred_dosage()` | Derive phase from inbred dosage, under an explicit policy |
| `founder_haplotypes()` | Validate and package phased founders |
| `founder_population()` | Build the AlphaSimR population, calibrated to $\mathbf{G}$ |
| `simulate_selection_cycles()` | Run one desired-gain direction forward |
| `optimize_desired_gains()` | Search directions, surrogate-assisted |

---

## 9. References

- Bulmer MG (1971). The effect of selection on genetic variability.
  *The American Naturalist* **105**:201-211.
- Gaynor RC, Gorjanc G, Hickey JM (2021). AlphaSimR: an R package for breeding
  program simulations. *G3 Genes|Genomes|Genetics* **11**:jkaa017.
  <https://doi.org/10.1093/g3journal/jkaa017>
- Knowles J (2006). ParEGO: a hybrid algorithm with on-line landscape
  approximation for expensive multiobjective optimization problems.
  *IEEE Transactions on Evolutionary Computation* **10**:50-66.
- Vieira RA, Nogueira APO, Fritsche-Neto R (2025). Optimizing the selection of
  quantitative traits in plant breeding using simulation.
  *Frontiers in Plant Science* **16**:1495662.
  <https://doi.org/10.3389/fpls.2025.1495662>
