---
title: "Multiple-trait selection: methods, choice, and comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple-trait selection: methods, choice, and comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(DesiredGainR)
data(dgr_traits, package = "DesiredGainR")
data(dgr_G, package = "DesiredGainR")
data(dgr_P, package = "DesiredGainR")
data(dgr_candidates, package = "DesiredGainR")
data(dgr_gebv, package = "DesiredGainR")
traits <- dgr_traits$trait
lower_is_better <- dgr_traits$trait[dgr_traits$direction == "decrease"]
```

## 1. Learning objectives

Plant breeding programmes usually improve several traits at the same time.
Yield, quality, adaptation, maturity, and disease resistance may all influence
the release decision. Progress in one trait can also reduce progress in
another. Therefore, the method must represent both the breeding objective and
the genetic relationships among traits.

This vignette has four objectives. They are to: (i) define multiple-trait
selection, (ii) explain its main methods, (iii) show when each method is
appropriate, and (iv) provide a comprehensive comparison analysis.

At the end, the breeder should be able to justify the selected method. The
justification should refer to the objective, the available information, the
expected responses, and the stability of the final ranking.

## 2. What is multiple-trait selection?

Multiple-trait selection is the choice of candidates using two or more traits.
The traits can enter the decision at the same stage or at different stages.
The breeder must first decide how trade-offs are allowed.

Four broad strategies are recognised by Beavis, Lamkey, Mahama, and Suza
(2023).

| Strategy | Definition | Suitable situation | Main limitation |
|---|---|---|---|
| Multistage selection | Different traits are evaluated at different stages. | Expensive traits are measured after inexpensive screening. | Early rejection can remove candidates that would perform well later. |
| Tandem selection | One trait is improved across one or more cycles before another trait becomes the focus. | One trait has temporary programme priority. | Other traits can deteriorate before the focus changes. |
| Independent culling | Every candidate must satisfy a limit for every required trait. | Release standards or biological limits are firm. | Excellence elsewhere gives no compensation for a failed limit. |
| Index selection | Traits are combined into one score of overall merit. | Trade-offs are allowed and simultaneous improvement is required. | The score is useful only when its objective and inputs are defensible. |

These strategies answer different questions. Hence, a comparison should never
treat them as interchangeable algorithms. The breeder's decision rule comes
first. The statistical method follows.

### 2.1 Choose the strategy before choosing the index

No strategy is universally best. The suitable strategy follows from four
programme decisions.

1. **Are any limits non-negotiable?** Use independent culling for those limits.
   An index can rank the candidates that pass.
2. **Are traits measured at different stages or costs?** Use a multistage
   design. Place inexpensive or highly heritable screens early. Reserve costly
   traits for the reduced set.
3. **Must one trait dominate several breeding cycles?** Use a true tandem
   schedule. Monitor correlated deterioration in the other traits before the
   focal trait changes.
4. **Can progress in one trait compensate for less progress in another?** Use
   index selection. Then choose the index family from the biological objective.

These choices can be combined. For example, disease susceptibility can define
a firm culling limit. A desired-gain index can then rank the eligible lines for
yield, quality, and maturity. This hybrid rule represents the programme more
accurately than forcing every trait into one score.

### 2.2 What DesiredGainR implements

The four strategies have different software coverage. The distinction below
is important.

| Strategy | Current package support | What can be compared directly |
|---|---|---|
| Multistage selection | No dedicated stage-allocation optimiser | Analyse the candidate set retained after each stage. Record attrition, cost, and final response separately. |
| Classical tandem selection across cycles | No dedicated alternating-cycle simulator | Define separate cycle scenarios outside the one-call comparison. Monitor response in every trait. |
| Within-cohort sequential screening | `selection_index(method = "tandem")` | Rankings, selected candidates, and observed differentials within one cohort. |
| Independent culling | `selection_index(method = "independent_culling")` | Passing rate, failed limits, selected candidates, and observed differentials. |
| Index selection | `selection_index()`, `restricted_index()`, `generalized_index()`, `run_dgsi()`, and `run_qgsi()` | The available criteria depend on the fitted index family. They are defined in Section 6. |

The option named `method = "tandem"` is retained for compatibility. It applies
successive trait screens to one candidate cohort. Classical tandem selection
changes the focal trait across cycles or generations. These procedures should
not be interpreted as the same strategy.

### 2.3 Compare strategies on programme outcomes

Strategy comparison requires more than a common candidate score. Use the same
initial population, selection budget, time horizon, and release constraints.
Then compare the following outcomes.

| Outcome | Why it matters |
|---|---|
| Candidates retained after each decision | Shows where useful material is lost. |
| Cost per selected candidate | Gives multistage selection its operational value. |
| Expected transmitted response by trait | Measures the genetic consequence of selection. |
| Frequency of failed mandatory limits | Protects non-negotiable requirements. |
| Cycle time | Distinguishes rapid simultaneous selection from a tandem schedule. |
| Rank and selected-set agreement | Shows whether strategies change the immediate decision. |
| Multi-cycle response and genetic variance | Shows whether short-term gain persists. |
| Validation in a later cohort | Tests whether the rule transfers beyond the fitted candidates. |

Therefore, the preferred strategy is the simplest rule that respects every
firm constraint and gives the strongest validated response within the
programme's cost and time limits. A larger index statistic alone cannot choose
among the four strategies.

## 3. The quantities used by an index

Let \(\mathbf{x}\) contain the information recorded for a candidate. Let
\(\mathbf{g}\) contain the additive genetic values of the objective traits.
The two vectors can contain the same traits. They can also differ. For example,
records from relatives, testcrosses, environments, or correlated traits can
help predict a smaller set of objective traits.

Three covariance matrices describe this problem.

| Symbol | Definition | Breeding meaning |
|---|---|---|
| \(\mathbf{P}=\operatorname{Var}(\mathbf{x})\) | Covariance among information variables | Describes the variation and redundancy in the evidence used for selection. |
| \(\mathbf{C}=\operatorname{Cov}(\mathbf{x},\mathbf{g})\) | Covariance between information and objective traits | Describes how strongly each information source predicts genetic value. |
| \(\mathbf{G}=\operatorname{Var}(\mathbf{g})\) | Genetic covariance among objective traits | Describes available genetic variation and correlated response. |

The aggregate genetic merit is

$$
H=\mathbf{a}^{\mathsf T}\mathbf{g},
$$

where \(\mathbf{a}\) contains economic weights. A linear selection index is

$$
I=\mathbf{b}^{\mathsf T}\mathbf{x},
$$

where \(\mathbf{b}\) contains index coefficients. Under truncation selection,
the expected response vector is

$$
\Delta
=i\frac{\mathbf{C}^{\mathsf T}\mathbf{b}}
{\sqrt{\mathbf{b}^{\mathsf T}\mathbf{P}\mathbf{b}}}.
$$

The term \(i\) is the selection intensity. It increases when a smaller
proportion of candidates is selected. When information and objective traits
are identical, \(\mathbf{C}=\mathbf{G}\). This gives the familiar expression
used throughout classical selection-index theory.

Use `selection_information()` and `generalized_index()` when the information
and objective traits differ. Read [Using predictions, prediction errors, and
restricted responses](DesiredGainR-information.html) for the complete
formulation.

## 4. The methods available in DesiredGainR

The table below links each statistical method to its exact package call. The
value in the `method` column is supplied to the stated function. This makes the
mathematical name and the R interface explicit.

| Index family | Function and method value | Main input that defines the objective |
|---|---|---|
| Smith-Hazel | `selection_index(method = "smith_hazel")` | Economic weights |
| Base index | `selection_index(method = "base")` | Economic weights |
| Pesek-Baker | `selection_index(method = "pesek_baker")` | Desired gains |
| Yamada | `selection_index(method = "yamada")` | Desired gains |
| Mulamba-Mock rank sum | `selection_index(method = "mulamba_mock")` | Rank weights, or equal weights |
| Elston multiplicative index | `selection_index(method = "elston")` | Trait-specific culling limits |
| Independent culling | `selection_index(method = "independent_culling")` | Trait-specific culling limits |
| Within-cohort sequential screen | `selection_index(method = "tandem")` | Ordered trait screens in one cohort |
| Kempthorne-Nordskog | `restricted_index(method = "kempthorne_nordskog")` | Economic weights and zero-response traits |
| Restricted Smith-Hazel | `restricted_index(method = "restricted_smith_hazel")` | Economic weights and zero-response traits |
| Tallis | `restricted_index(method = "tallis")` | Economic weights and proportional target gains |
| Mallard | `restricted_index(method = "mallard")` | Economic weights and target gain amounts |
| Harville | `restricted_index(method = "harville")` | Economic weights, target gains, and a finite penalty |
| General economic index | `generalized_index(method = "economic")` | Economic weights and distinct information sources |
| General desired-gain index | `generalized_index(method = "desired_gain")` | Desired gains and distinct information sources |
| Desired-gain index with iterative search | `run_dgsi()` | Desired gains and candidate values |
| QGSI | `run_qgsi()` | Linear, squared, and cross-product economic values |

DGSI means desired-gain selection index. QGSI means quadratic genomic
selection index. The general index applies when the information variables and
the objective traits differ. For example, testcross records may predict the
additive genetic values of parental lines.

### 4.1 Economic-value methods

The Smith-Hazel index maximises expected response in aggregate genetic merit.
Its coefficients are

$$
\mathbf{b}=\mathbf{P}^{-1}\mathbf{G}\mathbf{a}.
$$

Choose this method when the relative economic values are credible. It is also
the correct reference when every method is compared against one fixed
definition of merit.

The base index uses \(\mathbf{b}=\mathbf{a}\). It applies the economic weights
directly. It provides a useful benchmark because it avoids covariance
inversion. A small advantage for Smith-Hazel over the base index suggests that
the estimated covariance structure contributes little to the decision.

### 4.2 Desired-gain methods

The Pesek-Baker and Yamada formulations use a desired response direction
\(\mathbf{d}\). Their coefficient expressions are

$$
\mathbf{b}=\mathbf{G}^{-1}\mathbf{d}
$$

and

$$
\mathbf{b}=\mathbf{P}^{-1}\mathbf{G}
(\mathbf{G}\mathbf{P}^{-1}\mathbf{G})^{-1}\mathbf{d}.
$$

They give the same coefficient direction when \(\mathbf{G}\) is square and
invertible. Therefore, they are two computational routes to the same linear
index. A material difference between their answers indicates numerical
instability or a rank-deficient matrix.

Choose this family when the breeder can state relative desired gains more
confidently than economic weights. The vector fixes response proportions. The
selection intensity fixes the attainable magnitude. Multiplying every desired
gain by the same positive constant leaves the ranking unchanged.

`run_dgsi()` uses the established desired-gain selection index (DGSI) formula.
It adds the iterative search applied by Joukhadar et al. (2024). The search
changes the input desired-gain vector. It seeks a candidate ranking whose
selected differential is close to the breeder's target. Joukhadar et al. did
not introduce a separate index family. Read [Optimising desired
gains](DesiredGainR-dgsi.html) before using the iterative search.

The closed-form Pesek-Baker and Yamada methods define a response direction
from covariance matrices. DGSI adds candidate-level iteration. It searches for
a ranking whose realised selected differential approaches the requested
direction. Fit both when candidate-level attainment matters. Agreement gives a
useful implementation check. A difference directs attention to finite
candidate supply, selection intensity, and the shape of the candidate cloud.

### 4.3 Rank and threshold methods

The Mulamba-Mock rank-sum method ranks candidates within traits and combines
the ranks. It requires neither economic weights nor covariance matrices when
equal rank weights are used. It is valuable as a simple comparator. Its score
does not provide a closed-form predicted genetic response in DesiredGainR.

Independent culling applies firm acceptance limits. Supply each limit in the
original trait units. Use a minimum for traits where larger values are
favourable. Use a maximum for traits listed in `lower_is_better`.

The Elston multiplicative index also applies firm limits. It then ranks the
eligible candidates using their joint margins above those limits. It is useful
when balanced superiority among eligible candidates matters.

`selection_index(method = "tandem")` applies ordered trait screens within one
candidate cohort. It is useful as an operational comparator. It does not
represent classical tandem selection across breeding cycles.

### 4.4 Restricted methods

A restricted index maximises merit while controlling response in specified
traits. DesiredGainR provides five forms.

| Method | Breeding question |
|---|---|
| Kempthorne-Nordskog | Which index holds specified traits at zero expected change? |
| Restricted Smith-Hazel | Which projection form gives the same zero-response restriction? |
| Tallis | Which index preserves stated response proportions? |
| Mallard | Which index targets stated response amounts? |
| Harville | Which index balances merit against a finite restriction penalty? |

Restrictions usually reduce response in unconstrained net merit. That loss is
the cost of respecting the biological or commercial condition. Report both
the restricted response and the loss relative to the unrestricted index.

Satoh (2024) gives a geometric interpretation of proportional restriction.
The scalar \(\beta\) measures progress along the desired direction. The
Mahalanobis residual measures departure from that direction after accounting
for genetic covariance. DesiredGainR reports both quantities through
`evaluate_restricted_response()`.

All five restricted forms use the same interface. The restriction determines
which form has a clear interpretation. Use Kempthorne-Nordskog or restricted
Smith-Hazel for zero expected change. Use Tallis for a response direction. Use
Mallard for stated response amounts. Use Harville when the programme accepts a
measured departure from the target in exchange for greater aggregate merit.

### 4.5 Non-linear economic value

The quadratic genomic selection index (QGSI) represents linear, squared, and
cross-product economic values. Choose it when biological or economic merit is
genuinely curved. A desired gain is a response target. It is a different
quantity from a quadratic weight. Read [Quadratic genomic
selection](DesiredGainR-qgsi.html) for the full demonstration.

`run_qgsi_desired_gain()` remains as a compatibility entry point. It raises an
informative error because desired gains and quadratic economic values define
different objectives. Use `run_dgsi()` for desired gains. Use `run_qgsi()` for
curved economic merit.

## 5. How to choose a strategy and an index family

Begin with the programme structure. Then choose the simplest index family that
represents the allowed trade-offs.

| Breeder's statement | Starting method | Required evidence |
|---|---|---|
| "I can defend relative economic values." | Smith-Hazel | \(\mathbf{G}\), \(\mathbf{P}\), and economic weights |
| "I can state the response direction." | Pesek-Baker or Yamada | \(\mathbf{G}\), and \(\mathbf{P}\) for Yamada |
| "I can state acceptable gain intervals." | `suggest_desired_gains()` | \(\mathbf{G}\), \(\mathbf{P}\), intervals, and candidates |
| "Every candidate must satisfy these limits." | Independent culling | Candidate values and limits |
| "Eligible candidates should have balanced margins." | Elston | Candidate values and limits |
| "Some traits should remain unchanged." | Restricted index | \(\mathbf{G}\), \(\mathbf{P}\), merit weights, and restrictions |
| "Different traits become available at defined stages." | Multistage selection | Measurement cost, reliability, stage capacity, and attrition |
| "One trait must dominate successive cycles." | Classical tandem schedule | Cycle schedule and correlated response in every trait |
| "One cohort must pass ordered trait screens." | `selection_index(method = "tandem")` | Screen order and candidate values |
| "I need a simple comparator with few assumptions." | Base or Mulamba-Mock | Weights for the base index, or candidate ranks |
| "Merit contains curvature or trait interactions." | QGSI | Genomic estimated breeding values and quadratic economic values |

Use the following decision sequence.

1. Define the objective traits and their favourable directions.
2. Identify firm limits that allow no compensation.
3. Record when each trait becomes available and what it costs to measure.
4. Decide whether selection is simultaneous, multistage, or spread across
   cycles.
5. For simultaneous selection, define an economic, desired-gain, restricted,
   rank-based, threshold, or quadratic objective.
6. Confirm the covariance matrices and units required by that objective.
7. Fit one simple benchmark and one objective-matched method.
8. Compare biological response before comparing scalar criteria.
9. Examine ranking stability and uncertainty.
10. Use validation and simulation when the decision extends beyond one cohort.

## 6. What can be compared and how

### 6.1 Exact scope of the comparison functions

`compare_selection_methods()` accepts fitted objects from `selection_index()`,
`restricted_index()`, `generalized_index()`, `run_dgsi()`, and `run_qgsi()`.
Internal adapters harmonise candidates, favourable trait directions and
response units. They compare responses, rankings and selected sets. They never
compare coefficients across families.

Use `comparison_objective()` for a cross-family analysis. It fixes one desired
response target, one aggregate-weight vector, one optional quadratic utility,
and one genetic covariance matrix. Every fitted method is then evaluated
against that same objective.

`compare_dg_and_qgsi()` accepts one `run_dgsi()` result and one `run_qgsi()`
result. It compares candidate scores, ranks, and selected sets. Its
`decision_summary` reports selected counts, overlap and Jaccard similarity.
The comparison is descriptive. A desired response objective and a quadratic economic
objective are different biological statements. Rank agreement cannot decide
which statement is appropriate.

A QGSI with non-zero curvature retains its documented linear-regression
approximation. The special case with $\mathbf{W}=0$ is a linear genomic index
and has the exact linear response. A quadratic common utility is evaluated
exactly from common validation values. This distinction appears in
`Expected_response_basis` and must remain visible when methods are ranked.

### 6.2 Conditions for a fair comparison

A comprehensive comparison requires common conditions. Otherwise, a numerical
difference can reflect the setup rather than the method.

Use the same:

1. candidates and trait records;
2. trait directions and units;
3. one common genetic covariance for the objective traits;
4. selected proportion or selection intensity;
5. aggregate-merit definition when comparing \(R_{HI}\) or \(\Delta H\);
6. training, validation, and test partitions;
7. simulation scenarios, replicate seeds, and computational budget.

The comparison also needs several outcomes. No single statistic is sufficient.

| Outcome | Question answered |
|---|---|
| Expected response for every trait | What genetic change is predicted? |
| Target attainment | How much of each requested gain is reached? |
| Feasibility | Can the complete target be reached at the planned intensity? |
| \(R_{HI}\) | How closely does the index rank one fixed aggregate merit? |
| \(\Delta H\) | How much change in that fixed merit is expected? |
| Relative efficiency (RE) | How much response in the declared main trait remains relative to direct selection? |
| Index heritability | How repeatable is the composite score under the fitted covariance model? |
| Mahalanobis alignment | Does the response follow the requested multivariate direction? |
| Selected-set overlap | Do two methods choose the same candidates? |
| Rank correlation | Do two methods order all candidates similarly? |
| Selection probability | Does candidate choice remain stable after prediction uncertainty is propagated? |
| Information-deletion efficiency | Which records contribute useful selection information? |
| Multi-cycle response and diversity | Does the apparent advantage persist across breeding cycles? |

The coefficient of variation of the index, \(CV_I\), is retained for
reproduction of published analyses. It depends on the score origin and becomes
undefined after centring. Therefore, it should never decide the preferred
method.

### 6.3 Evaluate each family with the criteria it can support

| Family or strategy | Primary criteria | Additional evidence |
|---|---|---|
| Smith-Hazel and base | Expected response by trait, $R_{HI}$, $\Delta H$, relative efficiency, and index accuracy | Weight sensitivity, covariance uncertainty, and validation |
| Closed-form desired-gain index | Expected response by trait, feasibility, target attainment, Satoh's $\beta$, and Mahalanobis alignment | Covariance uncertainty and selected-set stability |
| Desired-gain index with iterative search | The same desired-gain criteria, plus selected differential, holdout objective, replicate stability, and comparison with the closed-form starting index | Independent-cohort validation |
| Restricted index | Expected response, restriction residual, loss in common aggregate merit, and Satoh projection where relevant | Sensitivity to the restriction and penalty |
| Mulamba-Mock rank sum | Rank stability, selected-set overlap, and observed differential | Later-cohort validation or simulation |
| Independent culling | Passing rate, failed limits, selected count, and observed differential | Sensitivity to each threshold |
| Elston multiplicative index | Passing rate, margins among survivors, selected set, and observed differential | Threshold sensitivity and validation |
| Within-cohort sequential screen | Attrition at each screen, selected set, and observed differential | Screen-order sensitivity and operational cost |
| Generalised index | Expected response in the objective traits, accuracy for the economic form, and information-deletion efficiency | Prediction-error propagation and validation |
| Quadratic genomic selection index | Linear and quadratic score contributions, observed genomic estimated breeding value differential, approximate expected gains, and selection agreement with a linear benchmark | Out-of-sample utility, true-value accuracy in simulation, and multi-cycle response |

The absence of one criterion is not evidence of poor quality. For example,
independent culling has no linear-index accuracy. Its purpose is to enforce
limits. Conversely, a high $R_{HI}$ cannot rescue a method that violates a
mandatory disease threshold.

## 7. A comprehensive comparison analysis

### 7.1 State the objective

The example seeks gains in grain yield (GY), ears per plant (EPP), and lower
values for plant height (PHT), anthesis date (AD), anthesis-silking interval
(ASI), and grey leaf spot score (GLS). The target is expressed in genetic
standard deviations.

```{r comparison-objective}
desired_gains_sd <- c(
  GY = 1.0,
  PHT = 0.4,
  AD = 0.6,
  ASI = 0.5,
  EPP = 0.4,
  GLS = 0.6
)
```

First test feasibility. This separates an unrealistic target from a poor
method.

```{r comparison-feasibility}
feasibility <- gain_feasibility(
  desired_gains_sd,
  dgr_G,
  dgr_P,
  n_candidates = nrow(dgr_candidates),
  n_select = 20L,
  lower_is_better = lower_is_better,
  gain_units = "genetic_sd"
)
feasibility
```

### 7.2 Put every method on one explicit scale

The following transformation expresses candidates and covariance matrices in
genetic-standard-deviation units. `selection_index()` then receives
`scale_traits = FALSE`. Hence, the desired-gain vector and every reported
response use the same units.

```{r prepare-comparison}
genetic_sd <- sqrt(diag(dgr_G))[traits]
inverse_sd <- diag(1 / genetic_sd, nrow = length(traits))
dimnames(inverse_sd) <- list(traits, traits)

G_sd <- inverse_sd %*% dgr_G %*% inverse_sd
P_sd <- inverse_sd %*% dgr_P %*% inverse_sd

candidates_sd <- dgr_candidates
centred <- sweep(
  as.matrix(dgr_candidates[, traits]),
  2L,
  colMeans(dgr_candidates[, traits]),
  "-"
)
candidates_sd[, traits] <- sweep(centred, 2L, genetic_sd, "/")
```

The economic weights below are implied by the desired-gain direction. This
creates one common merit definition for the mathematical comparison.

```{r common-merit}
common_weights <- implied_economic_weights(
  desired_gains_sd,
  G_sd,
  P_sd,
  lower_is_better = lower_is_better,
  gain_units = "trait"
)
round(common_weights, 3)
```

Negative implied weights can occur. They are mathematically valid under
correlated response. The desired response remains the biological statement to
interpret.

### 7.3 Fit objective-matched methods and benchmarks

```{r fit-comparison}
common_arguments <- list(
  values = candidates_sd,
  trait_cols = traits,
  id_col = "GenoID",
  G = G_sd,
  P = P_sd,
  lower_is_better = lower_is_better,
  center_traits = FALSE,
  scale_traits = FALSE,
  n_select = 20L,
  main_trait = "GY"
)

smith_hazel <- do.call(selection_index, c(
  common_arguments,
  list(method = "smith_hazel", economic_weights = common_weights)
))

pesek_baker <- do.call(selection_index, c(
  common_arguments,
  list(
    method = "pesek_baker",
    desired_gains = desired_gains_sd,
    aggregate_weights = common_weights
  )
))

base_index <- do.call(selection_index, c(
  common_arguments,
  list(method = "base", economic_weights = common_weights)
))

rank_sum <- do.call(selection_index, c(
  common_arguments,
  list(method = "mulamba_mock")
))
```

The first two methods should agree because the economic weights were derived
from the desired gains. This is a mathematical verification. The base and
rank-sum methods provide simpler benchmarks.

Threshold methods require limits rather than a response vector. These limits
use the units of `candidates_sd`. A maximum is supplied for each trait listed
in `lower_is_better`.

```{r fit-threshold-methods}
limits_sd <- c(
  GY = -0.50,
  PHT = 0.75,
  AD = 0.75,
  ASI = 0.75,
  EPP = -0.75,
  GLS = 0.75
)

culling <- do.call(selection_index, c(
  common_arguments,
  list(
    method = "independent_culling",
    culling_thresholds = limits_sd
  )
))

elston <- do.call(selection_index, c(
  common_arguments,
  list(method = "elston", culling_thresholds = limits_sd)
))

sequential_screen <- do.call(selection_index, c(
  common_arguments,
  list(method = "tandem", tandem_order = c("GY", "GLS", "AD"))
))
```

### 7.4 Assemble the comparison

```{r run-comparison}
comparison <- compare_selection_methods(
  list(
    Smith_Hazel = smith_hazel,
    Pesek_Baker = pesek_baker,
    Base = base_index,
    Rank_sum = rank_sum,
    Culling = culling,
    Elston = elston,
    Sequential_screen = sequential_screen
  ),
  target_gains = desired_gains_sd
)
comparison
```

Read the fairness table first.

```{r comparison-fairness}
comparison$fairness
```

A culling method may select fewer than 20 candidates. That result is useful.
It shows that the limits and the requested selection quota conflict.

### 7.5 Compare biological response

```{r comparison-responses}
comparison$responses[
  , c(
    "Method", "Trait", "Expected_response",
    "Observed_differential", "Expected_attainment"
  )
]
```

Expected response and observed differential answer different questions.
Expected response predicts transmitted change under the covariance model.
Observed differential describes the candidates selected from this data set.
Rank, threshold, and sequential-screen methods retain the second quantity because a
closed-form expected response is unavailable here.

Target attainment equals expected response divided by the requested gain. A
value of 1 reaches the target for that trait. A value of 0.70 reaches 70%.
Negative values move in an unfavourable direction. Read the worst trait before
the mean because one severe failure can be hidden by strong gains elsewhere.

### 7.6 Compare summary criteria

```{r comparison-summary}
comparison$summary[
  , c(
    "Method", "N_selected", "R_HI", "Delta_H", "RE",
    "Index_heritability", "Worst_expected_attainment",
    "Mahalanobis_alignment", "Mahalanobis_residual"
  )
]
```

Use each quantity for its defined purpose.

1. Read `R_HI` and `Delta_H` only where the same aggregate merit was supplied.
2. Read `RE` as efficiency for GY alone.
3. Read index heritability as repeatability of the composite score.
4. Read worst attainment as protection against hidden trait failure.
5. Read Mahalanobis alignment as agreement with the complete desired
   direction.
6. Read the residual as departure from the requested proportions.

### 7.7 Compare rankings and selected candidates

```{r comparison-decisions}
round(comparison$rank_correlation, 2)
round(comparison$selected_jaccard, 2)
comparison$selected_overlap
```

Rank correlation uses every candidate. Jaccard similarity uses the selected
sets. A high rank correlation can coexist with a different parent list near
the selection boundary. Therefore, report both.

### 7.8 Compare the closed-form and iteratively optimised desired-gain index

The iterative procedure remains part of the desired-gain family. Therefore,
compare it first with its closed-form starting index. Use the same candidates,
covariance matrices, trait directions, and selected count.

`run_dgsi()` expresses its target in candidate standard-deviation units. The
closed-form example above uses genetic standard deviations. The conversion
must be explicit.

```{r fit-iterative-dgsi, eval=FALSE}
candidate_sd <- vapply(
  dgr_candidates[, traits], stats::sd, numeric(1L)
)
target_original_units <- desired_gains_sd * genetic_sd
target_candidate_sd <- target_original_units / candidate_sd

iterative_dgsi <- run_dgsi(
  init_data = dgr_candidates["GenoID"],
  cand_data = dgr_candidates,
  trait_cols = traits,
  dg = target_candidate_sd,
  G = dgr_G,
  P = dgr_P,
  lower_is_better = lower_is_better,
  n_select = 20L,
  n_iter = 500L,
  n_rep = 10L,
  seed = 2026L
)
```

Construct one objective and include the iterative result in the same call.

```{r compare-iterative-objective, eval=FALSE}
comparison_target <- comparison_objective(
  desired_gains = desired_gains_sd,
  aggregate_weights = economic_weights,
  G = dgr_G,
  gain_units = "genetic_sd"
)

cross_family <- compare_selection_methods(
  list(
    Pesek_Baker = pesek_baker,
    Iterative_DGSI = iterative_dgsi
  ),
  objective = comparison_target,
  validation_data = dgr_candidates
)
```

The direct calculations below show the components returned by the unified
comparison.

```{r compare-iterative-dgsi, eval=FALSE}
closed_rank <- setNames(
  pesek_baker$ranking$rank,
  pesek_baker$ranking$id
)
iterative_rank <- setNames(
  iterative_dgsi$ranked_geno$Rank,
  iterative_dgsi$ranked_geno$GenoID
)

rank_correlation <- cor(
  closed_rank[names(closed_rank)],
  iterative_rank[names(closed_rank)],
  method = "spearman"
)
selected_overlap <- length(intersect(
  pesek_baker$selected$id,
  iterative_dgsi$selected_geno$GenoID
))

response_genetic_sd <- rbind(
  closed_form = pesek_baker$evaluation$expected_response,
  iterative =
    iterative_dgsi$theoretical_response$original_units / genetic_sd
)
attainment <- sweep(
  response_genetic_sd,
  2L,
  desired_gains_sd,
  "/"
)
```

Read four results: (i) transmitted response, (ii) target attainment, (iii)
rank correlation, and (iv) selected overlap. Then inspect the iterative
holdout objective and replicate stability. Improvement on the fitted candidate
set alone is insufficient.

### 7.9 Compare a linear and quadratic genomic index

A quadratic genomic selection index (QGSI) should have a linear genomic
selection index (LGSI) benchmark. Use the same genomic estimated breeding
values (GEBVs), linear economic weights, genomic covariance, transformation,
and selected count. Set the quadratic matrix to zero for the linear benchmark.

```{r compare-linear-quadratic, eval=FALSE}
q_traits <- c("GY", "PHT", "GLS")
linear_values <- c(GY = 1.0, PHT = 0.2, GLS = 0.4)
Gamma <- stats::cov(dgr_gebv[, q_traits])

W_zero <- matrix(
  0, length(q_traits), length(q_traits),
  dimnames = list(q_traits, q_traits)
)
W_curved <- W_zero
diag(W_curved) <- c(-0.05, -0.02, -0.04)

lgsi <- run_qgsi(
  init_data = dgr_gebv["GenoID"],
  gebv_data = dgr_gebv,
  trait_cols = q_traits,
  linear_weights = linear_values,
  W = W_zero,
  Gamma = Gamma,
  lower_is_better = c("PHT", "GLS"),
  n_select = 20L
)
qgsi <- run_qgsi(
  init_data = dgr_gebv["GenoID"],
  gebv_data = dgr_gebv,
  trait_cols = q_traits,
  linear_weights = linear_values,
  W = W_curved,
  Gamma = Gamma,
  lower_is_better = c("PHT", "GLS"),
  n_select = 20L
)

linear_rank <- setNames(lgsi$ranked_geno$Rank, lgsi$ranked_geno$GenoID)
quadratic_rank <- setNames(qgsi$ranked_geno$Rank, qgsi$ranked_geno$GenoID)
cor(linear_rank, quadratic_rank, method = "spearman")
length(intersect(
  lgsi$selected_geno$GenoID,
  qgsi$selected_geno$GenoID
))
lgsi$expected_gain_per_trait
qgsi$expected_gain_per_trait
qgsi$component_summary
```

The numerical weights above illustrate the comparison mechanics. They are not
a breeding recommendation. A quadratic analysis requires biological or
economic justification for every squared and cross-product term.

Evaluate the two fits using: (i) per-trait response, (ii) rank and selected-set
agreement, (iii) candidate-specific linear and quadratic contributions, and
(iv) out-of-sample nonlinear utility. In simulation, also compare accuracy
against true genetic merit and multi-cycle response. The linear model is
preferred when curvature does not improve validated utility or change the
decision reliably.

### 7.10 Compare every fitted family under one objective

The final comparison uses one declared yardstick. `common_G` is the genetic
covariance in original trait coordinates. `validation_values` contains one row
per fitted candidate and the same objective traits. These values can be
independent predictions, simulated true genetic values, or later measurements.
Their provenance determines the strength of the evidence.

```{r unified-cross-family-comparison, eval=FALSE}
common_objective <- comparison_objective(
  desired_gains = desired_gains_sd,
  aggregate_weights = common_aggregate_weights,
  W = common_quadratic_utility,
  G = common_G,
  gain_units = "genetic_sd"
)

all_methods <- compare_selection_methods(
  list(
    Smith_Hazel = smith_hazel,
    Restricted = restricted_fit,
    General = general_fit,
    Desired_gain = pesek_baker,
    Iterative_DGSI = iterative_dgsi,
    Linear_genomic = lgsi,
    Quadratic_genomic = qgsi
  ),
  objective = common_objective,
  validation_data = validation_values
)

all_methods$fairness
all_methods$responses
all_methods$validation_responses
all_methods$validation_utility
all_methods$summary
all_methods$rank_correlation
all_methods$selected_jaccard
```

The `R_HI`, `Delta_H`, and `Index_accuracy` columns retain native family
criteria. They can describe different merits. `Common_merit_response` is the
response in the linear component of the declared common utility. Use it for a
cross-family conclusion about that component. `Validation_utility_response`
evaluates the complete utility on the supplied validation values. A non-zero
quadratic term requires these values because mean trait responses do not
determine the mean of a quadratic function.

## 8. Add uncertainty and validation

The point comparison above is the beginning of the analysis. A breeding
recommendation also needs stability evidence.

### 8.1 Covariance and weight uncertainty

Use `index_uncertainty()` to resample the genetic and phenotypic covariance
matrices. Use `weight_sensitivity()` when economic weights are uncertain.
Report coefficient intervals, response intervals, rank stability, and
selected-set stability.

### 8.2 Candidate prediction uncertainty

Genomic estimated breeding values (GEBVs) have prediction error. When
candidate-specific prediction error covariance (PEC) is available, use
`candidate_score_uncertainty()`. Report each candidate's score interval and
selection probability. Give particular attention to candidates near the
selection boundary.

### 8.3 Value of each information source

Use `index_information_efficiency()` for a Smith-Hazel or general economic
index. It applies Cunningham's deletion-efficiency principle. A trait with
little information value can then be removed from future phenotyping plans if
the saving is operationally important.

### 8.4 Out-of-sample validation

Fit the index with training data. Fix its coefficients and transformations.
Apply `predict()` to a later cohort or a held-out set. Compare predicted
responses with realised breeding values, progeny means, or later-stage trial
performance. This separates genuine prediction from reuse of the fitting data.

### 8.5 Multi-cycle validation

Use `simulate_selection_cycles()` when the decision concerns several cycles.
Compare cumulative response, joint target attainment, genetic variance, and
diversity. Use matched scenarios and replicate seeds. Read [Multi-cycle
simulation](DesiredGainR-simulation.html) for calibration and stress testing.

## 9. Choosing the preferred method

A method is preferred when five conditions are satisfied.

1. It represents the breeder's actual objective.
2. Its required inputs have adequate quality.
3. Its expected response respects every critical trait.
4. Its ranking remains stable under credible uncertainty.
5. Its advantage persists in validation or relevant multi-cycle scenarios.

The method with the largest single statistic can fail this standard. For
example, a high aggregate-merit response can hide failure of a disease limit.
A stable rank can also follow the wrong biological direction. Therefore, the
final recommendation should report the complete decision panel.

Close the analysis with a practical statement. Name the selected method.
State why it matches the objective. Report the expected response for every
trait. State the main uncertainty. Then describe how the decision will be
validated in the breeding programme.

## 10. Package coverage and regression tests

DesiredGainR maintains an explicit contract for its public interface. The
contract currently covers 43 exported functions. Every export must have a
manual alias and at least one behavioural test file. Adding an exported
function without both entries causes the test suite to fail.

The family-level tests cover the complete method set described in this
vignette. They fit six classical index families and two operational
comparators on one labelled data set. They
also fit all five restricted families and both general-index objectives. The
tests check the method returned, the selected candidates, trait responses,
restriction residuals, target attainment, rank agreement, and selected-set
agreement. Separate tests cover DGSI, QGSI, uncertainty, feasibility,
simulation, covariance handling, and data validation.

Contributors should run the following commands after changing code or
documentation.

```r
devtools::document()
devtools::test()
devtools::check()
```

`devtools::document()` rebuilds the manual files from the documentation beside
each R function. `devtools::test()` checks scientific and programming
contracts. `devtools::check()` checks the complete source package in a clean
build environment.

## 11. References

- Beavis W, Lamkey K, Mahama AA, Suza W (2023). Multiple Trait Selection. In
  Suza WP and Lamkey KR, editors, *Quantitative Genetics for Plant Breeding*.
  Iowa State University Digital Press.
- Cunningham EP (1969). The relative efficiencies of selection indexes.
  *Acta Agriculturae Scandinavica* **19**:45-48.
- Elston RC (1963). A weight-free index for the purpose of ranking or selection
  with respect to several traits at a time. *Biometrics* **19**:85-97.
- Kempthorne O, Nordskog AW (1959). Restricted selection indices. *Biometrics*
  **15**:10-19.
- Mulamba NN, Mock JJ (1978). Improvement of yield potential of the Eto Blanco
  maize population by breeding for plant traits. *Egyptian Journal of Genetics
  and Cytology* **7**:40-51.
- Pesek J, Baker RJ (1969). Desired improvement in relation to selection
  indices. *Canadian Journal of Plant Science* **49**:803-804.
- Satoh M (2024). Characteristics of restricted selection indices and
  geometrical interpretation of restricted breeding values. *Journal of
  Animal Breeding and Genetics* **141**:353-363.
- Smith HF (1936). A discriminant function for plant selection. *Annals of
  Eugenics* **7**:240-250.
- Yamada Y, Yokouchi K, Nishida A (1975). Selection index when genetic gains of
  individual traits are of primary concern. *Japanese Journal of Genetics*
  **50**:33-41.
