---
title: "Introduction to DesiredGainR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to DesiredGainR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(DesiredGainR)
data(dgr_traits, package = "DesiredGainR")
data(dgr_G, package = "DesiredGainR")
data(dgr_P, package = "DesiredGainR")
data(dgr_candidates, package = "DesiredGainR")
traits <- dgr_traits$trait
lower_is_better <- dgr_traits$trait[dgr_traits$direction == "decrease"]
```

## 1. Overview

Computing a multi-trait selection index is arithmetic. Deciding what the index
should select for is the difficult part, and it is the part that determines
whether the resulting selection is any good.

The literature is consistent on this point. Guimarães et al. (2021) describe
the assignment of meaningful economic weights as the principal bottleneck to
routine index use, and demonstrate the consequence empirically: when they
supplied arbitrary weights to a Smith-Hazel and to a Tai index, the gains those
indices had delivered under sensible weights disappeared entirely.
Covarrubias-Pazaran (2021), writing the CGIAR implementation guideline, reaches
the same conclusion from the operational side and advises breeders not to
interpret index coefficients at all, because the desired response is the only
decision genuinely available to them.

DesiredGainR is therefore organised around the objective rather than around the
index. It provides the classical index families, but its distinctive content is
the layer that helps a breeder state an objective, test whether it can be
attained, and discover how much the resulting decision depends on it.

This vignette orients the reader. The other articles cover each layer in
depth, and Section 6 provides a direct link to each one.

---

## 2. What the package does, and what it does not

DesiredGainR begins after the genetic evaluation. It consumes genetic
predictions and covariance matrices produced by a fitted multi-trait model
elsewhere, and it never estimates them from raw phenotypes.

That boundary is deliberate. A covariance matrix of raw phenotypes contains
residual and environmental variation, and using it as a genetic covariance
silently mislabels the estimand. The package therefore requires `G` explicitly,
and where no fitted matrix exists,
[`estimate_genetic_covariance()`](../reference/estimate_genetic_covariance.html)
provides labelled approximations that record what they actually are.

| DesiredGainR does | DesiredGainR does not |
|---|---|
| Translate between economic weights and desired gains | Fit the multi-trait genetic model |
| Test whether a stated objective is attainable | Estimate breeding values or marker effects |
| Build and compare classical index families | Design a crossing plan or allocate matings |
| Report the standard index-evaluation criteria | Analyse field trials or adjust for design |
| Simulate desired-gain directions over cycles | Recommend which candidates to cross |

The last exclusion matters for anyone using both of the author's packages.
Parent, cross and mating decisions belong to a dedicated mating-design tool;
see [Working with other packages](DesiredGainR-interoperation.html).

---

## 3. The four layers

### 3.1 Defining the objective

The Smith-Hazel economic index and the Pesek-Baker desired-gain index are two
parameterisations of one linear index, so each desired-gain vector corresponds
to exactly one economic-weight vector. That duality is the foundation of the
layer: a breeder who cannot state weights but can state relative gains gets the
weights for free, and one who proposes weights can be shown what response they
actually request.

Around it sit a feasibility test, a method for recovering the objective a
programme has been using implicitly, and a sensitivity analysis that reports
whether the decision depends on the weights at all.

### 3.2 Building an index

`selection_index()` provides Smith-Hazel, the base index, both desired-gain
formulations, the Mulamba-Mock rank-sum index, and independent culling and
tandem selection as comparators, through one interface so that alternatives can
be compared on identical data. `run_dgsi()` implements the iterative
desired-gain method of Joukhadar et al. (2024), and `run_qgsi()` the quadratic
genomic index of Cerón-Rojas et al. (2026).

### 3.3 Evaluating an index

`evaluate_index()` reports: (i) the correlation between the index and one fixed
definition of aggregate genetic merit, $R_{HI}$; (ii) expected change in that
merit, $\Delta H$; (iii) expected response for every trait, $\Delta_j$; (iv)
relative efficiency (RE) for one declared main trait; (v) index heritability;
and (vi) the coefficient of variation of index scores, $CV_I$, when the score
has a meaningful non-zero mean.

These quantities are not interchangeable measures of index quality. In
particular, $R_{HI}$ and $\Delta H$ require the same aggregate-merit definition
for every index, RE describes only the main trait, and $CV_I$ is undefined for
a centred index. Therefore, DesiredGainR also directs the breeder to per-trait
attainment, response-direction alignment, selected-set stability, covariance
uncertainty and, for repeated cycles, diversity. The detailed interpretation is
given in [Choosing an index](DesiredGainR-index-families.html) and the breeder
guide.

### 3.4 Looking ahead

A single-cycle response calculation cannot distinguish desired-gain directions
beyond what the achievable-response ellipsoid already states. Over several
cycles the ranking can change, because truncation selection erodes the variance
that response depends on. The simulation layer represents that, using founders
built from the breeder's own phased marker data rather than a simulated genome.

---

## 4. The example programme

The shipped data describe a simulated tropical maize programme with six traits.
The material is patterned on a real programme structure but represents no real
germplasm and no real trial.

```{r traits}
dgr_traits
```

Three properties make these data useful for demonstration, and the previous
example data lacked all three.

**The objective is internally antagonistic.** Grain yield correlates positively
with anthesis date, yet the programme wants yield to rise while the cycle
shortens.

```{r correlation}
round(stats::cov2cor(dgr_G)["GY", ], 2)
```

**The trait scales differ by a factor of about one hundred and twenty**, from
ears per plant to plant height. That is what makes standardisation and matrix
conditioning matter rather than being formalities.

```{r scales}
round(range(dgr_traits$genetic_sd), 2)
```

**The population is internally consistent.** Genetic values were generated from
ninety quantitative trait loci drawn from the shipped marker panel, then
adjusted so that their realised covariance equals `dgr_G` to numerical
precision, and heritabilities are recoverable exactly.

```{r consistency}
round(diag(dgr_G) / diag(dgr_P), 2)
```

The nine datasets are documented together under `?"DesiredGainR-data"`.

---

## 5. A short worked path

The full narrative is in the *Complete workflow* vignette. This is the shape of
it in four calls.

**Check the covariance matrices before building anything on them.** Index
coefficients come from inverting these matrices, and an ill-conditioned matrix
produces coefficients that are numerically meaningless while looking plausible.

```{r conditioning}
matrix_diagnostics(dgr_G, "G")$condition_number
matrix_diagnostics(stats::cov2cor(dgr_G))$condition_number
```

Almost all of that conditioning comes from the trait scales rather than the
correlations, which is why the package standardises by default.

**State the objective and see what it implies.**

```{r implied}
desired_gains <- c(
  GY = 1.0, PHT = 0.4, AD = 0.6, ASI = 0.5,
  EPP = 0.4, GLS = 0.6
)
round(
  implied_economic_weights(
    desired_gains, dgr_G, dgr_P,
    lower_is_better = lower_is_better, gain_units = "genetic_sd"
  ),
  2
)
```

**Ask whether it can be attained.**

```{r feasibility}
gain_feasibility(
  desired_gains, dgr_G, dgr_P,
  n_candidates = nrow(dgr_candidates), n_select = 20L,
  lower_is_better = lower_is_better, gain_units = "genetic_sd"
)
```

It cannot, and not marginally: the target would require selecting fewer than
one candidate from two hundred. Roughly fifty-five per cent of the requested
gain is available at the intensity actually planned. Knowing that before
optimising is worth more than any subsequent refinement of the index.

**Build the index.**

```{r index}
selection_index(
  dgr_candidates, traits,
  method = "smith_hazel",
  G = dgr_G, P = dgr_P,
  economic_weights = c(
    GY = 1.0, PHT = 0.2, AD = 0.5,
    ASI = 0.4, EPP = 0.3, GLS = 0.5
  ),
  lower_is_better = lower_is_better, n_select = 20L, main_trait = "GY"
)
```

---

## 6. Which vignette to read next

| If you want to | Read |
|---|---|
| Follow one programme end to end | [The complete workflow](DesiredGainR-workflow.html) |
| Decide what to select for | [Defining a breeding objective](DesiredGainR-objective.html) |
| Define ranges when exact gains are uncertain | [Defining desired gains and acceptable intervals](DesiredGainR-desired-gain-intervals.html) |
| Obtain or judge a covariance matrix | [Obtaining G and P](DesiredGainR-covariance.html) |
| Choose among index families | [Choosing an index](DesiredGainR-index-families.html) |
| Understand the iterative desired-gain method | [Optimising desired gains](DesiredGainR-dgsi.html) |
| Use the quadratic genomic index | [Quadratic genomic selection](DesiredGainR-qgsi.html) |
| Compare objectives over several cycles | [Multi-cycle simulation](DesiredGainR-simulation.html) |
| Inspect validation with real programmes | [Empirical validation](DesiredGainR-empirical-validation.html) |
| Combine this with other breeding software | [Working with other packages](DesiredGainR-interoperation.html) |

A plain-language companion for readers who will approve a selection decision
but do not run R is available through `open_desiredgain_guide()`.

---

## 7. Three conventions worth knowing before you start

**Direction is declared, not signed.** Every entry point accepts
`lower_is_better`, naming the traits that improve by falling. Gains and weights
are then stated as improvements throughout, and the package orients the
covariance matrices internally. Signing values by hand is the commonest way to
obtain a confidently wrong answer.

**Units are stated explicitly.** Desired gains can be given in original trait
units, genetic standard deviations, or phenotypic standard deviations, through
`gain_units`. Results are returned in the units they were asked for.

**Coefficients are not comparable across traits.** A coefficient carries
inverse trait units, so a trait measured on a small scale receives a large
number for the same emphasis. `effective_weights()` reports the interpretable
form, the coefficient multiplied by the trait's genetic standard deviation.

---

## 8. References

- Cerón-Rojas JJ, Montesinos-López OA, Montesinos-López A, et al. (2026).
  Nonlinear genomic selection index accelerates multi-trait crop improvement.
  *Nature Communications* **17**:1991.
  <https://doi.org/10.1038/s41467-026-69890-3>
- Covarrubias-Pazaran G (2021). *Practical implementation of selection
  indices.* CGIAR Excellence in Breeding.
- Guimarães PHR, Melo PGS, Cordeiro ACC, Torga PP, Rangel PHN, de Castro AP
  (2021). Index selection can improve the selection efficiency in a rice
  recurrent selection population. *Euphytica* **217**:95.
  <https://doi.org/10.1007/s10681-021-02819-7>
- Joukhadar R, Li Y, Thistlethwaite R, Forrest KL, Tibbits JF, Trethowan R,
  Hayden MJ (2024). Optimising desired gain indices to maximise selection
  response. *Frontiers in Plant Science* **15**:1337388.
  <https://doi.org/10.3389/fpls.2024.1337388>
- Rahimi M, Debnath S (2023). Estimating optimum and base selection indices in
  plant and animal breeding programs by development new and simple SAS and R
  codes. *Scientific Reports* **13**:18977.
  <https://doi.org/10.1038/s41598-023-46368-6>
