---
title: "Optimising desired gains"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Optimising desired gains}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
library(DesiredGainR)
data(dgr_traits, package = "DesiredGainR")
data(dgr_G, package = "DesiredGainR")
data(dgr_P, package = "DesiredGainR")
data(dgr_candidates, package = "DesiredGainR")
traits <- dgr_traits$trait
lower_is_better <- dgr_traits$trait[dgr_traits$direction == "decrease"]
```

## 1. What the iterative method changes

The classical desired-gain coefficient solve fixes a response direction under
the covariance model. The iterative desired-gain selection index (DGSI) of
Joukhadar et al. searches nearby desired-gain vectors and evaluates the
selected candidates directly. Its purpose is to reduce disagreement between
the requested standardised gain vector and the differential of the selected
group.

`run_dgsi()` does not discover the breeding objective. The breeder must still
state the direction, minimum requirements or acceptable intervals. Test
attainability before interpreting an optimisation result.

### DGSI is not fused into `selection_index()`

The functions have separate contracts.

| Function | What it fits |
|---|---|
| `selection_index(method = "pesek_baker")` | The closed-form Pesek–Baker desired-gain index. |
| `selection_index(method = "yamada")` | The closed-form Yamada desired-gain index. For a square, invertible $\mathbf{G}$ it has the same coefficient direction as Pesek–Baker. |
| `run_dgsi()` | An iterative search over proposed desired-gain directions, evaluated through the candidates selected by each proposal. |

`selection_index()` never starts the iterative search. Conversely,
`run_dgsi()` retains its own non-iterated Yamada solution in
`dgsi$non_iterated`, so the breeder can determine what iteration changed.

For a proposed direction $\mathbf{d}$, the coefficient map is

$$
\mathbf{b}(\mathbf{d})=
\mathbf{P}^{-1}\mathbf{G}
(\mathbf{G}^{\mathsf T}\mathbf{P}^{-1}\mathbf{G})^{-1}\mathbf{d}.
$$

After ranking candidates with $\mathbf{b}(\mathbf{d})$, the search compares the
standardised differential $r_j$ with the requested gain $d_j$ through

$$
L(\mathbf{d})=\sum_j v_j
\left\{\frac{r_j-d_j}{\max(|d_j|,0.25)}\right\}^{2},
$$

where $v_j$ is the optional `objective_weights` value. The denominator floor
prevents a near-zero target from dominating the search. Because the selected
set changes in steps as coefficients move, the objective is not smooth and
the implementation uses a reproducible derivative-free search.

```{r feasibility}
desired_gains <- c(
  GY = 1.0, PHT = 0.4, AD = 0.6,
  ASI = 0.5, EPP = 0.4, GLS = 0.6
)

gain_feasibility(
  desired_gains, dgr_G, dgr_P,
  n_candidates = nrow(dgr_candidates), n_select = 20L,
  lower_is_better = lower_is_better,
  gain_units = "genetic_sd"
)
```

If the exact ray is unattainable, optimisation can find a closer selected set;
it cannot create unavailable genetic variation. If the entries are minimum
floors rather than an exact ratio, use
[`suggest_desired_gains()`](../reference/suggest_desired_gains.html) or interval
optimisation and assess joint attainment instead.

---

## 2. Fit independent searches

The example below is deliberately small enough for a vignette. Operational
analyses should use more iterations and independent replicates, followed by
holdout or external validation.

```{r fit-dgsi}
dgsi <- run_dgsi(
  init_data = dgr_candidates["GenoID"],
  cand_data = dgr_candidates,
  trait_cols = traits,
  dg = desired_gains,
  G = dgr_G,
  P = dgr_P,
  lower_is_better = lower_is_better,
  n_select = 20L,
  n_iter = 60L,
  n_rep = 3L,
  seed = 742L,
  replicate_selection = "holdout"
)
dgsi
```

`dg` and `realised_response` use candidate standard-deviation units. The
coefficient solve still respects the original units of $\mathbf{G}$ and
$\mathbf{P}$; the package performs the conversion internally.

The operational output is the ranked candidate table, not the coefficient
vector by itself.

```{r ranking}
head(dgsi$ranked_geno[, c(
  "GenoID", "SelectionIndex", "Rank", "Eligible", "Selected"
)])
dgsi$selected_geno[, c("GenoID", "SelectionIndex", "Rank")]
```

`Eligible` matters only when `select_mode = "eligible_top_n"`. In the default
`top_n` mode every complete candidate is eligible and the index selects the
requested number with deterministic tie-breaking.

---

## 3. Separate the selected differential from transmitted response

```{r responses}
round(rbind(
  requested = dgsi$desired_gain,
  optimised_direction = dgsi$optimised_d,
  selected_differential = dgsi$realised_response,
  expected_genetic_response_sd = dgsi$theoretical_response$standardised
), 3)
```

The selected differential describes the candidates retained. The theoretical
response is the model-based expectation transmitted to the next generation.
They are not synonyms. A large selected differential with a small theoretical
response indicates that much of the observed separation is not expected to be
inherited.

---

## 4. Inspect optimisation stability

```{r stability}
dgsi$replicate_diagnostics
dgsi$rank_correlation
dgsi$selected_set_agreement
```

The winning replicate is selected on held-out candidates by default. Report:

1. the training and holdout objectives;
2. whether the search reached a plateau;
3. coefficient and rank stability across replicates;
4. selected-set agreement;
5. the response for every trait; and
6. the covariance source and its uncertainty.

If independent replicates produce different selected sets with similar
objectives, the data do not identify one operational recommendation. Increase
the evidence or present the stable candidate core rather than hiding the
instability.

---

## 5. Compare with the non-iterated solution

```{r non-iterated}
comparison <- data.frame(
  Trait = traits,
  Requested = unname(dgsi$desired_gain[traits]),
  Non_iterated = unname(dgsi$non_iterated$realised_response[traits]),
  Iterative = unname(dgsi$realised_response[traits])
)
comparison$Absolute_error_non_iterated <- abs(
  comparison$Non_iterated - comparison$Requested
)
comparison$Absolute_error_iterative <- abs(
  comparison$Iterative - comparison$Requested
)
comparison[-1] <- round(comparison[-1], 3)
comparison
```

Compute both objective values from the full candidate set so that the
comparison uses the same observations. The stored `dgsi$objective` can instead
be a holdout objective and must not be compared directly with a full-sample
quantity.

```{r objective-comparison}
dgsi_loss <- function(response, target) {
  sum(((response - target) / pmax(abs(target), 0.25))^2)
}

c(
  non_iterated_full_sample = dgsi_loss(
    dgsi$non_iterated$realised_response, dgsi$desired_gain
  ),
  iterative_full_sample = dgsi_loss(
    dgsi$realised_response, dgsi$desired_gain
  )
)
```

In this reproducible run, the full-sample loss falls from about 3.280 to
2.108, a reduction of approximately 36%. That improvement is not uniform
across traits. The anthesis-silking interval changes from an unfavourable
differential to a favourable one, and plant height and anthesis date move much
closer to their requests. Grain yield, ears per plant and grey leaf spot do not
all become closer individually. DGSI minimises the declared multi-trait loss;
it does not guarantee a smaller error for every trait. The trait table must
therefore accompany the total objective.

The iterative method has value only if the improvement is meaningful on
held-out candidates and stable across searches. A smaller training objective
alone is not sufficient evidence.

The internal holdout is used to choose among stochastic replicates. It is not
an independent breeding cycle and does not establish transportability. When a
later cohort is available, pass it through `validation_data` and retain it
outside coefficient fitting.

---

## 6. What constitutes a sufficient DGSI demonstration

A defensible analysis reports all of the following:

1. the requested gain vector and whether it represents an exact ratio or
   minimum floors;
2. the exact-ray feasibility result at the planned selection intensity;
3. the non-iterated Yamada response;
4. the optimised direction, selected differential and expected transmitted
   response;
5. the selected candidates and their index ranks;
6. training and holdout objectives;
7. rank, coefficient and selected-set stability across replicates; and
8. validation in a later cohort when one is available.

The demonstration above now covers items 1–7 with the shipped population.
Item 8 necessarily requires data from a genuinely later breeding cohort.

For objectives stated as intervals rather than one exact vector, read
[Defining desired gains and acceptable intervals](DesiredGainR-desired-gain-intervals.html).

## References

Joukhadar R, Daetwyler HD, Bansal U, Gendall AR, Hayden MJ (2024). An iterative
desired-gain selection index for simultaneous improvement of multiple traits.
