---
title: "Full pipeline: from trait values to a defended selection decision"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Full pipeline: from trait values to a defended selection decision}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(DesiredGainR)
data(dgr_traits, package = "DesiredGainR")
data(dgr_G, package = "DesiredGainR")
data(dgr_P, package = "DesiredGainR")
data(dgr_candidates, package = "DesiredGainR")
data(dgr_gebv, package = "DesiredGainR")
data(dgr_history, package = "DesiredGainR")
data(dgr_hap1, package = "DesiredGainR")
data(dgr_hap2, package = "DesiredGainR")
data(dgr_map, package = "DesiredGainR")
```

## Purpose

This vignette assumes no prior knowledge of DesiredGainR. It runs the entire
package on the shipped example programme, one stage at a time, showing every
call, every output, and the decision each one feeds. Nothing is skipped and
nothing is left to be inferred from another document.

Sixteen stages in four phases:

| Phase | Stages | Question answered |
|---|---|---|
| **Evidence** | 1-4 | What do we have, and can it be trusted? |
| **Objective** | 5-8 | What should we select for, and is it possible? |
| **Index** | 9-12 | Which index, and what does it actually do? |
| **Looking ahead** | 13-16 | Does this objective still hold over several cycles? |

Every stage names the vignette that carries its full statistical detail. This
document is the connective walkthrough; those are the references.

---

# Phase 1: Evidence

## Stage 1 — What you need before you start

DesiredGainR begins *after* the genetic evaluation. It does not analyse field
trials, does not fit a mixed model, and does not estimate breeding values. You
must arrive with four things.

| Input | What it is | Where it comes from |
|---|---|---|
| Trait values | One adjusted mean, BLUP or GEBV per candidate per trait | Your multi-trait genetic evaluation |
| $\mathbf{G}$ | Genetic variance-covariance matrix | The same fitted model |
| $\mathbf{P}$ | Phenotypic variance-covariance matrix | The same fitted model |
| Trait directions | Which traits improve by rising, which by falling | The product profile |

If you have no fitted $\mathbf{G}$, Stage 3 shows the supported
approximations. If you have no trait directions written down, stop and write
them down: every later stage depends on them.

## Stage 2 — Load and inspect the data

```{r stage2}
dgr_traits
```

Read this table before anything else. It carries the units, the direction of
improvement, the heritability and the genetic standard deviation for each
trait. Two columns decide most of what follows.

```{r stage2-directions}
traits <- dgr_traits$trait
lower_is_better <- dgr_traits$trait[dgr_traits$direction == "decrease"]
lower_is_better
```

`lower_is_better` is passed to nearly every function in the package. Supplying
it means you state gains as *improvements* everywhere, and the package handles
the sign conventions internally. Signing values by hand is the commonest source
of a confidently wrong answer.

```{r stage2-candidates}
str(dgr_candidates, max.level = 1)
head(dgr_candidates[, c("GenoID", "GY", "PHT", "GLS")], 3)
```

## Stage 3 — Check the covariance matrices

Every index in this package is computed by inverting $\mathbf{P}$,
$\mathbf{G}$, or both. An ill-conditioned matrix produces coefficients that are
numerically meaningless while looking entirely plausible, so this check comes
before anything is built on top of them.

```{r stage3}
diagnostics_G <- matrix_diagnostics(dgr_G, "G")
c(
  condition_number = diagnostics_G$condition_number,
  positive_definite = diagnostics_G$positive_definite,
  numerical_rank = diagnostics_G$numerical_rank
)
```

A condition number near $1.9 \times 10^{4}$ looks alarming until it is
decomposed. Almost all of it comes from the trait scales rather than from the
correlations:

```{r stage3-decompose}
matrix_diagnostics(stats::cov2cor(dgr_G))$condition_number
```

Standardising the traits removes three orders of magnitude of ill-conditioning.
This is the practical form of an observation by Crosbie et al. (1980): equal
weights applied to unstandardised traits concentrate selection on whichever
trait happens to carry the largest variance. It is why DesiredGainR
standardises by default.

**If you have no fitted $\mathbf{G}$.** Use
`estimate_genetic_covariance()`, which returns labelled approximations rather
than relabelling a covariance of predictions as a genetic covariance:

```{r stage3-estimate}
G_working <- estimate_genetic_covariance(
  values = dgr_candidates, trait_cols = traits,
  method = "adjusted_means_surrogate"
)
G_working$estimand
```

Four methods are available: `prediction_covariance`, `pev_corrected`,
`se_diagonal_corrected`, `relationship_adjusted`, and
`adjusted_means_surrogate`. The last implements what the CGIAR implementation
guideline recommends for operational use when no fitted matrix exists.

## Stage 4 — Confirm units and heritabilities

The declared heritabilities must be recoverable from the two matrices. If they
are not, the metadata and the matrices are describing different things.

```{r stage4}
round(diag(dgr_G) / diag(dgr_P), 2)
dgr_traits$heritability
```

Check the spread of trait scales, because it determines whether standardisation
is optional or essential:

```{r stage4-scales}
genetic_sd <- stats::setNames(dgr_traits$genetic_sd, dgr_traits$trait)
round(sort(genetic_sd, decreasing = TRUE), 2)
round(max(genetic_sd) / min(genetic_sd))
```

A spread of 120-fold means standardisation is essential here. The package warns
automatically above fivefold.

---

# Phase 2: Objective

## Stage 5 — State a first objective

There are two ways to state a breeding objective, and they are equivalent.

**Economic weights** $\mathbf{w}$ say what a unit of each trait is worth.
**Desired gains** $\mathbf{d}$ say how much of each trait you want. The
Smith-Hazel index uses the first, $\mathbf{b} = \mathbf{P}^{-1}\mathbf{G}\mathbf{w}$;
the Pesek-Baker desired-gain index uses the second,
$\mathbf{b} = \mathbf{P}^{-1}\mathbf{G}(\mathbf{G}\mathbf{P}^{-1}\mathbf{G})^{-1}\mathbf{d}$.

Most breeders find desired gains easier, so start there. State them as
improvements, in genetic standard deviations, and let `lower_is_better` handle
direction.

```{r stage5}
desired_gains <- c(
  GY = 1.0, PHT = 0.4, AD = 0.6, ASI = 0.5,
  EPP = 0.4, GLS = 0.6
)
```

This says: one standard deviation of extra yield, four tenths of a standard
deviation shorter plants, six tenths earlier flowering, and so on.

## Stage 6 — Translate to economic weights

Setting the two index formulae equal gives an exact translation in both
directions:

$$\mathbf{w} = \mathbf{G}^{-1}\mathbf{P}\mathbf{G}^{-1}\mathbf{d},
\qquad
\mathbf{d} = \mathbf{G}\mathbf{P}^{-1}\mathbf{G}\mathbf{w}.$$

```{r stage6}
implied <- implied_economic_weights(
  desired_gains, dgr_G, dgr_P,
  lower_is_better = lower_is_better, gain_units = "genetic_sd"
)
round(implied, 2)
```

**Two of these are negative, for traits you asked to improve.** That is
correct. Once oriented, grain yield correlates $+0.55$ with the
anthesis-silking interval and $+0.45$ with ears per plant, so selecting hard
for yield would carry both *past* the smaller gains you requested. To deliver
the ratio you asked for, the index must hold them back. A negative weight never
means the trait should get worse.

**The magnitudes are not comparable.** A weight carries inverse trait units, so
a trait on a small scale gets a big number for the same emphasis. Always
rescale before comparing:

```{r stage6-rescale}
round(sort(implied * genetic_sd[names(implied)], decreasing = TRUE), 2)
```

Ears per plant had the largest raw number and is one of the smaller effects.

Confirm the translation inverts, which costs one line and catches unit errors:

```{r stage6-invert}
round(
  implied_desired_gains(
    implied, dgr_G, dgr_P,
    lower_is_better = lower_is_better, gain_units = "genetic_sd"
  ),
  3
)
```

## Stage 7 — Test whether the objective is attainable

**This is the stage most often skipped, and the one that most often changes the
answer.**

For any linear index the response is
$\mathbf{R} = i\,\mathbf{G}\mathbf{b}/\sqrt{\mathbf{b}^\mathsf{T}\mathbf{P}\mathbf{b}}$,
so the attainable responses lie on an ellipsoid and a stated target requires
exactly

$$i_{\text{required}} = \sqrt{\mathbf{d}^\mathsf{T}\mathbf{G}^{-1}\mathbf{P}\mathbf{G}^{-1}\mathbf{d}}.$$

```{r stage7}
feasibility <- gain_feasibility(
  desired_gains, dgr_G, dgr_P,
  n_candidates = nrow(dgr_candidates), n_select = 20L,
  lower_is_better = lower_is_better, gain_units = "genetic_sd"
)
feasibility
```

The request looked unremarkable and requires selecting fewer than one candidate
from two hundred. About fifty-five per cent of it is available at the intensity
actually planned:

```{r stage7-attainable}
round(feasibility$attainable_response_input_units, 3)
```

**Two consequences worth understanding now.** First, the classical desired-gain
index honours only the *direction* of $\mathbf{d}$; multiplying every element
by a constant changes nothing, because the magnitude is fixed by selection
intensity. Second, an antagonistic correlation structure can make a
modest-looking target unreachable, and no amount of optimisation will fix that.

## Stage 8 — Recover what the programme already selects for, then adjust

If your programme has been selecting without a formal index, the differentials
it achieved already encode its objective:

$$\mathbf{b} = \mathbf{P}^{-1}\mathbf{s}$$

where $\mathbf{s}$ holds the differences between the selected group and the
whole population.

```{r stage8}
recovered <- retrospective_weights(
  selected_values = dgr_candidates[dgr_history$selected, traits],
  population_values = dgr_candidates[, traits],
  trait_cols = traits
)
recovered
```

```{r stage8-differentials}
round(recovered$selection_differential_sd, 2)
```

This is how an index was introduced at both CIMMYT and IRRI. Treat it as a
starting point: it reproduces past decisions including any bias in them, and
must be adjusted deliberately against the product profile before it becomes a
forward-looking objective.

---

# Phase 3: Index

## Stage 9 — Build candidate indices

`selection_index()` provides every classical family through one interface.

```{r stage9}
economic_weights <- c(
  GY = 1.0, PHT = 0.2, AD = 0.5,
  ASI = 0.4, EPP = 0.3, GLS = 0.5
)

smith_hazel <- selection_index(
  dgr_candidates, traits,
  method = "smith_hazel",
  G = dgr_G, P = dgr_P, economic_weights = economic_weights,
  lower_is_better = lower_is_better, n_select = 20L, main_trait = "GY"
)
rank_sum <- selection_index(
  dgr_candidates, traits,
  method = "mulamba_mock",
  lower_is_better = lower_is_better, n_select = 20L
)
smith_hazel
```

| Method | Needs weights | Needs $\mathbf{G}$, $\mathbf{P}$ | Use when |
|---|---|---|---|
| `smith_hazel` | yes | yes | Weights are defensible |
| `base` | yes | no | A quick comparator; often nearly as good |
| `pesek_baker` | desired gains | $\mathbf{G}$ only | Gains easier than weights |
| `yamada` | desired gains | yes | Same index, better conditioned route |
| `mulamba_mock` | no | no | Weights unavailable or untrusted |
| `independent_culling` | thresholds | no | Non-compensatory requirements |
| `tandem` | trait order | no | Comparator for sequential selection |

## Stage 10 — Evaluate and compare

The four criteria reported are those of Rahimi and Debnath (2023), which the
established selection-index software also reports.

| Criterion | Meaning |
|---|---|
| $R_{HI}$ | Correlation between index and net merit |
| $\Delta H$ | Expected gain in aggregate merit |
| RE | Efficiency relative to direct selection on the main trait |
| $CV_I$ | Coefficient of variation of the index |

**Relative efficiency below 1 is not a failure.** It is the intended trade of
response in the main trait for response elsewhere. Every value reported by
Rahimi and Debnath was below 1.

Compare families by rank correlation *and* by overlap of the selected sets,
because the two must be consistent with each other:

```{r stage10}
paired <- merge(
  smith_hazel$ranking[, c("id", "score")],
  rank_sum$ranking[, c("id", "score")],
  by = "id", suffixes = c("_sh", "_mm")
)
round(stats::cor(paired$score_sh, paired$score_mm, method = "spearman"), 3)
length(intersect(smith_hazel$selected$id, rank_sum$selected$id))
```

A high correlation with a nearly empty intersection indicates a fault, not a
finding. That exact check caught a row-alignment bug during development.

Check which traits the index is really acting on:

```{r stage10-effective}
smith_hazel$effective_weights[, c("Trait", "Genetic_share")]
```

## Stage 11 — Optimise the desired-gain vector

`run_dgsi()` implements the iterative method of Joukhadar et al. (2024). It
samples desired-gain vectors, builds the index from each, and keeps the one
whose *realised* response in the selected set comes closest to your target.

```{r stage11}
dgsi <- run_dgsi(
  init_data = dgr_candidates[, c("GenoID", "Family")],
  cand_data = dgr_candidates,
  trait_cols = traits,
  dg = desired_gains,
  G = dgr_G, P = dgr_P,
  lower_is_better = lower_is_better,
  scale_traits = TRUE,
  n_select = 20L, n_iter = 200L, n_rep = 5L, seed = 42L
)
round(dgsi$realised_response, 3)
```

Read the deviation from target against Stage 7:

```{r stage11-deviation}
round(dgsi$realised_response - desired_gains, 3)
```

Feasibility constrained the requested *proportion*. The optimiser is not bound
to that proportion, so it distributes the shortfall unevenly, overshooting
grain yield while undershooting anthesis date. Whether that trade is acceptable
is a breeding decision, not a numerical one.

Always inspect the replicate diagnostics. The search is stochastic, so
agreement across replicates is the evidence that the answer is stable:

```{r stage11-replicates}
dgsi$replicate_diagnostics
```

And compare against the classical solution without iteration:

```{r stage11-comparator}
round(dgsi$non_iterated$realised_response, 3)
```

The anthesis-silking interval moves the *wrong way* there. Joukhadar et al.
reported precisely this failure of the unoptimised method.

## Stage 12 — Score genomic estimated breeding values

`run_qgsi()` implements the quadratic genomic selection index of Cerón-Rojas et
al. (2026):

$$I_{qg,i} = \mathbf{w}^\mathsf{T}\hat{\boldsymbol{\gamma}}_i +
\hat{\boldsymbol{\gamma}}_i^\mathsf{T}\mathbf{W}\hat{\boldsymbol{\gamma}}_i$$

where $\hat{\boldsymbol{\gamma}}_i$ is the candidate's vector of genomic
estimated breeding values, $\mathbf{w}$ the linear economic weights, and
$\mathbf{W}$ a symmetric matrix of squared and cross-product weights. Negative
diagonal curvature in $\mathbf{W}$ favours intermediate values; positive
curvature favours extremes.

```{r stage12}
quadratic_weights <- diag(c(
  GY = 0.05, PHT = -0.03, AD = -0.03, ASI = -0.04, EPP = 0.02, GLS = -0.03
))
dimnames(quadratic_weights) <- list(traits, traits)

qgsi <- run_qgsi(
  init_data = dgr_gebv["GenoID"], gebv_data = dgr_gebv,
  trait_cols = traits,
  linear_weights = economic_weights, W = quadratic_weights,
  lower_is_better = lower_is_better,
  scale_traits = TRUE, n_select = 20L
)
qgsi$expected_gain_per_trait
```

**Standardisation matters more here than anywhere else in the package**,
because economic weights multiply the breeding values directly. Left
unstandardised on these data, plant height carries half the linear index
despite holding nearly the smallest weight, and the expected gain in grain
yield comes out *negative* while the index selects for more of it.
`run_qgsi()` warns when the trait scales make this likely.

---

# Phase 4: Looking ahead

A single-cycle calculation cannot distinguish desired-gain directions beyond
what the feasibility ellipsoid already tells you. Over several cycles the
ranking can change, because truncation selection erodes the genetic variance
that the response depends on. Phase 4 represents that.

The simulation is calibrated to your own germplasm: founders are built from
real phased marker data, never from a simulated genome.

## Stage 13 — Build founders from phased marker data

```{r stage13}
founders <- founder_haplotypes(dgr_hap1, dgr_hap2, dgr_map)
founders
```

`hap1` and `hap2` are marker-by-individual matrices coded 0 or 1, giving the
allele on each chromosome copy. This is what a phase-preserving variant call
format reader returns.

**Dosage coded 0, 1, 2 is not accepted**, because it records how many
alternative alleles an individual carries but not which copy carries them, and
that co-occurrence is the linkage disequilibrium the simulation exists to
represent.

For highly inbred diploid material, dosage 0 and 2 are unambiguous, so phase
can be derived without external phasing. Measure the heterozygosity first:

```{r stage13-dosage}
dosage <- dgr_hap1 + dgr_hap2
diagnostics <- dosage_diagnostics(dosage)
c(
  heterozygosity = diagnostics$overall_heterozygosity,
  missing = diagnostics$overall_missing
)
```

```{r stage13-convert, eval=FALSE}
converted <- haplotypes_from_inbred_dosage(
  dosage,
  heterozygous_policy = "drop_variant"
)
```

No threshold is imposed on residual heterozygosity, because no universally
appropriate level exists; it is measured and reported instead, and heterozygous
calls are never resolved silently.

## Stage 14 — Calibrate the simulation

Genome structure comes from your markers. Trait architecture is calibrated to
the covariance matrix you already estimated, so the simulation reproduces both.

```{r stage14, eval=FALSE}
setup <- founder_population(
  founders,
  G = dgr_G,
  h2 = stats::setNames(dgr_traits$heritability, traits),
  n_qtl_per_chromosome = 100L,
  seed = 42L
)
setup
```

```
#> <desiredgainr_sim_setup>
#>   Founders: 200 individuals, 3 chromosomes, 100 QTL per chromosome
#>   Traits: GY, PHT, AD, ASI, EPP, GLS
#>   Dominance simulated: no
```

For a **clonal** programme add `dominance_degree`, and supply the *genotypic*
rather than the additive covariance as `G`, because the clone inherits
dominance intact.

## Stage 15 — Simulate one objective across cycles

```{r stage15, eval=FALSE}
simulation <- simulate_selection_cycles(
  setup,
  desired_gains = desired_gains,
  n_cycles = 5L,
  mating_system = "outcross",
  n_parents = 20L, n_crosses = 50L, n_progeny_per_cross = 10L,
  lower_is_better = lower_is_better,
  seed = 11L
)
simulation
```

```
#> <desiredgainr_simulation>
#>   outcross system, 5 cycles, 20 parents recycled
#>   Index re-estimated each cycle: yes
#>   Cumulative genetic gain:
#>       GY     PHT      AD     ASI     EPP     GLS
#>    1.842  -8.031  -1.226  -0.244  -0.019  -0.487
#>   Final mean relationship among parents: 0.0913
```

Output above is illustrative of the object's shape; your numbers depend on your
founders and parameters.

Three mating systems are supported, and they differ in more than naming:

| System | Crops | What differs |
|---|---|---|
| `"self"` | Wheat, rice, bean | Advanced by selfing or doubled haploidy; recombination releases variance slowly |
| `"outcross"` | Maize, sorghum, millet | Random mating each cycle; half the selection-induced disequilibrium decays per generation |
| `"clonal"` | Cassava, sweetpotato, banana | Selection on *total* genetic value, because dominance is inherited intact |

The per-cycle table records the genetic mean and variance for each trait, the
mean relationship among selected parents, and the implied effective population
size, so that gain and the loss of diversity can be read together.

`reestimate_index = TRUE`, the default, rebuilds the index from each cycle's own
data, which is what a programme actually does and which propagates estimation
error across cycles. Setting it `FALSE` isolates the effect of the
desired-gain direction alone. A large divergence between the two means the
recommendation is sensitive to estimation error.

## Stage 16 — Search for the best objective over cycles

Because only the *direction* of $\mathbf{d}$ matters, the search space is the
unit sphere with $p-1$ free parameters for $p$ traits — small enough to cover
densely.

```{r stage16, eval=FALSE}
optimisation <- optimize_desired_gains(
  setup,
  n_cycles = 5L,
  mode = "pareto",
  budget = 60L,
  n_replicates = 3L,
  mating_system = "outcross",
  n_parents = 20L, n_crosses = 50L, n_progeny_per_cross = 10L,
  lower_is_better = lower_is_better,
  checkpoint = "search.rds",
  seed = 42L
)
optimisation
```

```
#> <desiredgainr_optimisation>
#>   Mode: pareto   Cycles: 5   Evaluations: 60 (3 replicates each)
#>   Objectives: GY, PHT, AD, ASI, EPP, GLS, diversity
#>   Non-dominated directions: 9 of 60
#>   No single direction is recommended; choose a point on the frontier.
```

**The simulation is the objective function and is never replaced by a cheaper
approximation.** A Gaussian process decides only where to spend the next
simulation; it never filters or substitutes for one, and because the
acquisition function keeps exploring, no region can be permanently excluded.

Four ranking modes are available:

| Mode | Objective | Needs |
|---|---|---|
| `"pareto"` | Non-dominated outcomes | nothing beyond the setup |
| `"economic"` | Maximise $\mathbf{w}^\mathsf{T}\mathbf{R}$ | `economic_weights` |
| `"target"` | Minimise distance to absolute targets | `target_gains` |
| `"constrained"` | Maximise one trait subject to floors | `focal_trait`, `gain_floors` |

`"pareto"` is the default because it is the only mode that does not require the
weights breeders find hardest to state. Choosing a point on a frontier is an
easier judgement, and it is weight elicitation by revealed preference.

A realistic budget takes hours, so supply `checkpoint` to make the search
resumable.

---

## Sign-off checklist

Complete this before promoting any output to a breeding recommendation.

| Check | Where it came from |
|---|---|
| Covariance matrices positive definite, conditioning inspected | Stage 3 |
| Heritabilities consistent with $\mathbf{G}$ and $\mathbf{P}$ | Stage 4 |
| Trait directions declared, not signed by hand | Stage 2 |
| Objective feasible at the planned intensity, or knowingly not | Stage 7 |
| Index family chosen and justified against a comparator | Stages 9-10 |
| Effective weights inspected; no unintended trait dominance | Stage 10 |
| Replicate diagnostics agree | Stage 11 |
| Traits standardised, or the warning read and accepted | Stages 3, 12 |
| Decision robust to plausible weight perturbation | *Objective vignette* |
| Random seed, package version and session recorded | below |

```{r session}
packageVersion("DesiredGainR")
sessionInfo()
```

---

## Where each stage is covered in depth

| Stage | Vignette |
|---|---|
| 1-4 | *Obtaining the genetic and phenotypic covariance matrices* |
| 5-8 | *Defining a breeding objective* |
| 9-10 | *Choosing an index* |
| 11 | *Optimising desired gains* |
| 12 | *Quadratic genomic selection* |
| 13-16 | *Comparing objectives over several cycles* |

---

## References

- Cerón-Rojas JJ, Montesinos-López OA, Montesinos-López A, et al. (2026).
  Nonlinear genomic selection index accelerates multi-trait crop improvement.
  *Nature Communications* **17**:1991.
- Covarrubias-Pazaran G (2021). *Practical implementation of selection
  indices.* CGIAR Excellence in Breeding.
- Guimarães PHR et al. (2021). Index selection can improve the selection
  efficiency in a rice recurrent selection population. *Euphytica* **217**:95.
- Joukhadar R et al. (2024). Optimising desired gain indices to maximise
  selection response. *Frontiers in Plant Science* **15**:1337388.
- Rahimi M, Debnath S (2023). Estimating optimum and base selection indices.
  *Scientific Reports* **13**:18977.
