---
title: "DesiredGainR: DGSI and QGSI Workflows"
author: "Félicien Akohoue"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DesiredGainR: DGSI and QGSI Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(DesiredGainR)
set.seed(42)
```

# Scope

DesiredGainR provides two distinct multi-trait selection indices. DGSI
calibrates a linear score towards breeder-defined desired responses. QGSI
predicts a quadratic economic merit from genomic estimated breeding values
(GEBVs). The package does not infer genetic values, fit the upstream genomic
model, or treat desired gains as economic weights.

```{r data}
traits <- c("yield", "protein", "disease")
n <- 60
candidates <- data.frame(
  GenoID = paste0("G", seq_len(n)),
  Family = rep(paste0("F", 1:6), each = 10),
  yield = rnorm(n),
  protein = rnorm(n),
  disease = rnorm(n)
)
G <- cov(candidates[traits])
dimnames(G) <- list(traits, traits)
```

# Optimised desired-gain index

`run_dgsi()` requires the genetic covariance matrix `G`. If `P` is omitted,
the function estimates an empirical working covariance from the reference
values and reports that provenance explicitly.

```{r dgsi}
dgsi <- run_dgsi(
  init_data = candidates[c("GenoID", "Family")],
  cand_data = candidates,
  trait_cols = traits,
  dg = c(yield = 0.6, protein = 0.3, disease = 0.4),
  G = G,
  lower_is_better = "disease",
  n_select = 10,
  n_iter = 100,
  n_rep = 5,
  seed = 42
)

dgsi$best_replicate
dgsi$replicate_diagnostics
head(dgsi$ranked_geno)
```

All stochastic replicates run internally. DesiredGainR returns the replicate
with the smallest response objective automatically. The breeder reviews the
objective, plateau, coefficient, rank-correlation, and selected-set agreement
diagnostics but does not choose among runs manually.

In `select_mode = "eligible_top_n"`, thresholds define the eligible pool after
traits have been oriented so that larger is favourable. The index then ranks
eligible candidates. Missing data stop the analysis unless an explicit
complete-case or reference-mean-imputation policy is requested.

# Approximating genetic covariance

The covariance of BLUPs or GEBVs is a covariance of predictions. It is
generally smaller than total genetic covariance because predictions are
shrunken. DesiredGainR does not silently use it as `G`.

`estimate_genetic_covariance()` makes an approximation explicit and returns
its estimand, provenance, assumptions, component matrices, and numerical
diagnostics.

```{r covariance-working}
G_working <- estimate_genetic_covariance(
  values = candidates,
  trait_cols = traits,
  method = "prediction_covariance"
)
G_working
G_working$provenance
```

When full cross-trait prediction-error covariance (PEV) is available for
compatible multivariate BLUPs, the package can apply

\[
\widehat G =
\operatorname{Cov}(\widehat u) + \overline{\operatorname{PEV}}.
\]

```{r covariance-pev}
pev <- diag(c(yield = 0.10, protein = 0.08, disease = 0.12))
dimnames(pev) <- list(traits, traits)

G_approx <- estimate_genetic_covariance(
  values = candidates,
  trait_cols = traits,
  method = "pev_corrected",
  prediction_error_covariance = pev
)
G_approx$G
G_approx$assumptions
```

Pass the complete `G_approx` object as `G` in `run_dgsi()` to retain its
estimation method, estimand, assumptions, and diagnostics in
`dgsi$covariance_provenance$G`. A genuine covariance matrix from an upstream
model can still be passed directly.

A common trait-by-trait PEV matrix or a trait-by-trait-by-genotype PEV array
can be supplied. If only prediction standard errors are available,
`method = "se_diagonal_corrected"` corrects the diagonal but cannot recover
cross-trait prediction-error covariance. `method =
"relationship_adjusted"` accounts for a named genotype relationship matrix
but does not undo prediction shrinkage.

BLUEs, adjusted phenotypic means, and raw phenotypes contain residual or
environmental covariance. Their covariance is not a genetic covariance
estimate unless an upstream genetic model or suitable genetic and replication
structure separates those components.

# Quadratic genomic selection index

For each candidate, QGSI evaluates

\[
I_{qg} = \mathbf w^\mathsf{T}\hat{\boldsymbol\gamma}
+ \hat{\boldsymbol\gamma}^\mathsf{T}
\mathbf W\hat{\boldsymbol\gamma}.
\]

The breeder supplies linear economic weights \(\mathbf w\) and a symmetric
matrix \(\mathbf W\) of squared and cross-product economic weights. The
off-diagonal coefficient multiplying \(\gamma_i\gamma_j\) is \(2W_{ij}\).
Negative diagonal curvature represents stabilising selection around the
chosen origin, whereas positive curvature favours extremes. DesiredGainR does
not infer this economic objective from desired gains or correlations.

```{r qgsi}
W <- matrix(
  c(
     0.10, 0.02, -0.01,
     0.02, 0.05,  0.00,
    -0.01, 0.00, -0.08
  ),
  3,
  dimnames = list(traits, traits)
)

qgsi <- run_qgsi(
  init_data = candidates[c("GenoID", "Family")],
  gebv_data = candidates,
  trait_cols = traits,
  linear_weights = c(yield = 1, protein = 0.5, disease = 0.7),
  W = W,
  lower_is_better = "disease",
  n_select = 10
)

qgsi
head(qgsi$ranked_geno)
```

GEBVs are centred by default because QGSI theory assumes zero means.
`lower_is_better` changes the trait direction before scoring. Economic weights
must describe the resulting favourable-direction trait space. When scaling is
requested, the weights must also refer to standard-deviation units.

## Estimating genomic covariance

Without a supplied `Gamma`, the function estimates the genomic covariance
matrix from reference GEBVs with Supplementary Equation 19.2 of Cerón-Rojas
et al. (2026):

\[
\hat{\Gamma} = g^{-1}\hat{\gamma}^{\mathsf T}\hat{\gamma}.
\]

Candidate GEBVs serve as the reference when `reference_gebv_data` is absent.
A separate reference population may be supplied. With a named genomic
relationship matrix \(\Phi\), the function instead evaluates Supplementary
Equation 19.1:

\[
\hat{\Gamma} =
g^{-1}\hat{\gamma}^{\mathsf T}\Phi^{-1}\hat{\gamma}.
\]

A spectral Moore–Penrose inverse is used and reported for a rank-deficient
positive-semidefinite relationship matrix.

```{r qgsi-covariance}
qgsi$Gamma
qgsi$covariance_provenance
```

## Model-based parameters

DesiredGainR reports the model expected index, its linear and quadratic
variance components, normal-selection intensity, expected quadratic
net-merit response, and expected per-trait gains from Supplementary Equation
16.

```{r qgsi-theory}
qgsi$theoretical_parameters
qgsi$expected_gain_per_trait
```

Squared index–merit correlation and mean squared prediction error require
`true_G`. That matrix is generally known only in simulation. These quantities
remain unavailable for empirical data instead of being calculated under the
unverifiable assumption that estimated `Gamma` is the true covariance.

The selected candidates' mean GEBV shift is returned separately:

```{r qgsi-observed}
qgsi$observed_selection_differential
```

This is an observed differential among supplied predictions, not realised
genetic gain. Realised gain requires later-cycle observations.

Candidate-specific contribution tables reproduce the linear and quadratic
score:

```{r qgsi-contributions}
head(qgsi$linear_contributions)
head(qgsi$quadratic_contributions)
```

They explain each candidate's score. A nonlinear QGSI contribution is not one
global additive marker-effect vector.

# Comparing the methods

The combined pipeline requires separate objectives:

```{r both}
both <- run_dgsi_qgsi_pipeline(
  mode = "both",
  init_data = candidates[c("GenoID", "Family")],
  trait_cols = traits,
  dg = c(yield = 0.6, protein = 0.3, disease = 0.4),
  cand_data = candidates,
  G = G,
  n_select = 10,
  n_iter = 50,
  n_rep = 3,
  gebv_data = candidates,
  qgsi_linear_weights = c(yield = 1, protein = 0.5, disease = 0.7),
  W = W,
  qgsi_n_select = 10,
  lower_is_better = "disease",
  debug = FALSE
)

head(both$comparison_result$comparison_table)
```

Rank agreement is descriptive. DGSI targets desired responses; QGSI predicts
a quadratic economic merit. Agreement therefore does not validate either
method.

# Interpretation and reporting

Report the upstream genetic model, target population and environments,
covariance provenance, trait directions, transformations, selection
intensity, DGSI desired responses, QGSI economic weights, random seed,
optimisation diagnostics, and independent later-cycle evaluation.

The algorithms calculate results from the supplied genetic values and
parameters. Their breeding relevance is driven by whether those inputs
represent the target programme and whether the breeder's objectives are
scientifically and economically defensible.

```{r session}
packageVersion("DesiredGainR")
sessionInfo()
```
