---
title: "Defining a breeding objective"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Defining a breeding objective}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(DesiredGainR)
data(dgr_traits, package = "DesiredGainR")
data(dgr_G, package = "DesiredGainR")
data(dgr_P, package = "DesiredGainR")
data(dgr_candidates, package = "DesiredGainR")
data(dgr_history, package = "DesiredGainR")
traits <- dgr_traits$trait
lower_is_better <- dgr_traits$trait[dgr_traits$direction == "decrease"]
```

## 1. Why this layer exists

An index answers the question "given this objective, which candidates?".
Nothing in index theory answers "what should the objective be?", and that is
the question breeders find hard.

Guimarães et al. (2021) quantified the cost of getting it wrong. Working in a
lowland rice recurrent-selection population, they applied Smith-Hazel and Tai
indices under three weight schemes: genetic standard deviations, a
variation-based index, and an arbitrary vector. Under the arbitrary weights the
Smith-Hazel index produced *no* gain for grain yield, leaf scald or grain
discoloration, and unfavourable change in three of the six traits. The index
machinery was identical in all three runs. Only the objective differed.

Covarrubias-Pazaran (2021) reaches the same place from the operational side.
The CGIAR guideline instructs breeders to state a desired differential and then
adds, in the same paragraph, that the resulting coefficients should not be
interpreted, because they lack meaning under strong genetic correlations. The
desired response is the only decision genuinely available.

This vignette covers the six tools DesiredGainR provides for that decision.

---

## 2. The duality: two ways of saying one thing

### 2.1 The result

The Smith-Hazel index sets `b = P⁻¹Gw` from economic weights `w`. The
desired-gain index sets `b = P⁻¹G(GP⁻¹G)⁻¹d` from desired gains `d`. Setting
the two equal gives

$$w = G^{-1}PG^{-1}d, \qquad d = GP^{-1}Gw.$$

So they are not competing methods. They are two coordinate systems on one
index, and every objective stated in either can be read in the other.

### 2.2 Why it matters practically

A breeder who cannot put a currency value on grey leaf spot resistance can
usually say how much of it they want relative to yield. The duality turns that
into weights.

The reverse direction is the more useful one in practice, and is
under-exploited. When a programme proposes weights, showing them the response
those weights actually request often reveals that the objective is not what
anyone intended.

```{r duality}
desired_gains <- c(
  GY = 1.0, PHT = 0.4, AD = 0.6, ASI = 0.5,
  EPP = 0.4, GLS = 0.6
)

implied <- implied_economic_weights(
  desired_gains, dgr_G, dgr_P,
  lower_is_better = lower_is_better, gain_units = "genetic_sd"
)
round(implied, 3)
```

The translation is exactly invertible when the units match, and checking that
is worth the one line it costs:

```{r invert}
round(
  implied_desired_gains(
    implied, dgr_G, dgr_P,
    lower_is_better = lower_is_better, gain_units = "genetic_sd"
  ),
  3
)
```

### 2.3 Two ways to misread the output

Both of the following mislead readers who have not met them before, and both
appear in the vector above.

**Negative weights arise for traits the objective asks to improve.** Here the
anthesis-silking interval and ears per plant both receive negative weights,
although the objective requests gains in both. This is correct. Once oriented,
grain yield correlates +0.55 with the anthesis-silking interval and +0.45 with
ears per plant, so selecting for a full standard deviation of yield would carry
both *past* the smaller gains requested for them. Delivering the stated ratio
therefore requires the index to hold them back.

A negative weight never means the trait should decrease. `lower_is_better` has
already applied direction, so larger is better throughout.

**Magnitudes are not comparable across traits.** A weight carries inverse trait
units, so a trait on a small scale receives a large number for the same
emphasis. Rescale before comparing:

```{r rescale}
genetic_sd <- stats::setNames(dgr_traits$genetic_sd, dgr_traits$trait)
round(sort(implied * genetic_sd[names(implied)], decreasing = TRUE), 2)
```

Ears per plant carried the largest raw number in the vector and is one of the
smaller effects. Reading the unscaled vector would send a breeder to argue
about the wrong trait.

---

## 3. Feasibility: can the objective be attained at all?

### 3.1 The achievable-response ellipsoid

For any linear index the response vector is
`R = i·Gb/√(bᵀPb)`. Writing `u = b/√(bᵀPb)` so that `uᵀPu = 1`, the attainable
responses are `{i·Gu}`, which is the ellipsoid

$$R^\mathsf{T}G^{-1}PG^{-1}R = i^{2}.$$

A stated target `d` therefore requires exactly

$$i_{\mathrm{required}} = \sqrt{d^\mathsf{T}G^{-1}PG^{-1}d},$$

and the selected proportion delivering that intensity follows from the
normal-truncation relationship.

### 3.2 Two consequences

**The classical desired-gain index honours only the direction of `d`.**
Multiplying every desired gain by a constant leaves the index unchanged,
because the attainable magnitude is fixed by selection intensity. Joukhadar et
al. (2024) note this in passing when their INDEX1 and INDEX3, targeting 0.5 and
4.0 standard deviations respectively, produced identical rankings without
iteration. Breeders who set absolute targets and then read back a realised gain
are being misled unless the point is made explicit.

**An antagonistic correlation structure can make a modest-looking target
unreachable.** This is not a marginal effect.

```{r feasibility}
feasibility <- gain_feasibility(
  desired_gains, dgr_G, dgr_P,
  n_candidates = nrow(dgr_candidates), n_select = 20L,
  lower_is_better = lower_is_better, gain_units = "genetic_sd"
)
feasibility
```

The request looks unremarkable — between 0.4 and 1.0 genetic standard
deviations across six traits — and requires selecting fewer than one candidate
from two hundred.

```{r attainable}
round(feasibility$attainable_response_input_units, 3)
```

Fifty-five per cent of the request, in the requested proportion, is what the
planned intensity delivers.

### 3.3 Where this check belongs

The CGIAR guideline directs breeders to separate software to determine whether
"the progress of 1 or more σ is possible given your current trait genetic
correlations and selection intensity". That check belongs in the same place as
the index, and it belongs *before* the optimisation rather than after it
disappoints.

---

## 4. Recovering an objective from past decisions

### 4.1 The method

Where a programme has selected for years without a formal index, the
differentials it achieved already encode its objective. Covarrubias-Pazaran
recovers the corresponding linear rule as

$$b = P^{-1}s,$$

where `s` holds the differentials between the selected group and the full
population. This is how an index was introduced at both the International Maize
and Wheat Improvement Center (CIMMYT) and the International Rice Research
Institute (IRRI).

```{r retrospective}
recovered <- retrospective_weights(
  selected_values = dgr_candidates[dgr_history$selected, traits],
  population_values = dgr_candidates[, traits],
  trait_cols = traits
)
recovered
```

### 4.2 Checking the recovery

The example data record the weights that generated the historical decision, so
the recovery can be verified rather than asserted. Comparing on a common scale,
oriented and normalised to grain yield:

```{r verify}
direction <- ifelse(dgr_traits$direction == "increase", 1, -1)
names(direction) <- traits
oriented <- recovered$coefficients_per_sd * direction
generating <- attr(dgr_history, "generating_weights")

data.frame(
  trait = traits,
  recovered = round(oriented / oriented[["GY"]], 2),
  generating = round(generating / generating[["GY"]], 2)
)
```

The recovery captures the shape of the objective without reproducing it
exactly, which is the honest result. `b = P⁻¹s` returns the linear rule that
best reproduces the *observed differentials*, and those differentials are
themselves one realisation of a selection decision in a finite population.

### 4.3 What it is for, and what it is not for

A retrospective weight vector answers "what linear rule best approximates the
decisions already made?". It does not answer "what should this programme select
for next?".

The distinction is operational, not pedantic. The IRRI deployment reports
retrospective coefficients for four regional programmes in which the *same*
resistance-gene indicator carries opposite signs, and the fine-tuning step that
followed — doubling the Philippines yield coefficient — raised the reported
final gain from 2.230 to 3.772. Recovered weights reproduce past behaviour,
including any bias in it. Treat them as a starting point, inspect them, and
adjust deliberately against the target product profile.

---

## 5. How much does the objective matter?

### 5.1 The question

Weights are estimates, and refining them costs meetings. Before spending that
effort it is worth knowing whether the decision would change.

```{r sensitivity}
economic_weights <- c(
  GY = 1.0, PHT = 0.2, AD = 0.5,
  ASI = 0.4, EPP = 0.3, GLS = 0.5
)
sensitivity <- weight_sensitivity(
  economic_weights, dgr_candidates, dgr_G, dgr_P,
  n_select = 20L, trait_cols = traits, n_draws = 100L
)
sensitivity
```

Weights are perturbed multiplicatively on the log scale, the index is rebuilt
for each draw, and the resulting selected sets are compared with the set
obtained from the stated objective.

### 5.2 Reading the result

| Observation | Interpretation |
|---|---|
| High stability, high rank correlation | The decision is robust; further argument about weights will not change it |
| Low stability, high rank correlation | The ranking is stable but the threshold is contested; consider selecting more candidates |
| Low stability, low rank correlation | The objective, not the index, is the binding uncertainty |

### 5.3 Influence is not contribution

`weight_influence` measures how strongly perturbing each weight disturbs the
selected set. It is a different quantity from the share of the index a trait
contributes, and the two routinely disagree.

In the output above, plant height is the weight the decision is most sensitive
to, while contributing about six per cent of the index:

```{r effective}
smith_hazel <- selection_index(
  dgr_candidates, traits,
  method = "smith_hazel",
  G = dgr_G, P = dgr_P, economic_weights = economic_weights,
  lower_is_better = lower_is_better, n_select = 20L, main_trait = "GY"
)
smith_hazel$effective_weights[, c("Trait", "Genetic_share")]
```

There is no contradiction. A trait carrying little of the index can still be
the largest lever on the decision, because it moves the index in a direction
the remaining traits do not cover, and therefore reorders the candidates near
the selection threshold. A trait that dominates the index may conversely be a
weak lever, because the ranking already follows it.

Read the two together. The first says which weights are worth arguing about;
the second says which traits the index is acting on.

---

## 6. The Crosbie trap

Crosbie et al. (1980), as reported by Guimarães et al. (2021), observed that
assigning equal weights across unstandardised traits places most of the
selection pressure on whichever trait carries the largest genetic variance.
This is the single most common way for an objective to be quietly subverted,
and the CGIAR guideline's recommendation to start from a desired differential
of one for every trait makes it easy to walk into.

The example programme is built to show it. Trait standard deviations span a
factor of about one hundred and twenty:

```{r crosbie}
round(sort(genetic_sd, decreasing = TRUE), 2)
```

The consequence is visible in the conditioning of the covariance matrices. Most
of it comes from the scales rather than from the correlations:

```{r conditioning}
c(
  covariance = matrix_diagnostics(dgr_G)$condition_number,
  correlation = matrix_diagnostics(stats::cov2cor(dgr_G))$condition_number
)
```

Standardising removes three orders of magnitude of ill-conditioning.
`run_dgsi()` and `run_qgsi()` therefore warn when trait standard deviations
differ by more than fivefold and standardisation has been switched off, naming
the trait that dominates and its share.

---

## 7. Summary of the tools

| Function | Question it answers |
|---|---|
| `implied_economic_weights()` | What weights do these desired gains imply? |
| `implied_desired_gains()` | What response do these weights actually request? |
| `gain_feasibility()` | Is this objective attainable, and at what intensity? |
| `retrospective_weights()` | What has this programme been selecting for? |
| `weight_sensitivity()` | Would a different objective change the decision? |
| `effective_weights()` | Which traits is the index really acting on? |
| `matrix_diagnostics()` | Can these covariance matrices be inverted safely? |

---

## 8. What is not here

Two routes to an objective are outside the package's current scope, and both
are legitimate.

**A profit function.** The only principled definition of an economic weight is
the derivative of profit with respect to the trait mean. Where a programme can
write down prices, discounts and costs, differentiating that function is better
than any elicitation. DesiredGainR does not yet provide a helper for it.

**Structured elicitation.** Marginal rates of substitution are weight ratios
directly, so a short sequence of trade-off questions pins the vector up to
scale, with a consistency check. This is standard decision analysis and would
fit the package well.

Both are recorded as intended additions rather than omissions of principle.

---

## 9. References

- Covarrubias-Pazaran G (2021). *Practical implementation of selection
  indices.* CGIAR Excellence in Breeding.
- Crosbie TM, Mock JJ, Smith OS (1980), as discussed in Guimarães et al.
  (2021).
- Guimarães PHR, Melo PGS, Cordeiro ACC, Torga PP, Rangel PHN, de Castro AP
  (2021). Index selection can improve the selection efficiency in a rice
  recurrent selection population. *Euphytica* **217**:95.
  <https://doi.org/10.1007/s10681-021-02819-7>
- Joukhadar R, Li Y, Thistlethwaite R, Forrest KL, Tibbits JF, Trethowan R,
  Hayden MJ (2024). Optimising desired gain indices to maximise selection
  response. *Frontiers in Plant Science* **15**:1337388.
  <https://doi.org/10.3389/fpls.2024.1337388>
- Pesek J, Baker RJ (1969). Desired improvement in relation to selection
  indices. *Canadian Journal of Plant Science* **49**:803-804.
  <https://doi.org/10.4141/cjps69-137>
- Smith HF (1936). A discriminant function for plant selection. *Annals of
  Eugenics* **7**:240-250.
