---
title: "Choosing an index"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing an index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
library(DesiredGainR)
data(dgr_traits, package = "DesiredGainR")
data(dgr_G, package = "DesiredGainR")
data(dgr_P, package = "DesiredGainR")
data(dgr_candidates, package = "DesiredGainR")
traits <- dgr_traits$trait
lower_is_better <- dgr_traits$trait[dgr_traits$direction == "decrease"]
```

## 1. Begin with the objective, not the method

An index family is appropriate only relative to a stated breeding objective.
Economic value, a desired response, hard eligibility limits and non-linear
value judgements are different objectives. They should not be made to look
equivalent by applying the same function name to all of them.

| Breeder's statement | Appropriate starting point |
|---|---|
| “I can defend the relative economic value of each trait.” | Smith–Hazel |
| “The measurements themselves already form the score.” | Base index |
| “I can defend a desired response direction.” | Pesek–Baker or Yamada |
| “I know acceptable gain intervals, but not one exact vector.” | Population-driven desired-gain suggestions |
| “Every selected candidate must pass these limits.” | Independent culling or a restricted index |
| “I need a simple, weight-light comparator.” | Mulamba–Mock rank sum |
| “Trait value is genuinely curved or interactive.” | Quadratic genomic selection index |

Tandem selection is included as a comparator, but it postpones the other
traits rather than balancing them in the same decision.

---

## 2. Fit comparable linear families

All traits are oriented so that larger values are favourable. The objective
below is expressed in genetic standard-deviation units.

```{r objectives}
desired_gains <- c(
  GY = 1.0, PHT = 0.4, AD = 0.6,
  ASI = 0.5, EPP = 0.4, GLS = 0.6
)
economic_weights <- c(
  GY = 1.0, PHT = 0.2, AD = 0.5,
  ASI = 0.4, EPP = 0.3, GLS = 0.5
)
```

The two vectors are separate statements. `desired_gains` defines a response
direction, whereas `economic_weights` defines one common aggregate merit for
the merit-based comparison below.

```{r fit-families}
smith_hazel <- selection_index(
  dgr_candidates, traits,
  method = "smith_hazel", G = dgr_G, P = dgr_P,
  economic_weights = economic_weights,
  lower_is_better = lower_is_better,
  scale_traits = TRUE, scale_by = "phenotypic",
  n_select = 20L, main_trait = "GY"
)

pesek_baker <- selection_index(
  dgr_candidates, traits,
  method = "pesek_baker", G = dgr_G, P = dgr_P,
  desired_gains = desired_gains,
  aggregate_weights = economic_weights,
  lower_is_better = lower_is_better,
  scale_traits = TRUE, scale_by = "phenotypic",
  n_select = 20L, main_trait = "GY"
)

base_index <- selection_index(
  dgr_candidates, traits,
  method = "base", G = dgr_G, P = dgr_P,
  economic_weights = economic_weights,
  lower_is_better = lower_is_better,
  scale_traits = TRUE, scale_by = "phenotypic",
  n_select = 20L, main_trait = "GY"
)

rank_sum <- selection_index(
  dgr_candidates, traits,
  method = "mulamba_mock",
  G = dgr_G, P = dgr_P,
  lower_is_better = lower_is_better,
  scale_traits = TRUE, scale_by = "phenotypic",
  n_select = 20L, main_trait = "GY"
)
```

Smith–Hazel and Pesek–Baker are compared here under one common definition of
aggregate merit. Without that common definition, their index–merit correlation
and aggregate-merit response would not answer the same question.

---

## 3. Compare biological responses before summary criteria

```{r response-comparison}
response_table <- data.frame(
  Trait = traits,
  Smith_Hazel = unname(smith_hazel$evaluation$expected_response[traits]),
  Pesek_Baker = unname(pesek_baker$evaluation$expected_response[traits]),
  Base = unname(base_index$evaluation$expected_response[traits])
)
response_table[-1] <- round(response_table[-1], 3)
response_table
```

This table is the primary comparison. It shows what is expected to change in
every trait. A high aggregate statistic cannot rescue an index that moves a
critical trait in the wrong direction or misses its minimum requirement.

Then inspect the summary criteria.

```{r criteria}
criteria <- function(x) {
  with(x$evaluation, c(
    index_merit_correlation = R_HI,
    aggregate_merit_response = delta_H,
    relative_efficiency_GY = RE,
    index_heritability = h2_index
  ))
}
round(rbind(
  Smith_Hazel = criteria(smith_hazel),
  Pesek_Baker = criteria(pesek_baker),
  Base = criteria(base_index)
), 3)
```

Interpret these criteria with their limits: the index–merit correlation and
aggregate-merit response require one common merit definition; relative
efficiency describes only the declared main trait; index heritability describes
repeatability of the score, not attainment of the complete breeding objective.

---

## 4. Compare the actual decision

Two indices can have similar expected responses while selecting different
parents. Compare ranks and selected-set overlap, then repeat under covariance
or weight uncertainty.

```{r selected-overlap}
selected_ids <- function(x) {
  x$ranking$id[x$ranking$selected]
}

sets <- list(
  Smith_Hazel = selected_ids(smith_hazel),
  Pesek_Baker = selected_ids(pesek_baker),
  Base = selected_ids(base_index),
  Rank_sum = selected_ids(rank_sum)
)

length(Reduce(intersect, sets))
```

Use [`index_uncertainty()`](../reference/index_uncertainty.html) for covariance
resampling and [`weight_sensitivity()`](../reference/weight_sensitivity.html)
when economic weights are uncertain. For minimum gain floors, inspect the
worst-trait and joint-attainment probabilities supplied by the population-driven
tools.

---

## 5. Where DGSI and QGSI belong

The iterative desired-gain selection index (DGSI) searches for a selected-group
differential close to a desired vector. Read [Optimising desired gains](DesiredGainR-dgsi.html).

The quadratic genomic selection index (QGSI) ranks GEBVs under explicit linear,
squared and cross-product economic values. It does not accept desired gains as
quadratic weights. Read [Quadratic genomic selection](DesiredGainR-qgsi.html).

Choose the simplest family that represents the programme's real objective and
changes the decision in a stable, biologically defensible way.
