---
title: "Defining desired gains when one exact value is not known"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Defining desired gains when one exact value is not known}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(DesiredGainR)
```

## 1. What a desired gain means

A desired gain is the genetic response pattern that the breeding programme
wants the index to pursue. It is not an economic weight, an index coefficient,
or a guarantee about the response that selection will deliver.

For example, a breeder might describe the objective as:

* Increase yield by roughly 0.5 to 1.5 genetic standard deviations.
* Reduce disease severity by roughly 0.25 to 1.0 genetic standard deviations.
* improve quality by 0.1 to 0.8 genetic standard deviations.

The relative sizes describe the requested balance among traits. The signs
describe the direction of change in the original trait scale. A reduction in
disease severity is therefore negative before the trait is oriented.

The classical desired-gain index uses only the **direction** of this vector.
The vectors `(1.5, -0.5, 0.4)` and `(3, -1, 0.8)` produce the same index because
their ratios are identical. Actual response magnitude is determined by the
available genetic variation, correlations, accuracy and selection intensity.

## 2. Why genetic standard deviations are a useful language

Original units are easy to discuss within one trait but misleading across
traits. One tonne per hectare of yield and one disease-score unit do not
represent comparable amounts of selectable variation.

A genetic standard deviation (genetic SD) expresses change relative to the
breeder's estimate of additive genetic variation in the target population:

$$d_{j,\mathrm{SD}} = d_j / \sqrt{G_{jj}}.$$

Thus `+1.0 genetic SD` means one estimated genetic standard deviation of
improvement. For a lower-is-better trait, `-0.5 genetic SD` in the raw scale
means a favourable reduction of half a genetic SD. In a simulation setup this
scale is `sqrt(diag(setup$G_target))`. The random finite-QTL covariance stored
as `G_realised` is a calibration diagnostic, never a replacement for the
breeder's estimand. Otherwise, the same threshold would change with a simulator
seed.

Genetic SD is recommended for eliciting multi-trait intervals. It does not make
`1.5 SD` attainable. Feasibility must be checked against the covariance matrix
and planned selection intensity.

## 3. When the breeder cannot propose a desired-gain vector

Sometimes even an interval is more precision than the breeder can defend. The
breeder may know only the smallest useful response for each trait. Each trait
can have its own minimum:

```{r population-suggestion, eval=FALSE}
suggestion <- suggest_desired_gains(
  setup,
  minimum_gains = c(Yield = 1.0, Disease = 0.5, Quality = 0.25),
  lower_is_better = "Disease",
  programme = list(
    mating_system = "outcross",
    n_parents = 20,
    n_crosses = 50,
    n_progeny_per_cross = 10
  )
)
```

This asks for at least one estimated genetic SD of yield improvement, half an SD
of disease reduction, and one-quarter SD of quality improvement. The numbers
are favourable magnitudes, so the disease minimum is positive even though its
response in original disease units should be negative.

The result deliberately gives two answers:

* `minimum_recommendation` is the direction with the strongest conservative
  evidence of reaching **all three trait-specific minima**.
* `maximum_balanced_recommendation` is the direction with the highest robust
  lower bound on the worst-responding trait, after every trait is expressed in
  genetic SD.

The second answer needs a definition of “highest.” There is no single largest
multi-trait vector without economic weights or another preference rule: one
vector may provide more yield and another more resistance.
`suggest_desired_gains()` uses maximin because it treats the traits
symmetrically and prevents exceptional gain in one trait from concealing a weak
response in another.

The mathematical diagnostic first uses the achievable-response ellipsoid

$$R^\mathsf{T}G^{-1}PG^{-1}R=i^2$$

and solves an exact convex quadratic programme subject to the trait-specific
lower bounds. It reports whether all minima can be reached in one cycle, the
maximum common gain possible at the planned selection intensity, the limiting
traits, and the numerical KKT residual verifying the global solution.

Forward simulation then handles processes that the one-cycle ellipsoid cannot:
recombination, finite-family sampling, drift, variance erosion and repeated
selection. The adaptive search is used only for discovery. All discovered
directions are locked and rerun with new common random numbers before either
recommendation is chosen. The package ranks minimum attainment using an exact
multiplicity-adjusted Clopper--Pearson lower confidence bound, and ranks the
balanced solution using a simultaneous distribution-free lower confidence
bound for the median worst-trait gain. Several finalists for each objective are
then run a third time with independent seeds. The final vector is chosen only
from this confirmation stream. Exact simultaneous lower and upper probability
bounds classify it as `supported`, `uncertain`, or `not_supported` relative to
the declared required success probability. An uncertain best-available vector
is not printed as a recommendation.

Three further checks can prevent a recommendation: multi-start/half-budget
search disagreement, inadequate reproduction of `G_target` by the simulated
finite-QTL architecture, and failure across newly sampled architectures.
Covariance-estimation uncertainty is also propagated when defensible genetic
and residual degrees of freedom are supplied. Without them, any supported
answer is labelled `supported_conditional_on_covariance`. Simulation cannot
manufacture uncertainty information that the data analysis did not provide.

Read the result as follows:

```{r population-results, eval=FALSE}
suggestion$analytical_feasibility
suggestion$minimum_recommendation
suggestion$maximum_balanced_recommendation
suggestion$confirmation
suggestion$model_diagnostic
suggestion$search_stability
suggestion$outer_validation
```

The returned `desired_gain_direction` is a direction in favourable genetic-SD
space. Its length is fixed to one because multiplying the entire vector by a
constant does not change a classical desired-gain index. The confirmation table
reports expected realised responses and probabilities. Do not interpret the
direction entries themselves as guaranteed gains.

Defaults use three independent searches, 50 discovery replicates, 50
independent screening replicates, 200 independent confirmation replicates, and
30 newly sampled finite-QTL architectures. These are intentionally substantial
because the function may make a recommendation. Reduce them only for code
testing, not to support a breeding decision. To propagate sampling uncertainty
in covariance estimates, add for example:

```{r covariance-control, eval=FALSE}
control <- list(
  uncertainty = list(
    covariance_draws = 50,
    genetic_df = 120, # independent genetic units, not plot count
    residual_df = 400
  )
)
```

## 4. Ask for intervals, not false precision

Use `define_desired_gain_intervals()` to record what the breeder can defend:

```{r intervals}
gain_intervals <- define_desired_gain_intervals(
  lower = c(Yield = 0.5, Disease = -1.0, Quality = 0.1),
  upper = c(Yield = 1.5, Disease = -0.25, Quality = 0.8),
  lower_is_better = "Disease",
  gain_units = "genetic_sd",
  horizon_cycles = 5
)
gain_intervals
```

The input uses familiar raw changes, so disease reductions are negative. The
printed object uses the favourable-direction convention internally: every
positive value means improvement. This prevents a sign error later when the
phenotypes and covariance matrices are oriented.

The questions behind each interval should be recorded:

1. What is the smallest change that would matter to the programme?
2. What is an ambitious but still biologically credible change?
3. Is deterioration ever acceptable, and if so how much?
4. Is the interval stated per cycle or over the full planning horizon?
5. Does the genetic SD come from the current target population?

## 5. The interval-mode definition of “best”

The interval box tells DesiredGainR which desired-gain ratios the breeder is
willing to consider. In `mode = "interval"`, its lower bounds also define the
high-gain event at the declared planning horizon.

Let $R_j$ be the favourable cumulative response for trait $j$, expressed in
the interval's units, and let $L_j$ be its lower bound. One simulation replicate
is a joint success when

$$R_j \ge L_j \quad \text{for every trait } j.$$

This definition is deliberately strict. Exceptional yield cannot compensate
for disease response below the breeder's minimum.

If the joint event occurs in $s$ of $n$ independent simulation replicates, the
reported probability uses the Jeffreys binomial posterior

$$p \mid s,n \sim \operatorname{Beta}(s+1/2,n-s+1/2).$$

The package reports its posterior mean and credible interval. The recommended
vector maximises the **lower credible bound**, not the raw success proportion.
This conservative rule penalises poorly replicated apparent winners.

When probabilities tie, a balanced achievement score compares how far each
trait progressed through its interval:

$$A_j = \frac{R_j-L_j}{U_j-L_j},$$

$$S = \min_j A_j + 0.05\,\operatorname{mean}_j\{\min(A_j,1)\}.$$

The worst-satisfied trait is primary. The small bounded average term breaks
ties without allowing one exceptional trait to conceal another trait's
failure. This is an augmented achievement-scalarising rule, not an implicit
economic weighting scheme.

Other modes remain available when a different decision is intended:

`optimize_desired_gains()` offers four routes that do not require economic
weights:

| Breeder's decision | Recommended mode |
|---|---|
| Recommend the vector most reliably reaching every interval minimum | `"interval"` |
| Show the defensible trade-offs and do not force one answer | `"pareto"` |
| Come closest to operational absolute targets | `"target"` |
| Maximise one focal trait while protecting the others | `"constrained"` |

The economic mode is optional and is appropriate only when credible economic
weights exist.

## 6. Search only within the breeder's intervals

After constructing a simulation setup from the programme's founder population,
the probability-based interval search is:

```{r optimise, eval=FALSE}
optimisation <- optimize_desired_gains(
  setup,
  desired_gain_intervals = gain_intervals,
  mode = "interval",
  n_cycles = 5,
  budget = 80,
  n_replicates = 50,
  lower_is_better = "Disease",
  mating_system = "outcross",
  n_parents = 20,
  n_crosses = 50,
  n_progeny_per_cross = 10,
  seed = 42
)
```

The optimiser samples only desired-gain directions whose ray intersects the
specified interval box. `optimisation$desired_gains` contains a representative
request within the interval for every evaluated direction. Columns beginning
with `requested_` in `optimisation$results` report the same values.

Read the probability-based recommendation directly:

```{r interval-results, eval=FALSE}
optimisation$recommended_desired_gains
optimisation$recommended_probability
optimisation$recommendations[, c(
  "requested_Yield", "requested_Disease", "requested_Quality",
  "joint_high_gain_probability", "joint_probability_lower",
  "joint_probability_upper", "bootstrap_selection_frequency"
)]
```

Every row is an actually simulated desired-gain vector. The Gaussian-process
surrogate decides only where to spend the next simulation. An unsimulated
surrogate optimum is never recommended. Common random numbers make comparisons
more precise, and the paired bootstrap used for
`bootstrap_selection_frequency` preserves
the shared-seed dependence among vectors. It is a stability frequency under
resampling, not a probability that a vector is the unknown true best.

Twenty replicates is a screening minimum, not a guarantee of precision. Use at
least 50 when recommendations are close, and increase the count until credible
intervals and rankings are stable.

With `mode = "pareto"`, inspect `optimisation$pareto_set` instead. There is then
deliberately no single recommendation because one non-dominated solution can
provide more yield while another provides more disease resistance or retained
diversity.

If the programme has an operational target, use it to select within the
interval-constrained search:

```{r target, eval=FALSE}
optimisation <- optimize_desired_gains(
  setup,
  desired_gain_intervals = gain_intervals,
  mode = "target",
  # These are cumulative responses in original trait units after five cycles.
  target_gains = c(Yield = 1.2, Disease = -0.7, Quality = 0.5),
  n_cycles = 5,
  lower_is_better = "Disease",
  ...
)

optimisation$recommended_desired_gains
optimisation$stability_proportion
```

The first output is the desired-gain request selected from the breeder's
intervals. The stability proportion shows how much of the searched region is
practically indistinguishable from the numerical optimum. A broad stable region
is often more useful than a highly precise-looking point.

For covariance-uncertainty propagation, pass the complete `optimisation`
object rather than extracting its normalized search directions. The function
will use the corresponding original-trait-unit directions automatically:

```{r covariance-propagation, eval=FALSE}
propagate_covariance_uncertainty(
  setup,
  directions = optimisation,
  genetic_df = 80,
  P = P,
  residual_df = 300
)
```

## 7. Check feasibility separately

An interval can contain requests that are biologically impossible at the
planned selection intensity. Check representative points before spending a
large simulation budget. For example, the interval midpoint in favourable
genetic-SD units is:

```{r midpoint}
midpoint <- setNames(
  rowMeans(gain_intervals[c("lower", "upper")]),
  gain_intervals$trait
)
midpoint
```

With the programme covariance matrices, assess it using:

```{r feasibility, eval=FALSE}
gain_feasibility(
  midpoint,
  G,
  P,
  n_candidates = 500,
  n_select = 50,
  lower_is_better = "Disease",
  gain_units = "genetic_sd"
)
```

If the midpoint is infeasible, the breeder can revise the intervals or accept
that the desired proportions will be scaled down. Simulation cannot create
genetic variation that the population does not contain.

## 8. What the probability does and does not mean

The reported probability integrates the segregation, recombination, mating,
selection and phenotype-generation randomness represented by the simulation.
It is conditional on the supplied model. It is not the probability that an
unknown real breeding programme will succeed under every possible model.

Covariance-estimation uncertainty is separate. Pass the optimisation object to
`propagate_covariance_uncertainty()` and report whether its recommendation
remains competitive across plausible covariance matrices.

No finite simulation makes a biological conclusion irrefutable. What makes the
analysis defensible is that the event, probability model, assumptions, random
seeds, uncertainty interval and decision rule are all explicit and testable.

## 9. What “optimum” means in DesiredGainR

The selected solution is conditional on:

* The breeder-defined intervals and ranking mode.
* The founder germplasm and covariance matrices.
* Selection intensity, mating system and population size.
* The number of simulated cycles.
* the assumed genetic architecture and estimation error.

DesiredGainR therefore helps the breeder replace an unjustifiably precise
statement such as “the correct desired gain is exactly `+1.5 SD`” with the more
defensible statement:

> We considered the desired-gain ratios the breeding team judged acceptable,
> simulated their multi-cycle consequences under the planned programme, and
> selected a stable solution using a decision rule stated in advance.

That is the intended role of interval-based desired-gain optimisation.

## References

Brown LD, Cai TT, DasGupta A (2001). Interval estimation for a binomial
proportion. *Statistical Science* **16**:101–133.
doi:10.1214/ss/1009213286.

Wierzbicki AP (1982). A mathematical basis for satisficing decision making.
*Mathematical Modelling* **3**:391–405.
doi:10.1016/0270-0255(82)90038-0.

Yang W-N, Nelson BL (1991). Using common random numbers and control variates in
multiple-comparison procedures. *Operations Research* **39**:583–591.
doi:10.1287/opre.39.4.583.
