---
title: "Multiple-trait selection: methods, choice, and comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple-trait selection: methods, choice, and comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(DesiredGainR)
data(dgr_traits, package = "DesiredGainR")
data(dgr_G, package = "DesiredGainR")
data(dgr_P, package = "DesiredGainR")
data(dgr_candidates, package = "DesiredGainR")
traits <- dgr_traits$trait
lower_is_better <- dgr_traits$trait[dgr_traits$direction == "decrease"]
```

## 1. Learning objectives

Plant breeding programmes usually improve several traits at the same time.
Yield, quality, adaptation, maturity, and disease resistance may all influence
the release decision. Progress in one trait can also reduce progress in
another. Therefore, the method must represent both the breeding objective and
the genetic relationships among traits.

This vignette has four objectives. They are to: (i) define multiple-trait
selection, (ii) explain its main methods, (iii) show when each method is
appropriate, and (iv) provide a comprehensive comparison analysis.

At the end, the breeder should be able to justify the selected method. The
justification should refer to the objective, the available information, the
expected responses, and the stability of the final ranking.

## 2. What is multiple-trait selection?

Multiple-trait selection is the choice of candidates using two or more traits.
The traits can enter the decision at the same stage or at different stages.
The breeder must first decide how trade-offs are allowed.

Four broad strategies are recognised by Beavis, Lamkey, Mahama, and Suza
(2023).

| Strategy | Definition | Suitable situation | Main limitation |
|---|---|---|---|
| Multistage selection | Different traits are evaluated at different stages. | Expensive traits are measured after inexpensive screening. | Early rejection can remove candidates that would perform well later. |
| Tandem selection | One trait is improved first. Another trait becomes the focus later. | One urgent trait dominates a specific breeding stage. | Other traits can deteriorate while attention is placed elsewhere. |
| Independent culling | Every candidate must satisfy a limit for every required trait. | Release standards or biological limits are firm. | Excellence elsewhere gives no compensation for a failed limit. |
| Index selection | Traits are combined into one score of overall merit. | Trade-offs are allowed and simultaneous improvement is required. | The score is useful only when its objective and inputs are defensible. |

These strategies answer different questions. Hence, a comparison should never
treat them as interchangeable algorithms. The breeder's decision rule comes
first. The statistical method follows.

## 3. The quantities used by an index

Let \(\mathbf{x}\) contain the information recorded for a candidate. Let
\(\mathbf{g}\) contain the additive genetic values of the objective traits.
The two vectors can contain the same traits. They can also differ. For example,
records from relatives, testcrosses, environments, or correlated traits can
help predict a smaller set of objective traits.

Three covariance matrices describe this problem.

| Symbol | Definition | Breeding meaning |
|---|---|---|
| \(\mathbf{P}=\operatorname{Var}(\mathbf{x})\) | Covariance among information variables | Describes the variation and redundancy in the evidence used for selection. |
| \(\mathbf{C}=\operatorname{Cov}(\mathbf{x},\mathbf{g})\) | Covariance between information and objective traits | Describes how strongly each information source predicts genetic value. |
| \(\mathbf{G}=\operatorname{Var}(\mathbf{g})\) | Genetic covariance among objective traits | Describes available genetic variation and correlated response. |

The aggregate genetic merit is

$$
H=\mathbf{a}^{\mathsf T}\mathbf{g},
$$

where \(\mathbf{a}\) contains economic weights. A linear selection index is

$$
I=\mathbf{b}^{\mathsf T}\mathbf{x},
$$

where \(\mathbf{b}\) contains index coefficients. Under truncation selection,
the expected response vector is

$$
\Delta
=i\frac{\mathbf{C}^{\mathsf T}\mathbf{b}}
{\sqrt{\mathbf{b}^{\mathsf T}\mathbf{P}\mathbf{b}}}.
$$

The term \(i\) is the selection intensity. It increases when a smaller
proportion of candidates is selected. When information and objective traits
are identical, \(\mathbf{C}=\mathbf{G}\). This gives the familiar expression
used throughout classical selection-index theory.

Use `selection_information()` and `generalized_index()` when the information
and objective traits differ. Read [Using predictions, prediction errors, and
restricted responses](DesiredGainR-information.html) for the complete
formulation.

## 4. The methods available in DesiredGainR

### 4.1 Economic-value methods

The Smith-Hazel index maximises expected response in aggregate genetic merit.
Its coefficients are

$$
\mathbf{b}=\mathbf{P}^{-1}\mathbf{G}\mathbf{a}.
$$

Choose this method when the relative economic values are credible. It is also
the correct reference when every method is compared against one fixed
definition of merit.

The base index uses \(\mathbf{b}=\mathbf{a}\). It applies the economic weights
directly. It provides a useful benchmark because it avoids covariance
inversion. A small advantage for Smith-Hazel over the base index suggests that
the estimated covariance structure contributes little to the decision.

### 4.2 Desired-gain methods

The Pesek-Baker and Yamada formulations use a desired response direction
\(\mathbf{d}\). Their coefficient expressions are

$$
\mathbf{b}=\mathbf{G}^{-1}\mathbf{d}
$$

and

$$
\mathbf{b}=\mathbf{P}^{-1}\mathbf{G}
(\mathbf{G}\mathbf{P}^{-1}\mathbf{G})^{-1}\mathbf{d}.
$$

They give the same coefficient direction when \(\mathbf{G}\) is square and
invertible. Therefore, they are two computational routes to the same linear
index. A material difference between their answers indicates numerical
instability or a rank-deficient matrix.

Choose this family when the breeder can state relative desired gains more
confidently than economic weights. The vector fixes response proportions. The
selection intensity fixes the attainable magnitude. Multiplying every desired
gain by the same positive constant leaves the ranking unchanged.

`run_dgsi()` implements the iterative desired-gain selection index (DGSI). It
searches for a candidate ranking whose selected differential is close to the
requested vector. The closed-form desired-gain index remains its mathematical
reference. Read [Optimising desired gains](DesiredGainR-dgsi.html) before using
the iterative method.

### 4.3 Rank and threshold methods

The Mulamba-Mock rank-sum method ranks candidates within traits and combines
the ranks. It requires neither economic weights nor covariance matrices when
equal rank weights are used. It is valuable as a simple comparator. Its score
does not provide a closed-form predicted genetic response in DesiredGainR.

Independent culling applies firm acceptance limits. Supply each limit in the
original trait units. Use a minimum for traits where larger values are
favourable. Use a maximum for traits listed in `lower_is_better`.

The Elston multiplicative index also applies firm limits. It then ranks the
eligible candidates using their joint margins above those limits. It is useful
when balanced superiority among eligible candidates matters.

Tandem selection focuses on one trait at each stage. DesiredGainR includes it
as a historical and operational comparator.

### 4.4 Restricted methods

A restricted index maximises merit while controlling response in specified
traits. DesiredGainR provides five forms.

| Method | Breeding question |
|---|---|
| Kempthorne-Nordskog | Which index holds specified traits at zero expected change? |
| Restricted Smith-Hazel | Which projection form gives the same zero-response restriction? |
| Tallis | Which index preserves stated response proportions? |
| Mallard | Which index targets stated response amounts? |
| Harville | Which index balances merit against a finite restriction penalty? |

Restrictions usually reduce response in unconstrained net merit. That loss is
the cost of respecting the biological or commercial condition. Report both
the restricted response and the loss relative to the unrestricted index.

Satoh (2024) gives a geometric interpretation of proportional restriction.
The scalar \(\beta\) measures progress along the desired direction. The
Mahalanobis residual measures departure from that direction after accounting
for genetic covariance. DesiredGainR reports both quantities through
`evaluate_restricted_response()`.

### 4.5 Non-linear economic value

The quadratic genomic selection index (QGSI) represents linear, squared, and
cross-product economic values. Choose it when biological or economic merit is
genuinely curved. A desired gain is a response target. It is a different
quantity from a quadratic weight. Read [Quadratic genomic
selection](DesiredGainR-qgsi.html) for the full demonstration.

## 5. How to choose a method

Begin with the breeder's statement. Then choose the simplest method that
represents it accurately.

| Breeder's statement | Starting method | Required evidence |
|---|---|---|
| "I can defend relative economic values." | Smith-Hazel | \(\mathbf{G}\), \(\mathbf{P}\), and economic weights |
| "I can state the response direction." | Pesek-Baker or Yamada | \(\mathbf{G}\), and \(\mathbf{P}\) for Yamada |
| "I can state acceptable gain intervals." | `suggest_desired_gains()` | \(\mathbf{G}\), \(\mathbf{P}\), intervals, and candidates |
| "Every candidate must satisfy these limits." | Independent culling | Candidate values and limits |
| "Eligible candidates should have balanced margins." | Elston | Candidate values and limits |
| "Some traits should remain unchanged." | Restricted index | \(\mathbf{G}\), \(\mathbf{P}\), merit weights, and restrictions |
| "One trait is screened at each stage." | Tandem or multistage selection | Stage order and stage-specific evidence |
| "I need a simple comparator with few assumptions." | Base or Mulamba-Mock | Weights for the base index, or candidate ranks |
| "Merit contains curvature or trait interactions." | QGSI | Genomic estimated breeding values and quadratic economic values |

Use the following decision sequence.

1. Define the objective traits and their favourable directions.
2. Identify firm limits that allow no trade-off.
3. Decide whether the remaining objective is economic, desired-gain, or
   non-linear.
4. Confirm the covariance matrices and units required by that method.
5. Fit one simple benchmark and one objective-matched method.
6. Compare biological response before comparing scalar criteria.
7. Examine ranking stability and uncertainty.
8. Use simulation when repeated cycles or non-linear decisions matter.

## 6. Conditions for a fair comparison

A comprehensive comparison requires common conditions. Otherwise, a numerical
difference can reflect the setup rather than the method.

Use the same:

1. candidates and trait records;
2. trait directions and units;
3. genetic and phenotypic covariance estimates;
4. selected proportion or selection intensity;
5. aggregate-merit definition when comparing \(R_{HI}\) or \(\Delta H\);
6. training, validation, and test partitions;
7. simulation scenarios, replicate seeds, and computational budget.

The comparison also needs several outcomes. No single statistic is sufficient.

| Outcome | Question answered |
|---|---|
| Expected response for every trait | What genetic change is predicted? |
| Target attainment | How much of each requested gain is reached? |
| Feasibility | Can the complete target be reached at the planned intensity? |
| \(R_{HI}\) | How closely does the index rank one fixed aggregate merit? |
| \(\Delta H\) | How much change in that fixed merit is expected? |
| Relative efficiency (RE) | How much response in the declared main trait remains relative to direct selection? |
| Index heritability | How repeatable is the composite score under the fitted covariance model? |
| Mahalanobis alignment | Does the response follow the requested multivariate direction? |
| Selected-set overlap | Do two methods choose the same candidates? |
| Rank correlation | Do two methods order all candidates similarly? |
| Selection probability | Does candidate choice remain stable after prediction uncertainty is propagated? |
| Information-deletion efficiency | Which records contribute useful selection information? |
| Multi-cycle response and diversity | Does the apparent advantage persist across breeding cycles? |

The coefficient of variation of the index, \(CV_I\), is retained for
reproduction of published analyses. It depends on the score origin and becomes
undefined after centring. Therefore, it should never decide the preferred
method.

## 7. A comprehensive comparison analysis

### 7.1 State the objective

The example seeks gains in grain yield (GY), ears per plant (EPP), and lower
values for plant height (PHT), anthesis date (AD), anthesis-silking interval
(ASI), and grey leaf spot score (GLS). The target is expressed in genetic
standard deviations.

```{r comparison-objective}
desired_gains_sd <- c(
  GY = 1.0,
  PHT = 0.4,
  AD = 0.6,
  ASI = 0.5,
  EPP = 0.4,
  GLS = 0.6
)
```

First test feasibility. This separates an unrealistic target from a poor
method.

```{r comparison-feasibility}
feasibility <- gain_feasibility(
  desired_gains_sd,
  dgr_G,
  dgr_P,
  n_candidates = nrow(dgr_candidates),
  n_select = 20L,
  lower_is_better = lower_is_better,
  gain_units = "genetic_sd"
)
feasibility
```

### 7.2 Put every method on one explicit scale

The following transformation expresses candidates and covariance matrices in
genetic-standard-deviation units. `selection_index()` then receives
`scale_traits = FALSE`. Hence, the desired-gain vector and every reported
response use the same units.

```{r prepare-comparison}
genetic_sd <- sqrt(diag(dgr_G))[traits]
inverse_sd <- diag(1 / genetic_sd, nrow = length(traits))
dimnames(inverse_sd) <- list(traits, traits)

G_sd <- inverse_sd %*% dgr_G %*% inverse_sd
P_sd <- inverse_sd %*% dgr_P %*% inverse_sd

candidates_sd <- dgr_candidates
centred <- sweep(
  as.matrix(dgr_candidates[, traits]),
  2L,
  colMeans(dgr_candidates[, traits]),
  "-"
)
candidates_sd[, traits] <- sweep(centred, 2L, genetic_sd, "/")
```

The economic weights below are implied by the desired-gain direction. This
creates one common merit definition for the mathematical comparison.

```{r common-merit}
common_weights <- implied_economic_weights(
  desired_gains_sd,
  G_sd,
  P_sd,
  lower_is_better = lower_is_better,
  gain_units = "trait"
)
round(common_weights, 3)
```

Negative implied weights can occur. They are mathematically valid under
correlated response. The desired response remains the biological statement to
interpret.

### 7.3 Fit objective-matched methods and benchmarks

```{r fit-comparison}
common_arguments <- list(
  values = candidates_sd,
  trait_cols = traits,
  id_col = "GenoID",
  G = G_sd,
  P = P_sd,
  lower_is_better = lower_is_better,
  center_traits = FALSE,
  scale_traits = FALSE,
  n_select = 20L,
  main_trait = "GY"
)

smith_hazel <- do.call(selection_index, c(
  common_arguments,
  list(method = "smith_hazel", economic_weights = common_weights)
))

pesek_baker <- do.call(selection_index, c(
  common_arguments,
  list(
    method = "pesek_baker",
    desired_gains = desired_gains_sd,
    aggregate_weights = common_weights
  )
))

base_index <- do.call(selection_index, c(
  common_arguments,
  list(method = "base", economic_weights = common_weights)
))

rank_sum <- do.call(selection_index, c(
  common_arguments,
  list(method = "mulamba_mock")
))
```

The first two methods should agree because the economic weights were derived
from the desired gains. This is a mathematical verification. The base and
rank-sum methods provide simpler benchmarks.

Threshold methods require limits rather than a response vector. These limits
use the units of `candidates_sd`. A maximum is supplied for each trait listed
in `lower_is_better`.

```{r fit-threshold-methods}
limits_sd <- c(
  GY = -0.50,
  PHT = 0.75,
  AD = 0.75,
  ASI = 0.75,
  EPP = -0.75,
  GLS = 0.75
)

culling <- do.call(selection_index, c(
  common_arguments,
  list(
    method = "independent_culling",
    culling_thresholds = limits_sd
  )
))

elston <- do.call(selection_index, c(
  common_arguments,
  list(method = "elston", culling_thresholds = limits_sd)
))

tandem <- do.call(selection_index, c(
  common_arguments,
  list(method = "tandem", tandem_order = c("GY", "GLS", "AD"))
))
```

### 7.4 Assemble the comparison

```{r run-comparison}
comparison <- compare_selection_methods(
  list(
    Smith_Hazel = smith_hazel,
    Pesek_Baker = pesek_baker,
    Base = base_index,
    Rank_sum = rank_sum,
    Culling = culling,
    Elston = elston,
    Tandem = tandem
  ),
  target_gains = desired_gains_sd
)
comparison
```

Read the fairness table first.

```{r comparison-fairness}
comparison$fairness
```

A culling method may select fewer than 20 candidates. That result is useful.
It shows that the limits and the requested selection quota conflict.

### 7.5 Compare biological response

```{r comparison-responses}
comparison$responses[
  , c(
    "Method", "Trait", "Expected_response",
    "Observed_differential", "Expected_attainment"
  )
]
```

Expected response and observed differential answer different questions.
Expected response predicts transmitted change under the covariance model.
Observed differential describes the candidates selected from this data set.
Rank, threshold, and tandem methods retain the second quantity because a
closed-form expected response is unavailable here.

Target attainment equals expected response divided by the requested gain. A
value of 1 reaches the target for that trait. A value of 0.70 reaches 70%.
Negative values move in an unfavourable direction. Read the worst trait before
the mean because one severe failure can be hidden by strong gains elsewhere.

### 7.6 Compare summary criteria

```{r comparison-summary}
comparison$summary[
  , c(
    "Method", "N_selected", "R_HI", "Delta_H", "RE",
    "Index_heritability", "Worst_expected_attainment",
    "Mahalanobis_alignment", "Mahalanobis_residual"
  )
]
```

Use each quantity for its defined purpose.

1. Read `R_HI` and `Delta_H` only where the same aggregate merit was supplied.
2. Read `RE` as efficiency for GY alone.
3. Read index heritability as repeatability of the composite score.
4. Read worst attainment as protection against hidden trait failure.
5. Read Mahalanobis alignment as agreement with the complete desired
   direction.
6. Read the residual as departure from the requested proportions.

### 7.7 Compare rankings and selected candidates

```{r comparison-decisions}
round(comparison$rank_correlation, 2)
round(comparison$selected_jaccard, 2)
comparison$selected_overlap
```

Rank correlation uses every candidate. Jaccard similarity uses the selected
sets. A high rank correlation can coexist with a different parent list near
the selection boundary. Therefore, report both.

## 8. Add uncertainty and validation

The point comparison above is the beginning of the analysis. A breeding
recommendation also needs stability evidence.

### 8.1 Covariance and weight uncertainty

Use `index_uncertainty()` to resample the genetic and phenotypic covariance
matrices. Use `weight_sensitivity()` when economic weights are uncertain.
Report coefficient intervals, response intervals, rank stability, and
selected-set stability.

### 8.2 Candidate prediction uncertainty

Genomic estimated breeding values (GEBVs) have prediction error. When
candidate-specific prediction error covariance (PEC) is available, use
`candidate_score_uncertainty()`. Report each candidate's score interval and
selection probability. Give particular attention to candidates near the
selection boundary.

### 8.3 Value of each information source

Use `index_information_efficiency()` for a Smith-Hazel or general economic
index. It applies Cunningham's deletion-efficiency principle. A trait with
little information value can then be removed from future phenotyping plans if
the saving is operationally important.

### 8.4 Out-of-sample validation

Fit the index with training data. Fix its coefficients and transformations.
Apply `predict()` to a later cohort or a held-out set. Compare predicted
responses with realised breeding values, progeny means, or later-stage trial
performance. This separates genuine prediction from reuse of the fitting data.

### 8.5 Multi-cycle validation

Use `simulate_selection_cycles()` when the decision concerns several cycles.
Compare cumulative response, joint target attainment, genetic variance, and
diversity. Use matched scenarios and replicate seeds. Read [Multi-cycle
simulation](DesiredGainR-simulation.html) for calibration and stress testing.

## 9. Choosing the preferred method

A method is preferred when five conditions are satisfied.

1. It represents the breeder's actual objective.
2. Its required inputs have adequate quality.
3. Its expected response respects every critical trait.
4. Its ranking remains stable under credible uncertainty.
5. Its advantage persists in validation or relevant multi-cycle scenarios.

The method with the largest single statistic can fail this standard. For
example, a high aggregate-merit response can hide failure of a disease limit.
A stable rank can also follow the wrong biological direction. Therefore, the
final recommendation should report the complete decision panel.

Close the analysis with a practical statement. Name the selected method.
State why it matches the objective. Report the expected response for every
trait. State the main uncertainty. Then describe how the decision will be
validated in the breeding programme.

## 10. References

- Beavis W, Lamkey K, Mahama AA, Suza W (2023). Multiple Trait Selection. In
  Suza WP and Lamkey KR, editors, *Quantitative Genetics for Plant Breeding*.
  Iowa State University Digital Press.
- Cunningham EP (1969). The relative efficiencies of selection indexes.
  *Acta Agriculturae Scandinavica* **19**:45-48.
- Elston RC (1963). A weight-free index for the purpose of ranking or selection
  with respect to several traits at a time. *Biometrics* **19**:85-97.
- Kempthorne O, Nordskog AW (1959). Restricted selection indices. *Biometrics*
  **15**:10-19.
- Mulamba NN, Mock JJ (1978). Improvement of yield potential of the Eto Blanco
  maize population by breeding for plant traits. *Egyptian Journal of Genetics
  and Cytology* **7**:40-51.
- Pesek J, Baker RJ (1969). Desired improvement in relation to selection
  indices. *Canadian Journal of Plant Science* **49**:803-804.
- Satoh M (2024). Characteristics of restricted selection indices and
  geometrical interpretation of restricted breeding values. *Journal of
  Animal Breeding and Genetics* **141**:353-363.
- Smith HF (1936). A discriminant function for plant selection. *Annals of
  Eugenics* **7**:240-250.
- Yamada Y, Yokouchi K, Nishida A (1975). Selection index when genetic gains of
  individual traits are of primary concern. *Japanese Journal of Genetics*
  **50**:33-41.
