---
title: "Obtaining G and P"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Obtaining G and P}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
library(DesiredGainR)
data(dgr_traits, package = "DesiredGainR")
data(dgr_G, package = "DesiredGainR")
data(dgr_P, package = "DesiredGainR")
data(dgr_gebv, package = "DesiredGainR")
traits <- dgr_traits$trait
```

## 1. What the two matrices mean

The genetic covariance matrix $\mathbf{G}$ describes additive genetic
variation and covariation in the target population. The phenotypic or
index-variable covariance matrix $\mathbf{P}$ describes the variables on which
the index is calculated. They answer different questions and are not
interchangeable.

The strongest source is a fitted multi-trait genetic model that reports the
trait covariance components and their uncertainty. DesiredGainR begins after
that model has been fitted; it does not estimate breeding values from raw field
records.

Never use the covariance of raw phenotypes as $\mathbf{G}$. It contains
environmental, residual and design variation. Likewise, the covariance of best
linear unbiased predictions or genomic estimated breeding values (GEBVs) is a
covariance of predictions and is usually shrunken relative to total additive
genetic covariance.

---

## 2. Import a fitted covariance safely

`import_covariance()` matches trait names, restores the requested order,
distinguishes correlations from covariances and records the estimand.

```{r import}
shuffled_G <- dgr_G[rev(traits), rev(traits)]

imported <- import_covariance(
  shuffled_G,
  trait_cols = traits,
  source = "matrix",
  estimand = "additive genetic covariance"
)
imported
```

If software reports a correlation matrix and separate variances, declare that
explicitly.

```{r correlation}
imported_from_correlation <- import_covariance(
  stats::cov2cor(dgr_G),
  trait_cols = traits,
  source = "matrix",
  estimand = "additive genetic covariance",
  is_correlation = TRUE,
  variances = diag(dgr_G)
)
max(abs(imported_from_correlation$covariance - dgr_G))
```

---

## 3. When the fitted G matrix is unavailable

`estimate_genetic_covariance()` provides labelled working alternatives. The
method name is part of the scientific conclusion.

| Method | What it estimates | Principal limitation |
|---|---|---|
| `prediction_covariance` | Covariance of supplied genetic predictions | Does not undo prediction shrinkage |
| `pev_corrected` | Prediction covariance plus mean prediction-error covariance | Requires full, compatible cross-trait prediction-error covariance |
| `se_diagonal_corrected` | Prediction covariance with diagonal error correction | Cannot recover cross-trait prediction-error covariance |
| `relationship_adjusted` | Prediction covariance adjusted for relationships | Still does not undo prediction shrinkage |
| `adjusted_means_surrogate` | Covariance of adjusted means | Working surrogate, not additive genetic covariance |

For example:

```{r working-estimate}
working_G <- estimate_genetic_covariance(
  dgr_gebv,
  trait_cols = traits,
  method = "prediction_covariance"
)
working_G
```

The printed estimand should appear in every analysis report. Do not shorten it
to “genetic covariance” unless that is what the upstream model estimated.

---

## 4. Diagnose both matrices

```{r diagnostics}
matrix_diagnostics(dgr_G, "G")
matrix_diagnostics(dgr_P, "P")
```

Check the following before fitting an index:

1. Row and column names contain exactly the same traits in the analysis order.
2. Units agree with the candidate values and objective.
3. Both matrices are symmetric and finite.
4. Required matrices are positive definite, or any rank deficiency is
   understood.
5. $\mathbf{P}-\mathbf{G}$ is positive semidefinite when the matrices describe
   compatible phenotypic and additive genetic variation.
6. The condition number is not so large that small input errors dominate the
   coefficients.

`bend_covariance()` can repair a small numerical definiteness failure. It
cannot turn an inappropriate estimand into an appropriate one. Always report
the original minimum eigenvalue and the size of the adjustment.

---

## 5. What to report

For each matrix, record its source, population, traits, units, model term,
estimation method, software version, uncertainty information and any repair.
If only a working surrogate is available, repeat the analysis once a fitted
multi-trait covariance becomes available and examine whether the ranking and
selected set change.

The next step is [Choosing an index](DesiredGainR-index-families.html).

