---
title: "Reproducing published results"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Reproducing published results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(DesiredGainR)
```

## Why this vignette exists

Every other test in this package checks it against itself, or against an
equation it also implements. That establishes internal consistency and nothing
else. A formula can be implemented consistently and still be the wrong formula.

This vignette checks the package against numbers produced by other people, with
other software, on data this package has never seen: Rahimi and Debnath (2023),
*Scientific Reports* **13**, 18977, computed in SAS PROC IML and published in
tables.

The reference values are frozen in `rahimi_debnath_2023()`. Preserve them
exactly. The moment they are derived from this package they stop being
an external check.

```{r reference}
reference <- rahimi_debnath_2023()
reference$traits
```

## Scope of the reproduction

The article's genetic and phenotypic covariance matrices are in supplementary
data files that are absent from the publication. The index **coefficients**
therefore remain unavailable for direct recomputation. This limits the
quantities that can be checked.

It is not, because the Pesek–Baker index has an exact algebraic property that
the published table either satisfies or does not.

### The identity

Pesek–Baker sets $\mathbf{b} = \mathbf{G}^{-1}\mathbf{d}$. The expected
correlated response to selection is

$$\Delta = i\,\frac{\mathbf{G}\mathbf{b}}
{\sqrt{\mathbf{b}^\mathsf{T}\mathbf{P}\mathbf{b}}}
= i\,\frac{\mathbf{G}\mathbf{G}^{-1}\mathbf{d}}
{\sqrt{\mathbf{b}^\mathsf{T}\mathbf{P}\mathbf{b}}}
= \frac{i}{\sqrt{\mathbf{b}^\mathsf{T}\mathbf{P}\mathbf{b}}}\,\mathbf{d}.$$

The response is **exactly proportional to the desired gains**, with a single
constant of proportionality shared by every trait. No approximation is
involved.

So: divide the article's published gains by the article's published $d$ vector.
If the authors' implementation and this package's understanding of the method
agree, the seven ratios from the unrounded calculations must be the same
number. Ratios calculated from the printed table can differ because both the
numerator and denominator were rounded.

```{r identity}
ratio <- reference$pesek_baker_gain / reference$desired_gains
round(ratio, 5)
```

```{r identity-summary}
c(
  mean = mean(ratio),
  sd = stats::sd(ratio),
  relative_sd = stats::sd(ratio) / mean(ratio)
)
```

The displayed ratios do not agree to three or four significant figures. Six
round to 0.336 at three significant figures. `row_length` rounds to 0.335. All
seven round to 0.34 at two significant figures. The relative standard
deviation is about 0.00066. This small spread is descriptive evidence, but it
does not define the reproduction tolerance.

The appropriate check propagates the printed decimal precision. The desired
gains are printed to three decimal places. Their rounding half-unit is
0.0005. The responses are printed to four decimal places. Their rounding
half-unit is 0.00005.

```{r identity-rounding-intervals}
desired_half_unit <- 0.0005
response_half_unit <- 0.00005

ratio_lower <-
  (reference$pesek_baker_gain - response_half_unit) /
    (reference$desired_gains + desired_half_unit)
ratio_upper <-
  (reference$pesek_baker_gain + response_half_unit) /
    (reference$desired_gains - desired_half_unit)

common_ratio_interval <- c(
  lower = max(ratio_lower),
  upper = min(ratio_upper)
)
round(common_ratio_interval, 7)
```

The seven intervals share the common range 0.3358452 to 0.3358549. The value
0.33585 lies inside this range. One exact proportionality constant can
therefore produce every published value after rounding.

### Rounding rules used by the test

| Published quantity | Printed precision | Rounding interval used |
|---|---:|---:|
| Desired gain | Three decimal places | Printed value $\pm 0.0005$ |
| Expected response | Four decimal places | Printed value $\pm 0.00005$ |
| Common proportionality constant | Intersection across all seven ratio intervals | 0.3358452 to 0.3358549 |

These bounds come directly from the printed decimal places. They were not
estimated from the observed ratio spread. The joint interval check is stronger
than a loose comparison tolerance. It verifies the defining Pesek-Baker
response identity against seven published traits. Direct coefficient
reproduction still requires the unpublished covariance matrices.

## Why the reported $R_{HI}=0.0018$ does not invalidate Pesek-Baker

The article reports $R_{HI}=0.0018$ for Pesek-Baker and $0.9887$ for the
Smith-Hazel optimum index. These values answer a net-merit question. They do
not measure whether Pesek-Baker achieved its desired response direction.

Pesek-Baker starts from a desired-gain vector $\mathbf{d}$. This vector states
the relative genetic changes sought across traits. By contrast, $R_{HI}$ is the
correlation between the index $I$ and an aggregate merit
$H=\mathbf{a}^{\mathsf T}\mathbf{g}$. It requires a separate vector of merit
weights $\mathbf{a}$.

The article used $\mathbf{a}=\mathbf{d}$. With
$\mathbf{b}=\mathbf{G}^{-1}\mathbf{d}$, this gives

$$R_{HI} = \frac{\mathbf{d}^\mathsf{T}\mathbf{d}}
{\sqrt{\mathbf{d}^\mathsf{T}\mathbf{G}^{-1}\mathbf{P}\mathbf{G}^{-1}\mathbf{d}}
\;\sqrt{\mathbf{d}^\mathsf{T}\mathbf{G}\mathbf{d}}}.$$

This expression is mathematically the correlation between the Pesek-Baker
index and the constructed merit $H_{\mathbf d}=\mathbf{d}^{\mathsf T}\mathbf{g}$.
The difficulty is biological interpretation. Desired gains carry response
units. Merit weights carry value per trait unit. Substituting one for the other
creates a merit definition that changes when trait measurement units change.

The Pesek-Baker decision itself has the correct unit invariance. If traits are
rescaled by a diagonal matrix $\mathbf{S}$, then
$\mathbf{d}^{*}=\mathbf{S}\mathbf{d}$ and
$\mathbf{b}^{*}=\mathbf{S}^{-1}\mathbf{b}$. Therefore,
${\mathbf{b}^{*}}^{\mathsf T}\mathbf{x}^{*}=\mathbf{b}^{\mathsf T}\mathbf{x}$.
The candidate ranking is unchanged. However, treating desired gains as merit
weights gives
$H_{\mathbf d}^{*}=\mathbf{d}^{\mathsf T}\mathbf{S}^{2}\mathbf{g}$.
That merit definition changes with the units.

```{r scales}
range(reference$desired_gains)
max(reference$desired_gains) / min(reference$desired_gains)
```

The numerical desired gains span a factor of about 200. This scale disparity
makes the substitution especially fragile. The value 0.0018 therefore gives
little information about attainment of the desired-gain objective.

### How to evaluate a desired-gain index

Desired gains define a valid multivariate response objective. Economic weights
are unnecessary for fitting that objective. The primary evaluation should
report:

1. expected genetic response for every trait;
2. proportionality or Mahalanobis alignment with $\mathbf{d}$;
3. the attainable fraction at the planned selection intensity;
4. worst-trait and joint attainment when targets are minimum floors;
5. observed selection differentials for the current candidates; and
6. stability under covariance and prediction uncertainty.

$R_{HI}$ and $\Delta H$ remain useful when the breeding programme also defines
one common aggregate merit vector $\mathbf{a}$. They then provide an additional
economic comparison under the same merit definition for every method. They are
not prerequisites for a valid desired-gain analysis.

### How DesiredGainR handles the distinction

When `aggregate_weights` is absent, DesiredGainR still fits the desired-gain
index and reports its per-trait response and attainment diagnostics. It returns
`NA` for $R_{HI}$ and $\Delta H$ because the separate aggregate merit $H$ has
not been defined.

When the breeder supplies one common `aggregate_weights` vector, DesiredGainR
reports $R_{HI}$ and $\Delta H$ against that same merit for every fitted method.
The comparison then has a common economic yardstick.

The implied weights

$$\mathbf{w}=\mathbf{G}^{-1}\mathbf{P}\mathbf{G}^{-1}\mathbf{d}$$

identify the Smith-Hazel objective that reproduces a given Pesek-Baker
coefficient direction. `implied_economic_weights()` provides this translation
as a diagnostic. Using a different implied vector for each desired-gain
direction would also change the comparison yardstick. DesiredGainR therefore
keeps that choice explicit.

## The quadratic index: a reference implementation

The linear families are anchored above. `run_qgsi()` implements a different
paper — Cerón-Rojas et al. (2026) — and needs its own anchor.

That paper's replication deposit contains the authors' own R scripts. The
intermediate data files those scripts read were absent from the deposit. Their
published numbers therefore remain unavailable for direct recomputation. The
**algorithm** can still be reproduced. A reference implementation provides a
stronger comparison than a table because it can
be run on any input rather than only on theirs.

`tests/testthat/test-reference-implementation.R` transcribes their code
verbatim and runs both implementations on identical inputs. The quantities
compared are:

| Reference | This package |
|---|---|
| `b = P^-1 G w` | `selection_index(method = "smith_hazel")` coefficients |
| `corrHI = sqrt(b'Pb) / sqrt(w'Gw)` | `evaluate_index()$R_HI` |
| `Rs = ks * sqrt(b'Pb)` | `evaluate_index()$delta_H` |
| `Vpq = b'Pb + 2 tr(BB P BB P)` | `run_qgsi()` index variance |
| `VHq = w'Gw + 2 tr(A G A G)` | QGSI merit variance |

The second row is worth noting because the two expressions do not look alike.
The reference computes accuracy as a ratio of standard deviations. This package
computes a correlation. For the optimum index $\mathbf{b} =
\mathbf{P}^{-1}\mathbf{G}\mathbf{w}$ we have $\mathbf{b}'\mathbf{G}\mathbf{w} =
\mathbf{b}'\mathbf{P}\mathbf{b}$, so they are the same number by an identity
that is itself asserted as a separate test. Agreement establishes that both
parties mean the same thing by accuracy, which a coefficient comparison alone
would not.

## What this reproduction does not establish

Being explicit, because a reproduction vignette invites more confidence than it
earns.

- **The Rahimi and Debnath coefficients are not yet compared.** Their article
  points to BioStudies accession `S-BSST853`. The current vignette freezes the
  printed tables but does not yet redistribute or reanalyse that archive, so it
  verifies the defining property rather than claiming a complete rerun.
- **The Cerón-Rojas comparison is against code, not against published
  numbers.** It establishes that two independent implementations of the same
  algorithm agree. It does not establish that either reproduces the figures in
  the paper, because the inputs behind those figures were not deposited.
- **Neither anchor covers the simulation layer.** Forward simulation over
  breeding cycles is checked only against its own internal consistency.
- **The Monte Carlo tests remain the primary evidence** for the rest of the
  package. This vignette adds two external anchors. It does not replace them.

The qualitative orderings that *are* checkable, and hold:

```{r ordering}
c(
  optimum_RE = reference$optimum_method1_criteria[["RE"]],
  pesek_baker_RE = reference$pesek_baker_criteria[["RE"]]
)
c(
  optimum_yield_gain = reference$optimum_method1_gain[["yield"]],
  pesek_baker_yield_gain = reference$pesek_baker_gain[["yield"]]
)
```

An index using economic weights achieves more response in the main trait than
one constrained to a fixed proportional direction. That is the expected
trade-off, not a defect of either method.

## References

Rahimi, M. and Debnath, S. (2023) Estimating optimum and base selection indices
in plant and animal breeding programs by development new and simple SAS and R
codes. *Scientific Reports* **13**,
18977. \doi{10.1038/s41598-023-46368-6}

Cerón-Rojas, J.J., Montesinos-López, O.A., Montesinos-López, A., Vitale, P.,
Pérez-Rodríguez, P., Fernandes, S.B., Ortiz, R. and Crossa, J. (2026)
Replication Data for: Nonlinear Genomic Selection Index Accelerates Multi-Trait
Crop Improvement. CIMMYT Research Data & Software Repository Network, V1.
\doi{10.71682/10549385}
